### 2025-02-19 writing/notes.tex
Some notes on a dimensional ODE model.

\documentclass{article}

\usepackage[margin=1.5in]{geometry}
\usepackage{amsmath}

\begin{document}

\begin{flushleft}

\section{Collier et al., 1996}

Collier et al. make the following five modelling assumptions:
\begin{enumerate}
  \item Cells interact through Delta-Notch signalling if and only if they are adjacent.
  \item Production of Notch is an increasing (hill) function of Delta in neighbouring cells.
  \item Production of Delta is a decreasing (hill) function of Notch in the same cell.
  \item Production of Notch and Delta is balanced by decay proportional to concentration.
  \item Low levels of Notch cause the primary fate, high levels cause the secondary fate.
\end{enumerate}

Define the following variables:

\begin{itemize}
  \item $\tau$: time
  \item $N_{P}$: notch activity (concentration) in cell $P$
  \item $D_{P}$: delta activity (concentration) in cell $P$
  \item $\overline{D}_{P}$: average delta activity in neighbours of $P$
  \item $N_{0}$: typical notch activity (across all cells)
  \item $D_{0}$: typical delta activity (across all cells)
  \item $\mu$: The decay rate for notch (assumed constant)
  \item $\rho$: The decay rate for delta (assumed constant)
\end{itemize}

Then, define the following system of differential equations, where $F:[0, \infty) \rightarrow [0, \infty)$ is a continuous increasing function and $G: [0, \infty)\rightarrow [0, \infty)$ is a continuous decreasing function.

$$
\begin{aligned}
  \frac{d(N_{P} / N_{0})}{d\tau} = F(\overline{D}_{P} / D_{0}) - \mu N_{P} / N_{0} \\[5pt]
  \frac{d(D_{P} / D_{0})}{d\tau} = G(N_{P} / N_{0}) - \rho D_{P} / D_{0}
\end{aligned}
$$

This equation is inherently nondimensional, since $N_{P} / N_{0}$ and $D_{P} / D_{0}$ have no units.

\subsection{Writing Down a Dimensional Model}

Let us continue to use $N_{P}$, $D_{P}$, and $\overline{D}_{P}$ as defined above. Let $t$ denote the dimensional time and let $u$ and $r$ be the dimensional decay rates for notch and delta. Then, assuming the dimensional analogues of $F$ and $G$ are hill functions of order $1$, we have:

$$
\begin{aligned}
  \frac{dN_{P}}{dt} &= \frac{\overline{D}_{P}}{d_{0} + \overline{D}_{P}} - u N_{P} \\[5pt]
  \frac{dD_{P}}{dt} &= \frac{n_{0}}{n_{0} + N_{P}} - rD_{P}
\end{aligned}
$$

In the equation above, $d_{0}$ is the half-max of notch production induced by delta in neighbouring cells while $n_{0}$ is the half-min of delta production induced by notch.

\medskip

This model is a good start, but we will need to explicitly model the concentration of delta-notch complexes $C$ in order to write down a probabilistic model. Since $D_{P}$ and $N_{P}$ denote the level of delta and notch \emph{activation} (that is, the concentration of complexes),

\subsection{Deriving a Probabilistic Model}

TBD

\end{flushleft}

\end{document}

### 2025-02-20 code/DeltaNotch_Sim2_CellArrayHexVis3.py
Create DeltaNotch_Sim2_CellArrayHexVis3.py
Simulation and visualization of 2D hexagonal cell array with deterministic Delta/Notch dynamics.

@@ -1,0 +1,191 @@
+import numpy as np
+from scipy.integrate import solve_ivp
+import matplotlib.pyplot as plt
+import matplotlib.animation as animation
+from matplotlib.patches import RegularPolygon
+import matplotlib.colors as mcolors
+
+# Delta-Notch system for hexagonal grid (unchanged from corrected version)
+def delta_notch_hex_grid(t, y, alpha_D, beta_D, alpha_N, beta_N, k, n, grid_size):
+    """
+    y is arranged as [D1, N1, D2, N2, ..., Dn, Nn]
+    Each cell interacts with its six neighbors in a hexagonal grid
+    with periodic boundary conditions (toroidal surface)
+    """
+    rows, cols = grid_size
+    total_cells = rows * cols
+    dydt = np.zeros_like(y)
+    
+    # Helper function to get the 1D index from 2D coordinates
+    def get_index(r, c):
+        r %= rows  # Wrap around rows
+        c %= cols  # Wrap around columns
+        return 2 * (r * cols + c)
+    
+    # For each cell, calculate its derivatives
+    for r in range(rows):
+        for c in range(cols):
+            cell_idx = r * cols + c
+            D_idx = 2 * cell_idx
+            N_idx = D_idx + 1
+            
+            # Define neighbor offsets for hexagonal grid
+            if r % 2 == 0:  # Even row
+                neighbor_offsets = [
+                    (-1, -1), (-1, 0),   # Top-left, Top-right
+                    (0, -1), (0, 1),     # Left, Right
+                    (1, -1), (1, 0)      # Bottom-left, Bottom-right
+                ]
+            else:  # Odd row
+                neighbor_offsets = [
+                    (-1, 0), (-1, 1),    # Top-left, Top-right
+                    (0, -1), (0, 1),     # Left, Right
+                    (1, 0), (1, 1)       # Bottom-left, Bottom-right
+                ]
+            
+            # Get neighbor Delta values
+            neighbor_deltas = []
+            for dr, dc in neighbor_offsets:
+                nr, nc = r + dr, c + dc
+                neighbor_idx = get_index(nr, nc)
+                neighbor_deltas.append(y[neighbor_idx])
+            
[…]

### 2025-02-20 code/DeltaNotch_Sim3_CellArrayHexVis_Stochastic.py
Create DeltaNotch_Sim3_CellArrayHexVis_Stochastic.py
Simulation and visualization of 2D hexagonal cell array of stochastic (added noise) differential equation dynamics.

@@ -1,0 +1,203 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.animation as animation
+from matplotlib.patches import RegularPolygon
+import matplotlib.colors as mcolors
+
+# Deterministic Delta-Notch system for hexagonal grid
+def delta_notch_hex_grid(t, y, alpha_D, beta_D, alpha_N, beta_N, k, n, grid_size):
+    """
+    y is arranged as [D1, N1, D2, N2, ..., Dn, Nn]
+    Each cell interacts with its six neighbors in a hexagonal grid
+    with periodic boundary conditions (toroidal surface)
+    """
+    rows, cols = grid_size
+    total_cells = rows * cols
+    dydt = np.zeros_like(y)
+    
+    # Helper function to get the 1D index from 2D coordinates
+    def get_index(r, c):
+        r %= rows  # Wrap around rows
+        c %= cols  # Wrap around columns
+        return 2 * (r * cols + c)
+    
+    # For each cell, calculate its derivatives
+    for r in range(rows):
+        for c in range(cols):
+            cell_idx = r * cols + c
+            D_idx = 2 * cell_idx
+            N_idx = D_idx + 1
+            
+            # Define neighbor offsets for hexagonal grid
+            if r % 2 == 0:  # Even row
+                neighbor_offsets = [
+                    (-1, -1), (-1, 0),   # Top-left, Top-right
+                    (0, -1), (0, 1),     # Left, Right
+                    (1, -1), (1, 0)      # Bottom-left, Bottom-right
+                ]
+            else:  # Odd row
+                neighbor_offsets = [
+                    (-1, 0), (-1, 1),    # Top-left, Top-right
+                    (0, -1), (0, 1),     # Left, Right
+                    (1, 0), (1, 1)       # Bottom-left, Bottom-right
+                ]
+            
+            # Get neighbor Delta values
+            neighbor_deltas = []
+            for dr, dc in neighbor_offsets:
+                nr, nc = r + dr, c + dc
+                neighbor_idx = get_index(nr, nc)
+                neighbor_deltas.append(y[neighbor_idx])
+            
+            # Calculate Delta inhibition by cell's own Notch
[…]

### 2025-02-20 code/DeltaNotch_Sim1_TwoCell.py
Create DeltaNotch_Sim1_TwoCell.py
Simple two cell simulation of deterministic Delta/Notch dynamics with graphical output of Delta/Notch levels in each cell.

@@ -1,0 +1,62 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import solve_ivp
+
+# --- Revised Parameter Choices ---
+k = 2.0       # Hill exponent for Notch activation
+h = 2.0       # Hill exponent for Delta inhibition
+beta = 1.5    # Production scaling for Notch; >1 to promote instability
+theta = 0.5   # Threshold for Delta activation of Notch
+mu = 0.5      # Threshold for Notch inhibition of Delta
+
+# --- Revised Hill Functions ---
+def f(D, k, theta):
+    """
+    Notch activation by Delta from the neighboring cell.
+    A threshold parameter theta sharpens the response.
+    """
+    return D**k / (theta**k + D**k)
+
+def g(N, h, mu):
+    """
+    Delta production inhibited by own Notch.
+    The parameter mu sets the inhibition threshold.
+    """
+    return 1 / (1 + (N/mu)**h)
+
+# --- Define the Revised ODE System ---
+def ode_system(t, y):
+    # y = [N1, D1, N2, D2] for the two-cell system.
+    N1, D1, N2, D2 = y
+    dN1_dt = beta * f(D2, k, theta) - N1
+    dD1_dt = g(N1, h, mu) - D1
+    dN2_dt = beta * f(D1, k, theta) - N2
+    dD2_dt = g(N2, h, mu) - D2
+    return [dN1_dt, dD1_dt, dN2_dt, dD2_dt]
+
+# --- Simulation Settings ---
+t_span = (0, 50)
+t_eval = np.linspace(t_span[0], t_span[1], 1000)
+# Choose initial conditions that break symmetry
+y0 = [0.1, 0.9, 0.15, 0.85]
+
+# --- Solve the ODEs ---
+sol = solve_ivp(ode_system, t_span, y0, t_eval=t_eval)
+
+# --- Plotting the Deterministic Dynamics ---
+plt.figure(figsize=(10, 8))
+plt.subplot(2, 1, 1)
+plt.plot(sol.t, sol.y[0], label='N1')
+plt.plot(sol.t, sol.y[2], label='N2')
+plt.xlabel('Time')
+plt.ylabel('Notch Level')
+plt.legend()
+
+plt.subplot(2, 1, 2)
+plt.plot(sol.t, sol.y[1], label='D1')
+plt.plot(sol.t, sol.y[3], label='D2')
+plt.xlabel('Time')
[…]

### 2025-02-20 code/DeltaNotch_Sim1_TwoCell_VaryStart.py
Create DeltaNotch_Sim1_TwoCell_VaryStart.py
Simple two-cell deterministic Delta/Notch simulation with noisy initial conditions for seeing sensitivity. Plots all solutions and mean.

@@ -1,0 +1,103 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import solve_ivp
+
+# --- Revised Parameter Choices ---
+k = 2.0       # Hill exponent for Notch activation
+h = 2.0       # Hill exponent for Delta inhibition
+beta = 1.5    # Production scaling for Notch; >1 to promote instability
+theta = 0.5   # Threshold for Delta activation of Notch
+mu = 0.5      # Threshold for Notch inhibition of Delta
+
+# --- Revised Hill Functions ---
+def f(D, k, theta):
+    """
+    Notch activation by Delta from the neighboring cell.
+    A threshold parameter theta sharpens the response.
+    """
+    return D**k / (theta**k + D**k)
+
+def g(N, h, mu):
+    """
+    Delta production inhibited by own Notch.
+    The parameter mu sets the inhibition threshold.
+    """
+    return 1 / (1 + (N/mu)**h)
+
+# --- Define the Revised ODE System ---
+def ode_system(t, y):
+    # y = [N1, D1, N2, D2] for the two-cell system.
+    N1, D1, N2, D2 = y
+    dN1_dt = beta * f(D2, k, theta) - N1
+    dD1_dt = g(N1, h, mu) - D1
+    dN2_dt = beta * f(D1, k, theta) - N2
+    dD2_dt = g(N2, h, mu) - D2
+    return [dN1_dt, dD1_dt, dN2_dt, dD2_dt]
+
+# --- Simulation Settings ---
+t_span = (0, 50)
+t_eval = np.linspace(t_span[0], t_span[1], 1000)
+
+# Number of different initial conditions to simulate
+num_simulations = 50
+
+# Prepare an array to store each solution (shape: num_simulations x 4 x len(t_eval))
+all_solutions = []
+
+# Loop over simulations with different starting positions.
+for i in range(num_simulations):
+    # Create a random perturbation about the baseline initial condition.
+    # Baseline: [0.1, 0.9, 0.15, 0.85]
+    y0 = [
+        0.1 + np.random.uniform(-0.05, 0.05),
+        0.9 + np.random.uniform(-0.05, 0.05),
+        0.15 + np.random.uniform(-0.05, 0.05),
+        0.85 + np.random.uniform(-0.05, 0.05)
+    ]
[…]

### 2025-02-20 code/DeltaNotch_Sim1_TwoCell_VaryStart.py
Update DeltaNotch_Sim1_TwoCell_VaryStart.py

@@ -39,7 +39,7 @@ t_span = (0, 50)
 t_eval = np.linspace(t_span[0], t_span[1], 1000)
 
 # Number of different initial conditions to simulate
-num_simulations = 50
+num_simulations = 30
 
 # Prepare an array to store each solution (shape: num_simulations x 4 x len(t_eval))
 all_solutions = []
@@ -49,10 +49,10 @@ for i in range(num_simulations):
     # Create a random perturbation about the baseline initial condition.
     # Baseline: [0.1, 0.9, 0.15, 0.85]
     y0 = [
-        0.1 + np.random.uniform(-0.05, 0.05),
-        0.9 + np.random.uniform(-0.05, 0.05),
-        0.15 + np.random.uniform(-0.05, 0.05),
-        0.85 + np.random.uniform(-0.05, 0.05)
+        0.92 + np.random.uniform(-0.05, 0.05),
+        0.92 + np.random.uniform(-0.05, 0.05),
+        0.92 + np.random.uniform(-0.05, 0.05),
+        0.92 + np.random.uniform(-0.05, 0.05)
     ]
     sol = solve_ivp(ode_system, t_span, y0, t_eval=t_eval, rtol=1e-6, atol=1e-9)
     all_solutions.append(sol.y)

### 2025-02-20 code/DeltaNotch_Sim1_TwoCell_CollierRecreate.py
Create DeltaNotch_Sim1_TwoCell_CollierRecreate.py
Attempt to recreate the two-cell deterministic numerical results in Collier et al. 1996 (Fig 3, page 434).

@@ -1,0 +1,74 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import solve_ivp
+
+# --- Adjusted Parameter Choices ---
+k = 2.0         # Hill exponent for Notch activation
+h = 2.0         # Hill exponent for Delta inhibition
+beta = 1.0      # Production scaling for Notch
+theta = 0.01    # Threshold for Delta activation of Notch
+mu = 100        # Threshold for Notch inhibition of Delta
+
+# --- Hill Functions ---
+def f(D, k, theta):
+    """Notch activation by Delta from neighboring cell."""
+    return D**k / (theta + D**k)
+
+def g(N, h, mu):
+    """Delta production inhibited by own Notch."""
+    return 1 / (1 + mu*(N)**h)
+
+# --- ODE System ---
+def ode_system(t, y):
+    N1, D1, N2, D2 = y
+    dN1_dt = beta * f(D2, k, theta) - N1
+    dD1_dt = g(N1, h, mu) - D1
+    dN2_dt = beta * f(D1, k, theta) - N2
+    dD2_dt = g(N2, h, mu) - D2
+    return [dN1_dt, dD1_dt, dN2_dt, dD2_dt]
+
+# --- Adjusted Simulation Settings ---
+t_span = (0, 25)  # Match the x-axis in your plots
+t_eval = np.linspace(t_span[0], t_span[1], 1000)
+
+# Initial conditions - adjusted to match plot patterns
+y0 = [1.0, 1.0, 0.99, 0.99]  # Slight asymmetry in D2 initial value
+
+# --- Solve the ODEs ---
+sol = solve_ivp(ode_system, t_span, y0, t_eval=t_eval, method='RK45', rtol=1e-6, atol=1e-9)
+
+# --- Create the four-panel plot matching the image ---
+fig, axs = plt.subplots(2, 2, figsize=(12, 10))
+
+lw = 1.0  # Line width for the plots
+
+# Top left: n₁
+axs[0, 0].plot(sol.t, sol.y[0], 'k-', linewidth=lw)
+axs[0, 0].set_xlim(0, 25)
+axs[0, 0].set_ylim(0, 1)
[…]

### 2025-02-20 code/DeltaNotch_Sim1_TwoCell_CollierRecreate.py
Update DeltaNotch_Sim1_TwoCell_CollierRecreate.py
Adjusted f function input to assume zero boundary condition: D_bar = (D + 0) / 2 = D/2

@@ -21,9 +21,9 @@ def g(N, h, mu):
 # --- ODE System ---
 def ode_system(t, y):
     N1, D1, N2, D2 = y
-    dN1_dt = beta * f(D2, k, theta) - N1
+    dN1_dt = beta * f(D2/2, k, theta) - N1
     dD1_dt = g(N1, h, mu) - D1
-    dN2_dt = beta * f(D1, k, theta) - N2
+    dN2_dt = beta * f(D1/2, k, theta) - N2
     dD2_dt = g(N2, h, mu) - D2
     return [dN1_dt, dD1_dt, dN2_dt, dD2_dt]
 

### 2025-02-20 code/DeltaNotch_Sim1_TwoCell_CollierRecreate.py
Updated model parameter to better align with Collier

@@ -5,7 +5,7 @@ from scipy.integrate import solve_ivp
 # --- Adjusted Parameter Choices ---
 k = 2.0         # Hill exponent for Notch activation
 h = 2.0         # Hill exponent for Delta inhibition
-beta = 1.0      # Production scaling for Notch
+v = 1.0      # Production scaling for Delta
 theta = 0.01    # Threshold for Delta activation of Notch
 mu = 100        # Threshold for Notch inhibition of Delta
 
@@ -21,10 +21,10 @@ def g(N, h, mu):
 # --- ODE System ---
 def ode_system(t, y):
     N1, D1, N2, D2 = y
-    dN1_dt = beta * f(D2/2, k, theta) - N1
-    dD1_dt = g(N1, h, mu) - D1
-    dN2_dt = beta * f(D1/2, k, theta) - N2
-    dD2_dt = g(N2, h, mu) - D2
+    dN1_dt = f(D2/2, k, theta) - N1
+    dD1_dt = v * (g(N1, h, mu) - D1)
+    dN2_dt = f(D1/2, k, theta) - N2
+    dD2_dt = v * (g(N2, h, mu) - D2)
     return [dN1_dt, dD1_dt, dN2_dt, dD2_dt]
 
 # --- Adjusted Simulation Settings ---
@@ -38,7 +38,7 @@ y0 = [1.0, 1.0, 0.99, 0.99]  # Slight asymmetry in D2 initial value
 sol = solve_ivp(ode_system, t_span, y0, t_eval=t_eval, method='RK45', rtol=1e-6, atol=1e-9)
 
 # --- Create the four-panel plot matching the image ---
-fig, axs = plt.subplots(2, 2, figsize=(12, 10))
+fig, axs = plt.subplots(2, 2, figsize=(10, 8))
 
 lw = 1.0  # Line width for the plots
 

### 2025-02-20 code/DeltaNotch_Sim1_TwoCell_CollierRecreate.py
Formatting tweaks

@@ -5,9 +5,9 @@ from scipy.integrate import solve_ivp
 # --- Adjusted Parameter Choices ---
 k = 2.0         # Hill exponent for Notch activation
 h = 2.0         # Hill exponent for Delta inhibition
-v = 1.0      # Production scaling for Delta
-theta = 0.01    # Threshold for Delta activation of Notch
-mu = 100        # Threshold for Notch inhibition of Delta
+v = 1.0         # Production scaling for Delta
+theta = 0.01    # Threshold for Delta activation of Notch ('a' in paper)
+mu = 100        # Threshold for Notch inhibition of Delta ('b' in paper)
 
 # --- Hill Functions ---
 def f(D, k, theta):
@@ -16,7 +16,7 @@ def f(D, k, theta):
 
 def g(N, h, mu):
     """Delta production inhibited by own Notch."""
-    return 1 / (1 + mu*(N)**h)
+    return 1 / (1 + mu * (N)**h)
 
 # --- ODE System ---
 def ode_system(t, y):
@@ -28,11 +28,12 @@ def ode_system(t, y):
     return [dN1_dt, dD1_dt, dN2_dt, dD2_dt]
 
 # --- Adjusted Simulation Settings ---
-t_span = (0, 25)  # Match the x-axis in your plots
+t_span = (0, 25)  
 t_eval = np.linspace(t_span[0], t_span[1], 1000)
 
-# Initial conditions - adjusted to match plot patterns
-y0 = [1.0, 1.0, 0.99, 0.99]  # Slight asymmetry in D2 initial value
+# Initial conditions 
+# y = [N1, D1, N2, D2]
+y0 = [1.0, 1.0, 0.99, 0.99]  # Slight asymmetry in N2,D2 initial values
 
 # --- Solve the ODEs ---
 sol = solve_ivp(ode_system, t_span, y0, t_eval=t_eval, method='RK45', rtol=1e-6, atol=1e-9)
@@ -40,7 +41,8 @@ sol = solve_ivp(ode_system, t_span, y0, t_eval=t_eval, method='RK45', rtol=1e-6,
 # --- Create the four-panel plot matching the image ---
 fig, axs = plt.subplots(2, 2, figsize=(10, 8))
 
-lw = 1.0  # Line width for the plots
+# Line width for the plots
+lw = 1.0 
 
 # Top left: n₁
 axs[0, 0].plot(sol.t, sol.y[0], 'k-', linewidth=lw)

### 2025-02-24 writing/462 Project _Stochastic.tex
Add files via upload

\documentclass{article}

%some useful packages
\usepackage{amsmath,amsthm,amsfonts,amssymb}  %math stuff
\usepackage{graphicx} %to embed images
\usepackage{enumitem} %fancy lists
\usepackage{fullpage} %smaller margins
\usepackage{color} %smaller margins
\usepackage{float}
\usepackage{array} % For better table control
\usepackage{stackrel}
\usepackage{nicematrix}
\NiceMatrixOptions{
code-for-first-row = \color{blue} ,
code-for-first-row = \color{blue} ,
code-for-last-row = \color{blue} ,
code-for-first-col = \color{blue} ,
code-for-last-col = \color{blue}
}

\title{Math 462 Project}
\author{Authors:}
\date{February 2025}

\begin{document}

\maketitle

\section{Introduction}
The numbers of Notch (\(N_P\)) and Delta (\(D_P\)) molecules changing over time due to random reactions motivates us to employ a stochastic model. Each reaction happens with a certain probability per unit time. \\ 
However, we want to show that, in the large-number limit, the expected values \(E[N_P]\) and \(E[D_P]\) follow the deterministic ODEs:

\[
\frac{dN_P}{dt} = F(D_{\text{neighbour},P}) - \mu N_P
\]
\[
\frac{dD_P}{dt} = G(N_P) - \rho D_P
\]

\hfill

We establish a detailed stochastic model for Notch-Delta signalling, ensuring that it converges to the deterministic ODE model under suitable regularity conditions. The system is described by a continuous-time Markov process (CTMC) and then derive the expected value equations to show how it approximates the deterministic model.

\section{Stochastic Model for Notch-Delta Signalling}
\subsection{Chemical Reactions and Propensities}
To construct a probabilistic model, we define the stochastic dynamics of the Delta-Notch interaction as a set of discrete reaction events. We introduce the following reactions:

\begin{align*}
  N_{P} + D_{P} &\xrightarrow{k_1} C_{P} \quad \text{(Binding of Notch and Delta)} \\
  C_{P} &\xrightarrow{k_2} N_{P}^{*} + D_{P} \quad \text{(Activation of Notch)} \\
  C_{P} &\xrightarrow{k_{-1}} N_{P} + D_{P} \quad \text{(Dissociation of the complex)}
\end{align*}

where:

\begin{itemize}
  \item $C_{P}$ represents the Notch-Delta complex in cell $P$.
  \item $k_1$ is the rate at which Notch and Delta bind to form a complex.
  \item $k_2$ is the rate at which the complex activates Notch, producing an intracellular signal.
  \item $k_{-1}$ is the dissociation rate of the complex back into Notch and Delta.
\end{itemize}

Reactions describing Notch and Delta production and degradation  are summarized in the table below:

\begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|} \hline 
         Reaction&  Reaction Equation& Associated term\\ \hline 
         Notch production&  \( N_P \to N_P + 1 \)& \( F(D_{\text{neighbour},P}) \)\\ \hline 
         Notch degradation&  \( N_P \to N_P - 1 \)& \( \mu N_P \)\\ \hline 
         Delta production&  \( D_P \to D_P + 1 \)& \( G(N_P) \)\\ \hline 
         Delta degradation&  \( D_P \to D_P - 1 \)& \( \rho D_P \)\\\hline
    \end{tabular}
    \caption{A summary of Notch and Delta production and degradation reactions}
    \label{1}

\end{table}

where:
\begin{itemize}
    \item \( F(D_{\text{neighbour},P}) \) is an increasing function describing the effect of Delta on Notch production.
    \item \( G(N_P) \) is a decreasing function representing Notch inhibiting Delta.
\end{itemize}

\subsection{Chemical Master Equation and State Probabilities}
Let \( P(N_P, D_P, C_P, t) \) be the probability that the system is in state \( (N_P, D_P, C_P) \) at time \( t \). The probability evolves according to the chemical master equation (CME):

\[
\frac{dP}{dt} = \sum_{\text{all reactions}} \Big[ \text{gain term} - \text{loss term} \Big]
\]

\begin{align*}
  \frac{dP(N_P, D_P, C_P, t)}{dt} &= k_1 (N_P - 1)(D_P - 1) P(N_P - 1, D_P - 1, C_P + 1, t) \\
  &\quad + k_{-1} (C_P + 1) P(N_P + 1, D_P + 1, C_P - 1, t) \\
  &\quad + k_2 (C_P + 1) P(N_P, D_P, C_P - 1, t) \\
  &\quad + F(D_{\text{neighbour},P}) P(N_P - 1, D_P, C_P, t) \\
  &\quad + G(N_P) P(N_P, D_P - 1, C_P, t) \\
  &\quad + \mu (N_P + 1) P(N_P + 1, D_P, C_P, t) \\
  &\quad + \rho (D_P + 1) P(N_P, D_P + 1, C_P, t) \\
  &\quad - (k_1 N_P D_P + k_{-1} C_P + k_2 C_P + \mu N_P + \rho D_P) P(N_P, D_P, C_P, t).
\end{align*}

\subsection{Expected Value Approximation}
To derive the deterministic equations, we compute the expected values:

\[
E[N_P] = \sum_{N_P} N_P P(N_P, D_P, C_P, t)
\]

\[
E[D_P] = \sum_{D_P} D_P P(N_P, D_P, C_P, t)
\]

\[
E[C_P] = \sum_{C_P} C_P P(N_P, D_P, C_P, t)
\]

Taking the time derivative:

\[
\frac{dE[N_P]}{dt} = \sum_{N_P} N_P \frac{dP(N_P, D_P, C_P, t)}{dt}.
\]

Substituting from the CME, we obtain:

\[
\frac{dE[N_P]}{dt} = -k_1 E[N_P]E[D_P] + k_{-1}E[C_P] + k_2 E[C_P] + F(E[D_{\text{neighbour},P}]) - \mu E[N_P].
\]

Similarly, for Delta:

\[
\frac{dE[D_P]}{dt} = -k_1 E[N_P]E[D_P] + k_{-1}E[C_P] + G(E[N_P]) - \rho E[D_P].
\]

For Complexes, we start with the transition terms in the CME.
\begin{align*}
\frac{dP(N_P, D_P, C_P, t)}{dt} = &+ 
 k_1 (N_P - 1)(D_P - 1) P(N_P - 1, D_P - 1, C_P + 1, t)\\
& + k_{-1} (C_P + 1) P(N_P + 1, D_P + 1, C_P - 1, t)\\
& + k_2 (C_P + 1) P(N_P, D_P, C_P - 1, t)
\end{align*}

We compute the expected value for complexes, \( E[C_P] \), and substitute into the time derivative expression:

\[
\frac{dE[C_P]}{dt} = k_1 E[N_P] E[D_P] - (k_{-1} + k_2) E[C_P] - k_{-1}
\]

Rearranging, we get the final form:
\[
\frac{dE[C_P]}{dt} = k_1 E[N_P]E[D_P] - (k_{-1} + k_2) E[C_P].
\]

Using the main equation derived above, we substitute \( dP/dt \) for each reaction.

For Notch production, we expand:

\[
\sum_{N_P} N_P \left[ F(D_{\text{neighbour},P}) P(N_P - 1, D_P, C_P, t) - F(D_{\text{neighbour},P}) P(N_P, D_P, C_P, t) \right]
\]

Shifting the sum over \( N_P \) gives:

\[
\sum_{N_P} (N_P + 1) F(D_{\text{neighbour},P}) P(N_P, D_P, C_P, t) - \sum_{N_P} N_P F(D_{\text{neighbour},P}) P(N_P, D_P, C_P, t)
\]

Rearranging:

\[
\sum_{N_P} F(D_{\text{neighbour},P}) P(N_P, D_P, C_P, t) = E[F(D_{\text{neighbour},P})]
\]

Thus, the equation simplifies to:

\[
\frac{dE[N_P]}{dt} = E[F(D_{\text{neighbour},P})] - \mu E[N_P]
\]

Similarly, for Delta:

\[
\frac{dE[D_P]}{dt} = E[G(N_P)] - \rho E[D_P]
\]

\subsection{The Mean-Field Approximation}
In general, expected values of nonlinear functions like \( E[F(D_P)] \) are not necessarily equal to \( F(E[D_P]) \). However, under large-number assumptions, fluctuations around the mean become small (due to the law of large numbers), and we approximate:

\[
E[F(D_{\text{neighbour},P})] \approx F(E[D_{\text{neighbour},P}])
\]

\[
E[G(N_P)] \approx G(E[N_P])
\]

This assumes that each individual cell’s fluctuations average out in large populations.

\subsection{Deterministic Model Recovery}

Substituting these approximations into our equations:

\[
\frac{dE[N_P]}{dt} = F(E[D_{\text{neighbour},P}]) - \mu E[N_P]
\]

\[
\frac{dE[D_P]}{dt} = G(E[N_P]) - \rho E[D_P]
\]

Thus, we recover the deterministic system:

\[
\frac{dN_P}{dt} = F(D_{\text{neighbour},P}) - \mu N_P
\]

\[
\frac{dD_P}{dt} = G(N_P) - \rho D_P
\]

The calculation confirms that in the large-number limit, the stochastic model converges to the deterministic ordinary differential equation (ODE) system. This implies that as the molecular count increases, transitioning the system to a more macroscopic scale, the stochastic model naturally aligns with the deterministic framework.

\subsection{Convergence to the Deterministic Model}

To show convergence, define fluctuations:

\[
X = N_P - E[N_P], \quad Y = D_P - E[D_P], \quad Z = C_P - E[C_P].
\]

Taking expectations:

\[
\frac{dE[X]}{dt} = -\mu E[X], \quad \frac{dE[Y]}{dt} = -\rho E[Y], \quad \frac{dE[Z]}{dt} = -(k_{-1} + k_2)E[Z].
\]

These decay exponentially, meaning that as time progresses:

\[
E[X], E[Y], E[Z] \to 0.
\]

Thus, the stochastic model converges to the deterministic ODE model as molecule numbers increase.

\end{document}

### 2025-02-25 writing/notes.tex
Write down a probabilistic model under constant delta/notch assumptions.

@@
 
-\usepackage[margin=1.5in]{geometry}
+\usepackage[left=1.5in, right=1.5in, top=1in, bottom=1in]{geometry}
 \usepackage{amsmath}
+\usepackage{amssymb}
 
@@
 
-\subsection{Writing Down a Dimensional Model}
+\medskip
 
-Let us continue to use $N_{P}$, $D_{P}$, and $\overline{D}_{P}$ as defined above.
-Let $t$ denote the dimensional time and let $u$ and $r$ be the dimensional decay rates for notch and delta.
-Then, assuming the dimensional analogues of $F$ and $G$ are hill functions of order $1$, we have:
+\textbf{Remark:} The notch activity can also be taken to be the number of complexes formed by the binding of notch to its activated ligand delta (Collier et al. 1996).
+
+\newpage
+
+\subsection{Deriving a Probabilistic Model}
+
+We will make the assumption that each cell has $N$ notch molecules (which can be in a complex, or not).
+As we did above, define $\overline{D}$ to be the average number of active delta molecules in neighbouring cells.
+Let $k_{b}$ be the binding rate and $k_{u}$ the unbinding rate.
+Let $p_{n}(t)$ be the probability there are $n$ complexes at time $t$.
+Then, define the Kolmogorov forward equation:
 
@@
 \begin{aligned}
-  \frac{dN_{P}}{dt} &= \frac{\overline{D}_{P}}{d_{0} + \overline{D}_{P}} - u N_{P} \\[5pt]
-  \frac{dD_{P}}{dt} &= \frac{n_{0}}{n_{0} + N_{P}} - rD_{P}
+  p_{n}(t + \Delta t)= &\,\,p_{n-1}(t) \cdot \underbrace{(N - (n - 1))(\overline{D} - (n - 1)) k_{b} \Delta t}_{\text{P(binding)}}  \\[5pt]
+                      &+ p_{n}(t) \cdot \underbrace{(1 - (N - n)(\overline{D} - n)k_{b}\Delta t)}_{\text{P(no binding)}} \cdot \underbrace{(1 - n k_{u} \Delta t)}_{\text{P(no unbinding)}} \\[5pt]
+                      &+ p_{n+1}(t) \cdot \underbrace{(n + 1)k_{u}\Delta t}_{\text{P(unbinding)}}
 \end{aligned}
@@
 
-In the equation above, $d_{0}$ is the half-max of notch production induced by delta in neighbouring cells while $n_{0}$ is the half-min of delta production induced by notch.
+Rearranging this equation (and cancelling out the term with $\Delta t^2$) we get:
 
-\medskip
+$$
+\begin{aligned}
+p_{n}'(t) = &\,\,p_{n-1}(t) \cdot (N - (n - 1))(\overline{D} - (n - 1)) k_{b} \\[5pt]
+            &-p_{n}(t)\cdot (N-n)(\overline{D} - n)k_{b} - p_{n}(t) \cdot n k_{u} \\[5pt]
+            &+ p_{n+1}(t) \cdot (n + 1)k_{u}
+\end{aligned}
+$$
 
-This model is a good start, but we will need to explicitly model the concentration of delta-notch complexes $C$ in order to write down a probabilistic model.
-Since $D_{P}$ and $N_{P}$ denote the level of delta and notch \emph{activation} (that is, the concentration of complexes),
+Let $C(t)$ be the number of complexes at time $t$. From the equation above, we have:
 
-\subsection{Deriving a Probabilistic Model}
+$$
+y(t) = \mathbb{E}[C(t)] = \sum_{n = 0}^{N} np_{n}(t) \Rightarrow y'(t) = \sum_{n = 0}^{N} np_{n}'(t)
+$$
 
-TBD
+Now, we substitute $p_{n}'(t)$ for the expression derived above:
+
+$$
+\begin{aligned}
+  y'(t) = &\,\,\sum_{n = 2}^{N}  np_{n-1}(t)(N - (n - 1))(\overline{D} - (n - 1)) k_{b} \\[5pt]
+          &-  \sum_{n = 1}^{N - 1} np_{n}(t) (N - n)(\overline{D} - n )k_{b} -  \sum_{n = 1}^{N}  n^2 p_{n}(t) k_{u} \\[5pt]
+          &- \sum_{n = 1}^{N - 1}  np_{n+1}(t) (n+1)k_{u}
+\end{aligned}
+$$
+
+We will need to reindex the $p_{n-1}$ and $p_{n+1}$ terms. To do this, begin by rewriting:
+
+$$
+\begin{aligned}
+  y'(t) = &\,\,\sum_{n = 2}^{N}  (n-1 + 1)p_{n-1}(t)(N - (n - 1))(\overline{D} - (n - 1)) k_{b} \\[5pt] 
+          &-  \sum_{n = 1}^{N - 1} np_{n}(t) (N - n)( \overline{D} - n)k_{b} -  \sum_{n = 1}^{N}  n^2 p_{n}(t) k_{u} \\[5pt]
+          &- \sum_{n = 1}^{N - 1}  (n + 1 - 1)p_{n+1}(t) (n+1)k_{u}
+\end{aligned}
+$$
+
+Now, perform reindexing on the $p_{n- 1}$ and $p_{n+1}$ terms:
+
+$$
+\begin{aligned}
+y'(t) = &\,\,\sum_{n = 1}^{N - 1} (n + 1)p_{n}(t)(N - n) (\overline{D} - n)k_{b} \\[5pt]
+        &-  \sum_{n = 1}^{N - 1} np_{n}(t) (N - n)(\overline{D} - n)k_{b} -  \sum_{n = 1}^{N}  n^2 p_{n}(t) k_{u} \\[5pt]
+        &+  \sum_{n = 2}^{N}  (n - 1 )p_{n}(t) nk_{u} 
+\end{aligned}
+$$
+
+Now, we cancel out like terms to get the following:
+
+$$
+y'(t) = \sum_{n = 1}^{N} p_{n}(t)(N - n)(\overline{D} - n)k_{b} - \sum_{n = 1}^{N} np_{n}(t)  k_{u}
+$$
+
+Therefore,
+
+$$
+y'(t) = (N - y)(\overline{D} - y)k_{b} -  yk_{u}
+$$
+
+Note that the $(N - y)(\overline{D} - y)$ term is an increasing function of $\overline{D}$, which means that it satisfies the conditions imposed by the Collier et al.
+(1996).
+However, it is not a hill function, which means the authors made different assumptions about the system.
 

### 2025-02-25 writing/462 Project _Stochastic.tex
Add feedback on the stochastic model write-up.

@@
 
+\medskip
+
+% COMMENTS BEGIN HERE
+\color{red}
+
+Shouldn't the binding equation use $\overline{D}_{P}$ (i.e.
+the delta concentration in neighbouring cells) instead of $D_{P}$?
+Also, I think this would be a good place to state the assumptions you used to derive these equations.
+The start of section "2.
+The Model" in Collier et al.
+contains some more information about interpreting $N$ and $D$ that might be useful to read through.
+
+\medskip
+
+I'm also unsure about the activation equation, because it implies that a complex is splitting to form an $N_{P}^{*}$ (what makes this different from $N_{P}$, by the way?) and a $D_{P}$.
+I think something like $C_{P} \xrightarrow{k_{2}} N_{P}^{*} + D_{P}$ might be more accurate.
+
+\color{black}
+% COMMENTS END HERE
+
+\medskip
+
 where:
@@
 
+% COMMENTS BEGIN HERE
+\color{red}
+
+I think there are a couple mistakes in this equation.
+For example, in the first term, we're looking at the probability of a binding event, which creates one $C_{P}$ and destroys a $N_{P}$ and a $D_{P}$.
+Therefore, I believe we need to be in state $(N_{P} + 1, D_{P} + 1, C_{P} - 1)$  in order for a binding event to take us to state $(N_{P}, D_{P}, C_{P})$.
+I think the first term should be the following, although I could be mistaken.
+
+$$
+k_{1} (N_{P} + 1)(D_{P} + 1) P(N_{P} + 1, D_{P} + 1, C_{P} - 1, t)
+$$
+
+\medskip
+
+I think some of the other terms have similar issues. I'm also still unsure about the difference between the unbinding ($k_{-1}$) and activation ($k_{2}$) reactions. What assumptions are you making?
+
+\color{black}
+% COMMENTS END HERE
+
 \subsection{Expected Value Approximation}
@@
 
+% COMMENTS BEGIN HERE
+\medskip
+\color{red}
+This looks good and makes sense to me.
+\color{black}
+\medskip
+% COMMENTS END HERE
+
 Substituting from the CME, we obtain:
@@
 \]
+
+% COMMENTS BEGIN HERE
+\medskip
+\color{red}
+I'm not sure how you got to here, but the equation makes sense.
+\color{black}
+\medskip
+% COMMENTS END HERE
 
@@
 
+% COMMENTS BEGIN HERE
+\medskip
+\color{red}
+Isn't this just the previous equation without the $k_{-1}$? I also think you're missing an $E[C_{P}]$:
+
+$$
+\frac{dE[C_{P}]}{dt} = k_{1}E[N_{P}]E[D_{P}] - (k_{-1} + k_{2})E[C_{P}] - k_{-1}E[C_{P}]
+$$
+
+and then you can do some simple algebra to get rid of the last term.
+\color{black}
+\medskip
+% COMMENTS END HERE
+
 Using the main equation derived above, we substitute \( dP/dt \) for each reaction.
@@
 \]
+
+% COMMENTS BEGIN HERE
+\medskip
+\color{red}
+Where does this equation come from?
+\color{black}
+\medskip
+% COMMENTS END HERE
 

### 2025-03-04 writing/notes.tex
Update notes.tex with some background information.

@@
 \usepackage{amssymb}
+\usepackage{hyperref}
 
@@
 \begin{flushleft}
+
+\section{Background Information}
+
+Sources: \href{https://www.pnas.org/doi/full/10.1073/pnas.1416287112}{Boareto et al., 2015}, \href{https://www.nature.com/articles/nrm2009}{Bray, 2006}
+
+\medskip
+
+Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor.
+This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch and \emph{inhibits} Delta.
+
+\begin{itemize}
+  \item The Serrate (Jagged) ligand can also bind to Notch. Unlike Delta, NICD promotes Jagged, which creates a double-positive feedback loop.
+  \item Interaction between Delta ligands and Notch receptors from the same cell (cis-interaction) removes both molecules and does not trigger the release of NICD.
+\end{itemize}
 

### 2025-03-05 writing/notes.tex
Update notes.tex with a new proposal for a model.

--- 2025-02-19 writing/notes.tex
@@
 
-\usepackage[margin=1.5in]{geometry}
+\usepackage[left=1.5in, right=1.5in, top=1in, bottom=1in]{geometry}
 \usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{hyperref}
 
@@
 
-\section{Collier et al., 1996}
+\section{Background Information}
+
+Sources: \href{https://www.pnas.org/doi/full/10.1073/pnas.1416287112}{Boareto et al., 2015}, \href{https://www.nature.com/articles/nrm2009}{Bray, 2006}
+
+\medskip
+
+Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor.
+This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch and \emph{inhibits} Delta.
+
+\begin{itemize}
+  \item The Serrate (Jagged) ligand can also bind to Notch. Unlike Delta, NICD promotes Jagged, which creates a double-positive feedback loop.
+  \item Interaction between Delta ligands and Notch receptors from the same cell (cis-interaction) removes both molecules and does not trigger the release of NICD.
+\end{itemize}
+
+\section{Ideas and Issues with Collier et al., 1996}
+
+\subsection{Model}
 
@@
 
-Define the following variables:
+Then they define the following variables:
 
@@
 
-\subsection{Writing Down a Dimensional Model}
+\newpage
 
-Let us continue to use $N_{P}$, $D_{P}$, and $\overline{D}_{P}$ as defined above.
-Let $t$ denote the dimensional time and let $u$ and $r$ be the dimensional decay rates for notch and delta.
-Then, assuming the dimensional analogues of $F$ and $G$ are hill functions of order $1$, we have:
+\subsection{Issues}
+
+The Collier et al.
+model makes numerous simplifications that violate our current understanding of the delta-notch signalling network.
+These simplifications make it difficult to derive a probabilistic model from the Collier et al.
+model, since the biological implications of such a model would be contradictory.
+Some of the issues include:
+
+\begin{itemize}
+  \item Production of activated Notch (NICD) is exclusively a function of the Delta concentration in neighbouring cells, with no regard for the concentration of extracellular Notch receptors.
+  \item Furthermore, the concentration of extracellular Notch isn't accounted for at all, which means there is no way to create a meaningful reaction term that captures the dynamics of the system.
+  \item There are smaller issues, like the absence of cis-inhibition and the Serrate (Jagged) protein, but these aren't necessary to capture the fundamental behaviour of the system.
+\end{itemize}
+
+\subsection{Qualitative Results}
+
+In the two-cell periodic boundary condition case, Collier et al.
+show that there always exists a homogeneous steady state, although this state is unstable if the feedback in the Delta-Notch signalling network is sufficiently strong.
+If the line of cells is given Dirichlet boundary conditions, the inhomogeneity arises on the outside edges before gradually spreading inwards to the center of the domain.
+On the two-dimensional hexagonal lattice, the system exhibits different ratios of primary to secondary fated cells depending on the parameters.
+In particular, Collier et al.
+observed ratios of $1/2$, $1/3$, $1/4$, $1/5$, and $1/6$.
+
+\section{Probabilistic Models of Increasing Complexity}
+
+\textbf{Idea}: Our first goal is to develop a probabilistic model based on a modern understanding of the biochemistry of the delta-notch signalling network.
+Then, we will use this model to attempt to replicate qualitative results from Collier et al.
+(1996).
+
+\subsection{A First Model}
+
+Assume \emph{a priori} (i.e. from previous research) the following, where $I$ is NICD:
+
+\begin{itemize}
+  \item Basal extracellular Notch ($N$) production has the form $H^{+}(I) = N_{0}I^2/(1 + I^2)$. 
+  \item Basal extracellular Delta ($D$) production has the form $H^{-}(I) = D_{0}/(1 + I^2)$.
+  \item Extracellular Notch and Delta have the same decay rate, $\gamma$.
+\end{itemize}
+
+Then, define the following parameters:
+
+\begin{itemize}
+  \item $N_{0}$ and $D_{0}$ are the maximum rates of Notch/Delta production.
+  \item $k_{T}$ is the binding rate between extracellular Notch and Delta.
+  \item $N_{ext}$ and $D_{ext}$ refer to the average Notch/Delta levels in neighbouring cells.
+  \item $\gamma_{I}$ is the decay rate of NICD.
+\end{itemize}
+
+\newpage
+
+Furthermore, we will assume that there is no cis-interaction and no Serrate ligand. This gives us the following set of chemical reactions that can occur in a single time step:
+
+\begin{enumerate}
+  \item A Notch receptor (from cell $A$) binds to a Delta ligand (from cell $B$), triggering the release of NICD in cell $A$.
+This can be written as $N_{A} + D_{B} \rightarrow I_{A}$.
+The probability of this event is $k_{T}N_{A}D_{B}$.
+  \item A Notch receptor in cell $A$ decays ($N_{A} \rightarrow \emptyset$) with probability $\gamma N_{A}$.
+  \item A Delta ligand in cell $A$ decays ($D_{A} \rightarrow \emptyset$) with probability $\gamma D_{A}$.
+  \item A NICD molecule in cell $A$ decays ($I_{A} \rightarrow \emptyset$) with probability $\gamma_{I}I_{A}$.
+  \item A Notch receptor in cell $A$ is produced ($\emptyset \rightarrow N_{A}$) with probability $H^{+}(I)$.
+  \item A Delta ligand in cell $A$ is produced ($\emptyset \rightarrow D_{A}$) with probability $H^{-}(I)$.
+\end{enumerate}
+
+\subsection{Computational Implementation}
+
+We will make the assumption that reaction times follow a Poisson point process.
+Let $E$ be the set of all possible reaction events (i.e.
+reactions 1-6 above for all cells).
+Furthermore, let $|E| = N$ be the number of possible reactions.
+For each event, let $X_{i} \sim \text{Exp}(\lambda_{i})$ denote the time until the next event of type $i$, where $1 \leq i \leq N$.
+Then, the waiting time $T$ in between \emph{any} two events is:
+
+$$
+T = \text{min} \{ X_{1}, \dots, X_{N} \} \sim \text{Exp}\left( \sum_{i = 1}^{N} \lambda_{i} \right) 
+$$
+
+This results follows from the definition of an exponential distribution and the memoryless property. Then, we can use Gillespie's algorithm to simulate the system as follows:
+
+\medskip
+
+\textbf{Step 1}: Sample a value $t_{i}$ from $T$ using inverse transform sampling.
+
+\medskip
+
+\textbf{Step 2}: Increment the current time $t$ by $t_{i}$.
+
+\medskip
+
+\textbf{Step 3}: Determine which event took place. To do this, recall that:
+
+$$P(X_{k} = \text{min} \{  X_{1}, \dots, X_{N} \}) = \frac{\lambda_{k}}{\lambda_{1} + \dots + \lambda_{N}}$$
+
+Use this fact to partition the interval $[0, 1]$. Then sample $y_{i}$ from a $\text{Unif}(0, 1)$ distribution, identify its partition, and find the corresponding event.
+
+\medskip
+
+\textbf{Step 4}: Update the state based on the event determined in the previous step.
+
+\subsection{Theoretical Justification}
+
+In this section, we need to show that in expectation, the model above converges to the following system of ODEs (adapted from Boareto et al., 2015):
 
@@
 \begin{aligned}
-  \frac{dN_{P}}{dt} &= \frac{\overline{D}_{P}}{d_{0} + \overline{D}_{P}} - u N_{P} \\[5pt]
-  \frac{dD_{P}}{dt} &= \frac{n_{0}}{n_{0} + N_{P}} - rD_{P}
+  \frac{dN}{dt} &= H^{+}(I) - k_{T}ND_{ext} - \gamma N \\[5pt]
+  \frac{dD}{dt} &= H^{-}(I) - k_{T}DN_{ext} - \gamma D \\[5pt]
+  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
 \end{aligned}
 $$
-
-In the equation above, $d_{0}$ is the half-max of notch production induced by delta in neighbouring cells while $n_{0}$ is the half-min of delta production induced by notch.
-
-\medskip
-
-This model is a good start, but we will need to explicitly model the concentration of delta-notch complexes $C$ in order to write down a probabilistic model.
-Since $D_{P}$ and $N_{P}$ denote the level of delta and notch \emph{activation} (that is, the concentration of complexes),
-
-\subsection{Deriving a Probabilistic Model}
-
-TBD
 

### 2025-03-07 code/gillespie-model/main.py
First attempt at the stochastic model with the Gillespie algorithm.

@@ -1,0 +1,179 @@
+# IMPORT PYTHON MODULES
+import matplotlib.pyplot as plt
+import numpy as np
+import logging
+
+# INITIALIZE PARAMETER VALUES
+KT = 0.0002  # Delta-Notch binding rate
+GE = 0.01    # Extracellular molecule decay rate (used in dN/dt and dD/dt equations)
+GI = 0.05    # NICD decay rate
+NM = 10      # Maximum rate of Notch production
+DM = 10      # Maximum rate of Delta production
+N0 = 100     # Notch hill function (H+) half-max
+D0 = 100     # Delta hill function (H-) half-max
+
+# INITIALIZE HILL FUNCTIONS
+HP = lambda i : (NM * i**2) / ((N0**2) + (i**2))
+HM = lambda i : (DM * D0**2) / ((D0**2) + (i**2))
+
+# INITIALIZE NEIGHBOURS (by index in state, include both directions)
+NEIGHBOURS = np.array([
+    [0, 1],
+    [1, 0],
+])
+
+# INITIALIZE SYSTEM STATE (each element contains integers [N, D, I] for one cell)
+state = np.array([
+    [300, 50, 100],
+    [300, 50, 100],
+])
+
+# INITIALIZE SYSTEM RATES (ri contains rates for reaction i)
+#  - Each pair of neighbours creates one reaction
+#  - Each cell has five other reactions (decay/production)
+rates = np.empty(len(NEIGHBOURS) + (len(state) * 5))
+
+# INITIALIZE SIMULATION TIME
+t = 0
+
+
+# Generates the reaction rates based on the current state
+def update_reaction_rates(state):
+
+    # Initialize empty rate arrays for each reaction
+    r1 = np.empty(len(NEIGHBOURS))
+    r2 = np.empty(len(state))
+    r3 = np.empty(len(state))
+    r4 = np.empty(len(state))
+    r5 = np.empty(len(state))
+    r6 = np.empty(len(state))
+
+    # Iterate over all cells to get the reaction/decay rates
+    for i, cell in enumerate(state):
+
+        # Set all of the decay rates (reactions 2-4)
+        r2[i] = cell[0] * GE  # Rate parameter of Notch decay
+        r3[i] = cell[1] * GE  # Rate parameter of Delta decay
+        r4[i] = cell[2] * GI  # Rate parameter of NICD decay
+
+        # Set all of the production rates (reactions 5-6)
[…]

### 2025-03-09 code/new-ode-model/NewModel_TwoCell_Deterministic.py
New model ODE simulations

@@ -1,0 +1,76 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import solve_ivp
+
+# --- Adjusted Parameter Choices ---
+k = 2.0         # Hill exponent for Notch activation
+h = 2.0         # Hill exponent for Delta inhibition
+v = 1.0         # Production scaling for Delta
+theta = 0.01    # Threshold for Delta activation of Notch ('a' in paper)
+mu = 100        # Threshold for Notch inhibition of Delta ('b' in paper)
+
+# --- Hill Functions ---
+def f(D, k, theta):
+    """Notch activation by Delta from neighboring cell."""
+    return D**k / (theta + D**k)
+
+def g(N, h, mu):
+    """Delta production inhibited by own Notch."""
+    return 1 / (1 + mu * (N)**h)
+
+# --- ODE System ---
+def ode_system(t, y):
+    N1, D1, N2, D2 = y
+    dN1_dt = f(D2/2, k, theta) - N1
+    dD1_dt = v * (g(N1, h, mu) - D1)
+    dN2_dt = f(D1/2, k, theta) - N2
+    dD2_dt = v * (g(N2, h, mu) - D2)
+    return [dN1_dt, dD1_dt, dN2_dt, dD2_dt]
+
+# --- Adjusted Simulation Settings ---
+t_span = (0, 25)  
+t_eval = np.linspace(t_span[0], t_span[1], 1000)
+
+# Initial conditions 
+# y = [N1, D1, N2, D2]
+y0 = [1.0, 1.0, 0.99, 0.99]  # Slight asymmetry in N2,D2 initial values
+
+# --- Solve the ODEs ---
+sol = solve_ivp(ode_system, t_span, y0, t_eval=t_eval, method='RK45', rtol=1e-6, atol=1e-9)
+
+# --- Create the four-panel plot matching the image ---
+fig, axs = plt.subplots(2, 2, figsize=(10, 8))
+
+# Line width for the plots
+lw = 1.0 
+
+# Top left: n₁
+axs[0, 0].plot(sol.t, sol.y[0], 'k-', linewidth=lw)
+axs[0, 0].set_xlim(0, 25)
[…]

### 2025-03-09 code/new-ode-model/NewModel_TwoCell_VaryInitialConditions.py

@@ -1,0 +1,103 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import solve_ivp
+
+# --- Revised Parameter Choices ---
+k = 2.0       # Hill exponent for Notch activation
+h = 2.0       # Hill exponent for Delta inhibition
+beta = 1.5    # Production scaling for Notch; >1 to promote instability
+a = 0.01   # Threshold for Delta activation of Notch
+b = 100      # Threshold for Notch inhibition of Delta
+
+# --- Revised Hill Functions ---
+def f(D, k, a):
+    """
+    Notch activation by Delta from the neighboring cell.
+    A threshold parameter theta sharpens the response.
+    """
+    return D**k / (a + D**k)
+
+def g(N, h, b):
+    """
+    Delta production inhibited by own Notch.
+    The parameter mu sets the inhibition threshold.
+    """
+    return 1 / (1 + b * N**h)
+
+# --- Define the Revised ODE System ---
+def ode_system(t, y):
+    # y = [N1, D1, N2, D2] for the two-cell system.
+    N1, D1, N2, D2 = y
+    dN1_dt = beta * f(D2, k, a) - N1
+    dD1_dt = g(N1, h, b) - D1
+    dN2_dt = beta * f(D1, k, a) - N2
+    dD2_dt = g(N2, h, b) - D2
+    return [dN1_dt, dD1_dt, dN2_dt, dD2_dt]
+
+# --- Simulation Settings ---
+t_span = (0, 50)
+t_eval = np.linspace(t_span[0], t_span[1], 1000)
+
+# Number of different initial conditions to simulate
+num_simulations = 10
+
+# Prepare an array to store each solution (shape: num_simulations x 4 x len(t_eval))
+all_solutions = []
+
+# Loop over simulations with different starting positions.
+for i in range(num_simulations):
+    # Create a random perturbation about the baseline initial condition.
+    # Baseline: [0.1, 0.9, 0.15, 0.85]
+    y0 = [
+        0.91 + np.random.uniform(-0.05, 0.05),
+        0.9 + np.random.uniform(-0.05, 0.05),
+        0.89 + np.random.uniform(-0.05, 0.05),
+        0.85 + np.random.uniform(-0.05, 0.05)
+    ]
[…]

### 2025-03-09 code/new-ode-model/NewModel_TwoCell_Deterministic.py
Correct versions

@@ -2,43 +2,57 @@ import numpy as np
 import matplotlib.pyplot as plt
 from scipy.integrate import solve_ivp
 
-# --- Adjusted Parameter Choices ---
-k = 2.0         # Hill exponent for Notch activation
-h = 2.0         # Hill exponent for Delta inhibition
-v = 1.0         # Production scaling for Delta
-theta = 0.01    # Threshold for Delta activation of Notch ('a' in paper)
-mu = 100        # Threshold for Notch inhibition of Delta ('b' in paper)
+# --- Initialize Parameter Values ---
+k = 2           # Hill exponent for Notch activation
+NM = 10         # Maximum rates of Notch production
+DM = 10         # Maximum rates of Delta production
+N0 = 100        # Delta Hill function
+D0 = 100        # Delta Hill function
+KT = 0.0001     # Binding rate between estracellular Delta and Notch
+G = 0.01        # Decay rate for Delta and Notch
+GI = 0.025       # Decay rate for NICD
 
 # --- Hill Functions ---
-def f(D, k, theta):
-    """Notch activation by Delta from neighboring cell."""
-    return D**k / (theta + D**k)
+def H_plus(I):
+    """Notch activation by intracellular NICD."""
+    return (NM * I**k) / ((N0**k) + (I**k))
 
-def g(N, h, mu):
-    """Delta production inhibited by own Notch."""
-    return 1 / (1 + mu * (N)**h)
+def H_minus(I):
+    """Delta inhibited by intracellular NICD."""
+    return (DM * D0**k) / ((D0**k) + (I**k))
 
 # --- ODE System ---
 def ode_system(t, y):
-    N1, D1, N2, D2 = y
-    dN1_dt = f(D2/2, k, theta) - N1
-    dD1_dt = v * (g(N1, h, mu) - D1)
-    dN2_dt = f(D1/2, k, theta) - N2
-    dD2_dt = v * (g(N2, h, mu) - D2)
[…]

### 2025-03-09 code/new-ode-model/NewModel_TwoCell_VaryInitialConditions.py

@@ -1,103 +1,170 @@
 import numpy as np
 import matplotlib.pyplot as plt
 from scipy.integrate import solve_ivp
+import matplotlib.cm as cm
 
-# --- Revised Parameter Choices ---
-k = 2.0       # Hill exponent for Notch activation
-h = 2.0       # Hill exponent for Delta inhibition
-beta = 1.5    # Production scaling for Notch; >1 to promote instability
-a = 0.01   # Threshold for Delta activation of Notch
-b = 100      # Threshold for Notch inhibition of Delta
+# --- Initialize Parameter Values ---
+k = 2           # Hill exponent for Notch activation
+NM = 10         # Maximum rates of Notch production
+DM = 10         # Maximum rates of Delta production
+N0 = 100        # Delta Hill function
+D0 = 100        # Delta Hill function
+KT = 0.0001     # Binding rate between extracellular Delta and Notch
+G = 0.01        # Decay rate for Delta and Notch
+GI = 0.025      # Decay rate for NICD
 
-# --- Revised Hill Functions ---
-def f(D, k, a):
-    """
-    Notch activation by Delta from the neighboring cell.
-    A threshold parameter theta sharpens the response.
-    """
-    return D**k / (a + D**k)
+# --- Hill Functions ---
+def H_plus(I):
+    """Notch activation by intracellular NICD."""
+    return (NM * I**k) / ((N0**k) + (I**k))
 
-def g(N, h, b):
-    """
-    Delta production inhibited by own Notch.
-    The parameter mu sets the inhibition threshold.
-    """
-    return 1 / (1 + b * N**h)
+def H_minus(I):
+    """Delta inhibited by intracellular NICD."""
+    return (DM * D0**k) / ((D0**k) + (I**k))
 
-# --- Define the Revised ODE System ---
+# --- ODE System ---
 def ode_system(t, y):
-    # y = [N1, D1, N2, D2] for the two-cell system.
-    N1, D1, N2, D2 = y
-    dN1_dt = beta * f(D2, k, a) - N1
-    dD1_dt = g(N1, h, b) - D1
-    dN2_dt = beta * f(D1, k, a) - N2
-    dD2_dt = g(N2, h, b) - D2
[…]

### 2025-03-09 code/gillespie-model/main.py
Do bifurcation analysis.

@@ -4,17 +4,18 @@ import numpy as np
 import logging
 
 # INITIALIZE PARAMETER VALUES
-KT = 0.0002  # Delta-Notch binding rate
-GE = 0.01    # Extracellular molecule decay rate (used in dN/dt and dD/dt equations)
-GI = 0.05    # NICD decay rate
-NM = 10      # Maximum rate of Notch production
-DM = 10      # Maximum rate of Delta production
+KT = 0.0001  # Delta-Notch binding rate
+GE = 0.002   # Extracellular molecule decay rate (used in dN/dt and dD/dt equations)
+GI = 0.01    # NICD decay rate
+NM = 5       # Maximum rate of Notch production
+DM = 5       # Maximum rate of Delta production
 N0 = 100     # Notch hill function (H+) half-max
 D0 = 100     # Delta hill function (H-) half-max
+N = 2        # Hill function parameter
 
 # INITIALIZE HILL FUNCTIONS
-HP = lambda i : (NM * i**2) / ((N0**2) + (i**2))
-HM = lambda i : (DM * D0**2) / ((D0**2) + (i**2))
+HP = lambda i : (NM * i**N) / ((N0**N) + (i**N))
+HM = lambda i : (DM * D0**N) / ((D0**N) + (i**N))
 
 # INITIALIZE NEIGHBOURS (by index in state, include both directions)
 NEIGHBOURS = np.array([
@@ -24,8 +25,8 @@ NEIGHBOURS = np.array([
 
 # INITIALIZE SYSTEM STATE (each element contains integers [N, D, I] for one cell)
 state = np.array([
-    [300, 50, 100],
-    [300, 50, 100],
+    [500, 50, 200],
+    [500, 50, 200],
 ])
 
 # INITIALIZE SYSTEM RATES (ri contains rates for reaction i)
@@ -57,8 +58,8 @@ def update_reaction_rates(state):
         r4[i] = cell[2] * GI  # Rate parameter of NICD decay
 
         # Set all of the production rates (reactions 5-6)
-        r5[i] = HP(cell[0])   # Rate parameter of Notch production
-        r6[i] = HM(cell[1])   # Rate parameter of Delta production
+        r5[i] = HP(cell[2])   # Rate parameter of Notch production
+        r6[i] = HM(cell[2])   # Rate parameter of Delta production
 
[…]

### 2025-03-09 code/new-ode-model/NewModel_TwoCell_Deterministic.py
Param tweak (G = 0.02) and new cell line visualization

@@ -4,12 +4,12 @@ from scipy.integrate import solve_ivp
 
 # --- Initialize Parameter Values ---
 k = 2           # Hill exponent for Notch activation
-NM = 10         # Maximum rates of Notch production
+NM = 15         # Maximum rates of Notch production
 DM = 10         # Maximum rates of Delta production
 N0 = 100        # Delta Hill function
 D0 = 100        # Delta Hill function
 KT = 0.0001     # Binding rate between estracellular Delta and Notch
-G = 0.01        # Decay rate for Delta and Notch
+G = 0.02        # Decay rate for Delta and Notch
 GI = 0.025       # Decay rate for NICD
 
 # --- Hill Functions ---
@@ -43,8 +43,8 @@ t_eval = np.linspace(t_span[0], t_span[1], 1000)
 
 # Initial conditions 
 # C = [N, D, I]
-C1 = [200, 200, 100]  # Initial conditions for cell 1
-C2 = [300, 150, 120]  # Initial conditions for cell 2
+C1 = [280, 200, 100]  # Initial conditions for cell 1
+C2 = [300, 180, 120]  # Initial conditions for cell 2
 
 # y = [N1, D1, I1, N2, D2, I2]
 y0 = C1 + C2  # Concatenate the two cell initial conditions

### 2025-03-09 writing/notes.tex

@@
 \usepackage{hyperref}
+\usepackage{graphicx}
 
@@
 \begin{itemize}
-  \item Basal extracellular Notch ($N$) production has the form $H^{+}(I) = N_{0}I^2/(1 + I^2)$. 
-  \item Basal extracellular Delta ($D$) production has the form $H^{-}(I) = D_{0}/(1 + I^2)$.
+  \item Basal extracellular Notch ($N$) production has the form $H^{+}(I) = n_{m}I^2/(n_{0}^2 + I^2)$. 
+  \item Basal extracellular Delta ($D$) production has the form $H^{-}(I) = d_{m}/(d_{0}^2 + I^2)$.
   \item Extracellular Notch and Delta have the same decay rate, $\gamma$.
@@
 \begin{aligned}
-  \frac{dN}{dt} &= H^{+}(I) - k_{T}ND_{ext} - \gamma N \\[5pt]
-  \frac{dD}{dt} &= H^{-}(I) - k_{T}DN_{ext} - \gamma D \\[5pt]
+  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\[5pt]
+  \frac{dD}{dt} &= \frac{d_{m}}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\[5pt]
   \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
@@
 
+TBD
+
+\subsection{Bifurcation Analysis}
+
+We perform a bifurcation analysis for the two-cell system, where $D_{ext}$ and $N_{ext}$ are treated as parameters.
+In other words, we will examine the steady states for one cell as a function of the Notch and Delta levels of the other cell.
+For simplicity, assume all parameters other than $D_{ext}, N_{ext}$ are fixed.
+
+$$
+\begin{aligned}
+  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - (k_{T}D_{ext} + \gamma)N \\[5pt]
+  \frac{dD}{dt} &= \frac{d_{m}}{d_{0} + I^2} - (k_{T}N_{ext} + \gamma)D \\[5pt]
+  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I \\[5pt]
+\end{aligned}
+$$
+
+Solving for the nullclines gives us the following:
+
+$$
+\begin{aligned}
+  N &= \frac{1}{\gamma + k_{T}D_{ext}}\left(\frac{n_{m}I^2}{n_{0}^2 + I^2}\right) \\[5pt]
+  D &= \frac{1}{\gamma + k_{T}N_{ext}}\left( \frac{d_{m}}{d_{0}^2 + I^2} \right) \\[5pt]
+  I &= \frac{k_{T}ND_{ext}}{\gamma_{I}}
+\end{aligned}
+$$
+
+Note that each nullcline defines a surface in $\mathbb{R}^3$.
+To express different cell fates, we need at least two stable steady states.
+We know that $I^{*} = k_{T}ND_{ext} / \gamma_{I}$ is the unique steady state value of $I$.
+Furthermore, by substituting $I^{*}$ into the $D$-nullcline, we get the unique steady-state value of $D$.
+There also exists a trivial stable steady state at $N = I = 0$.
+
+\medskip
+
+It remains to find the other steady-state values of $N$ as a function of $D_{ext}$.
+To do this, we use the computer algebra system sympy.
+Substituting the equation for $I^{*}$ into the $N$-nullcline gives us the following nonzero steady-state values of $N$:
+
+\medskip
+
+% $$
+% N^{*} = \frac{n_{m}}{2(D_{ext}k_{T} + \gamma)} \pm \sqrt{-\frac{(2D_{ext}\gamma_{I}k_{T}n_{0} - D_{ext}k_{T}n_{m} + 2\gamma \gamma_{I}n_{0})(2 D_{ext} \gamma_{I}k_{T}n_{0} + D_{ext}k_{T}n_{m} + 2\gamma\gamma_{I}n_{0})}{2D_{ext}k_{T}(D_{ext}k_{T} + \gamma)}}
+% $$
+
+$$
+N^{*} = \frac{n_{m}}{2(D_{ext}k_{T} + \gamma)} \pm \sqrt{\frac{(D_{ext}k_{T}n_{m})^2 - 4(D_{ext}\gamma_{I}k_{T}n_{0} + \gamma \gamma_{I}n_{0})^2}{2D_{ext}k_{T}(D_{ext}k_{T} + \gamma)}}
+$$
+
+\textbf{Remark}: As mentioned above, we need two non-zero stable steady states for $N$ in order to represent the two different cell fates. This requires that the discriminant is positive:
+
+$$D_{ext}k_{T}n_{m} > 2D_{ext}\gamma_{I}k_{T}n_{0} + 2\gamma \gamma_{I} n_{0} \Rightarrow D_{ext}k_{T}(n_{m} - 2\gamma_{I}n_{0}) - 2\gamma \gamma_{I} n_{0} > 0$$
+
+As a general rule of thumb, increasing $n_{m}$ is a reliable way to recover these steady states.
+If we replace the $>$ sign with $=$ in the equation above, we find the location of a fold bifurcation.
+The lower branch corresponds to the $N = N_{ext}$, $D = D_{ext}$ unstable steady state.
+If the system is in this state, any perturbation will cause one cell to collapse to the high notch steady state and the other to collapse to the $N = 0$ steady state.
+Shown below is a rough sketch of the bifurcation diagram, including the corresponding cell fates.
+
+\includegraphics[width=\textwidth]{bifurcation-diagram}
+
 \end{flushleft}

### 2025-03-09 code/new-ode-model/NewModel_HexGrid_Visualization.py
Hex grid visualization for new model - kinda works!

@@ -1,0 +1,262 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.colors import LinearSegmentedColormap
+from matplotlib.patches import RegularPolygon
+from scipy.integrate import solve_ivp
+import matplotlib.animation as animation
+
+# --- Updated Parameter Values ---
+k = 2           # Hill exponent for Notch activation
+NM = 10         # Maximum rates of Notch production (increased from 10)
+DM = 10         # Maximum rates of Delta production
+N0 = 100        # Delta Hill function
+D0 = 100        # Delta Hill function
+KT = 0.0001     # Binding rate between estracellular Delta and Notch
+G = 0.019        # Decay rate for Delta and Notch (increased from 0.01)
+GI = 0.025      # Decay rate for NICD
+
+# Hexagonal grid size parameters
+nx = 8  # Number of cells in x direction
+ny = 8  # Number of cells in y direction
+
+# Animation settings
+save_animation = False
+
+# --- Hill Functions ---
+def H_plus(I):
+    """Notch activation by intracellular NICD."""
+    return (NM * I**k) / ((N0**k) + (I**k))
+
+def H_minus(I):
+    """Delta inhibited by intracellular NICD."""
+    return (DM * D0**k) / ((D0**k) + (I**k))
+
+# --- Hexagonal Grid Helper Functions ---
+def get_hex_neighbors(i, j, nx, ny):
+    """
+    Get the indices of the six neighboring hexagons with periodic boundary conditions.
+    In a hexagonal grid, each cell has 6 neighbors.
+    """
+    # The neighbor pattern depends on whether we're in an even or odd row
+    is_even_row = (i % 2 == 0)
+    
+    if is_even_row:
+        # Even rows: neighbors are at positions:
+        # (i-1,j-1), (i-1,j), (i,j-1), (i,j+1), (i+1,j-1), (i+1,j)
+        neighbors = [
+            ((i-1) % ny, (j-1) % nx),  # Top-left
+            ((i-1) % ny, j),           # Top-right
+            (i, (j-1) % nx),           # Left
+            (i, (j+1) % nx),           # Right
+            ((i+1) % ny, (j-1) % nx),  # Bottom-left
+            ((i+1) % ny, j)            # Bottom-right
+        ]
+    else:
+        # Odd rows: neighbors are at positions:
+        # (i-1,j), (i-1,j+1), (i,j-1), (i,j+1), (i+1,j), (i+1,j+1)
[…]

### 2025-03-12 writing/notes.tex
Add more features to gillespie simulations. Begin writing final code in /models

@@
   \item Basal extracellular Notch ($N$) production has the form $H^{+}(I) = n_{m}I^2/(n_{0}^2 + I^2)$. 
-  \item Basal extracellular Delta ($D$) production has the form $H^{-}(I) = d_{m}/(d_{0}^2 + I^2)$.
+  \item Basal extracellular Delta ($D$) production has the form $H^{-}(I) = d_{m}d_{0}^2/(d_{0}^2 + I^2)$.
   \item Extracellular Notch and Delta have the same decay rate, $\gamma$.
@@
 
+\subsubsection{Time-Scale Analysis}
+
+The expected time for event $i$ to occur is given by $\mathbb{E}[X_{i}] = 1/\lambda_{i}$.
+Therefore, the rate at which event $i$ occurs is $\lambda_{i}$, which is equal to the rate at which the event occurs in the original ODE.
+This means that the trajectories generated from the Gillespie's algorithm implementation of the model can be compared directly to the trajectories generated from the ODE itself, without any need for scaling.
+
 \subsection{Theoretical Justification}
@@
   \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\[5pt]
-  \frac{dD}{dt} &= \frac{d_{m}}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\[5pt]
+  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\[5pt]
   \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
@@
   \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - (k_{T}D_{ext} + \gamma)N \\[5pt]
-  \frac{dD}{dt} &= \frac{d_{m}}{d_{0} + I^2} - (k_{T}N_{ext} + \gamma)D \\[5pt]
+  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - (k_{T}N_{ext} + \gamma)D \\[5pt]
   \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I \\[5pt]
@@
   N &= \frac{1}{\gamma + k_{T}D_{ext}}\left(\frac{n_{m}I^2}{n_{0}^2 + I^2}\right) \\[5pt]
-  D &= \frac{1}{\gamma + k_{T}N_{ext}}\left( \frac{d_{m}}{d_{0}^2 + I^2} \right) \\[5pt]
+  D &= \frac{1}{\gamma + k_{T}N_{ext}}\left( \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} \right) \\[5pt]
   I &= \frac{k_{T}ND_{ext}}{\gamma_{I}}
@@
 \medskip
-
-% $$
-% N^{*} = \frac{n_{m}}{2(D_{ext}k_{T} + \gamma)} \pm \sqrt{-\frac{(2D_{ext}\gamma_{I}k_{T}n_{0} - D_{ext}k_{T}n_{m} + 2\gamma \gamma_{I}n_{0})(2 D_{ext} \gamma_{I}k_{T}n_{0} + D_{ext}k_{T}n_{m} + 2\gamma\gamma_{I}n_{0})}{2D_{ext}k_{T}(D_{ext}k_{T} + \gamma)}}
-% $$
 

### 2025-03-12 code/gillespie-model/bifurcation.py

@@ -2,7 +2,7 @@ from sympy import solve, symbols
 
 g, gI, kT, n, nm, n0, de = symbols('g gI kT n nm n0 de')
 iStar = (kT * n * de) / gI
-sol = solve((1 / (g + (kT * de))) * (nm * (iStar**2) / ((n0**2) + (iStar**2))) - n, n)
+sol = solve((1 / (g + (kT * de))) * ((nm * (iStar**2)) / ((n0**2) + (iStar**2))) - n, n)
 
 print(sol)
 

### 2025-03-12 code/gillespie-model/domains.py

@@ -1,0 +1,123 @@
+import numpy as np
+
+##################
+# LINEAR DOMAINS #
+##################
+
+def linear(size, boundary):
+
+    # Set the number of cell pairs
+    pairs = size
+    if boundary == "dirichlet":
+        pairs = size - 1
+
+    # Generate the cell pairs
+    output = [] 
+    for i in range(pairs):
+        output.append([i, (i+1) % size])
+        output.append([(i+1) % size, i])
+
+    return np.array(output), size
+
+# TEST CASES
+# print(linear(2, "dirichlet"))
+# print(linear(10, "dirichlet"))
+# print(linear(3, "periodic"))
+# print(linear(10, "periodic"))
+
+
+#####################
+# HEXAGONAL DOMAINS #
+#####################
+
+# Hexagonal domains paralellograms and are indexed from left to right, top to bottom 
+
+# EXAMPLE:
+#
+# 00 01 02 03 04 
+#  05 06 07 08 09
+#   10 11 12 13 14
+#    15 16 17 18 19
+#     20 21 22 23 24
+
+def hexagonal_dirichlet(height, width):
+
+    # Generate the parallelogram row by row
+    output = []
+    for h in range(height):
+
+        # Add all neighbours of each hexagon
+        for w in range(width):
+
+            # Get the index of the current cell
+            i = (h * width) + w
+
+            # Top left neighbour
+            if h > 0: output.append([i, i - width])
+
+            # Top right neighbour
+            if h > 0 and w < width - 1: output.append([i, i - width + 1])
+
+            # Left neighbour
+            if w > 0: output.append([i, i - 1])
+
+            # Right neighbour
+            if w < width - 1: output.append([i, i + 1])
+
+            # Bottom left neighbour
+            if h < height - 1 and w > 0: output.append([i, i + width - 1])
+
+            # Bottom right neighbour
+            if h < height - 1: output.append([i, i + width])
+
+    return np.array(output), (height * width)
+
+
+def hexagonal_periodic(height, width):
+
+    # Generate the parallelogram row by row
+    output = []
+    for h in range(height):
+
+        # Add all neighbours of each hexagon
+        for w in range(width):
+
+            # Get the row of this cell, and the row above/below
+            rt = ((h - 1) % height) * width
+            rm = (h * width)
+            rb = ((h + 1) % height) * width
+
+            # Top left neighbour
[…]

### 2025-03-12 code/gillespie-model/visualization.py

@@ -1,0 +1,45 @@
+from matplotlib.colors import LinearSegmentedColormap
+import matplotlib.pyplot as plt
+import numpy as np
+
+# Create a static visualization of Delta levels over time
+def linear(v_time, v_state):
+    cells = v_state.shape[1]
+
+    # Create a custom colormap for Delta levels (blue to red)
+    delta_cmap = LinearSegmentedColormap.from_list("DeltaMap", ["blue", "white", "red"])
+
+    # --- Static visualization of Delta patterns over time ---
+    plt.figure(figsize=(12, 8))
+    plt.subplot(3, 1, 1)
+
+    # Subplot 1: Line plot of Delta levels for each cell
+    for i in range(cells):
+        plt.plot(v_time, v_state[:, i, 1], label=f'Cell {i+1}')
+
+    plt.xlabel('Time')
+    plt.ylabel('Delta Level')
+    plt.title('Delta Expression Over Time for Each Cell')
+    plt.legend(loc='upper right', ncol=2)
+
+    # Subplot 3: Cell row visualization at the final time point
+    ax = plt.subplot(3, 1, 3)
+    d_final = v_state[-1, :, 1]
+    d_normal = (d_final - np.min(d_final)) / (np.max(d_final) - np.min(d_final)) 
+
+    # Draw cells as rectangles colored by Delta level
+    for i in range(cells):
+        color = delta_cmap(d_normal[i])
+        rect = plt.Rectangle((i, 0), 1, 1, facecolor=color, edgecolor='black')
+        plt.text(i + 0.5, 0.5, f"{d_final[i]:.1f}", ha='center', va='center')
+        ax.add_patch(rect)
+
+    plt.xlim(0, cells)
+    plt.ylim(0, 1)
+    plt.title('Final Delta Expression Pattern')
+    plt.xticks(np.arange(0.5, cells, 1), [f'Cell {i+1}' for i in range(cells)])
+    plt.yticks([])
+
+    # Display the static plots
+    plt.tight_layout()
+    plt.show()  

### 2025-03-12 models/NewModel_CellLine_Visualization.py

@@ -1,0 +1,186 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.colors import LinearSegmentedColormap
+from scipy.integrate import solve_ivp
+import matplotlib.animation as animation
+
+# --- Initialize Parameter Values ---
+k = 2           # Hill exponent for Notch activation
+NM = 10         # Maximum rates of Notch production
+DM = 10         # Maximum rates of Delta production
+N0 = 100        # Delta Hill function
+D0 = 100        # Delta Hill function
+KT = 0.0001     # Binding rate between estracellular Delta and Notch
+G = 0.02        # Decay rate for Delta and Notch
+GI = 0.025      # Decay rate for NICD
+
+# Number of cells in the row
+num_cells = 8
+
+# --- Hill Functions ---
+def H_plus(I):
+    """Notch activation by intracellular NICD."""
+    return (NM * I**k) / ((N0**k) + (I**k))
+
+def H_minus(I):
+    """Delta inhibited by intracellular NICD."""
+    return (DM * D0**k) / ((D0**k) + (I**k))
+
+# --- ODE System ---
+def ode_system(t, y):
+    # Reshape into a matrix where each row represents a cell [N, D, I]
+    state = y.reshape(num_cells, 3)
+    
+    # Initialize the derivatives
+    derivatives = np.zeros_like(state)
+    
+    for i in range(num_cells):
+        # Get left and right neighbor indices with periodic boundary conditions
+        left_idx = (i - 1) % num_cells
+        right_idx = (i + 1) % num_cells
+        
+        # Current cell values
+        N_i, D_i, I_i = state[i]
+        
+        # Neighbor Delta values (for Notch-Delta binding)
+        D_left = state[left_idx][1]
+        D_right = state[right_idx][1]
+        
+        # Average Delta signal from neighbors
+        D_neighbors = (D_left + D_right) / 2
+        
+        # Calculate derivatives for current cell
+        dN_dt = H_plus(I_i) - KT * N_i * D_neighbors - G * N_i
+        dD_dt = H_minus(I_i) - G * D_i
+        dI_dt = KT * N_i * D_neighbors - GI * I_i
+        
+        derivatives[i] = [dN_dt, dD_dt, dI_dt]
+    
[…]

### 2025-03-12 models/NewModel_TwoCell_Deterministic.py

@@ -1,0 +1,117 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import solve_ivp
+
+# --- Initialize Parameter Values ---
+k = 2           # Hill exponent for Notch activation
+NM = 15         # Maximum rates of Notch production
+DM = 10         # Maximum rates of Delta production
+N0 = 100        # Delta Hill function
+D0 = 100        # Delta Hill function
+KT = 0.0001     # Binding rate between estracellular Delta and Notch
+G = 0.02        # Decay rate for Delta and Notch
+GI = 0.025       # Decay rate for NICD
+
+# --- Hill Functions ---
+def H_plus(I):
+    """Notch activation by intracellular NICD."""
+    return (NM * I**k) / ((N0**k) + (I**k))
+
+def H_minus(I):
+    """Delta inhibited by intracellular NICD."""
+    return (DM * D0**k) / ((D0**k) + (I**k))
+
+# --- ODE System ---
+def ode_system(t, y):
+    N1, D1, I1, N2, D2, I2 = y
+    
+    # Cell 1
+    dN1_dt = H_plus(I1) - KT * N1 * D2 - G * N1
+    dD1_dt = H_minus(I1) - KT * D1 * N2 - G * D1
+    dI1_dt = KT * N1 * D2 - GI * I1
+
+    # Cell 2
+    dN2_dt = H_plus(I2) - KT * N2 * D1 - G * N2
+    dD2_dt = H_minus(I2) - KT * D2 * N1 - G * D2
+    dI2_dt = KT * N2 * D1 - GI * I2
+
+    return [dN1_dt, dD1_dt, dI1_dt, dN2_dt, dD2_dt, dI2_dt]
+
+# --- Simulation Settings ---
+t_span = (0, 1000)  
+t_eval = np.linspace(t_span[0], t_span[1], 1000)
+
+# Initial conditions 
+# C = [N, D, I]
+C1 = [280, 200, 100]  # Initial conditions for cell 1
+C2 = [300, 180, 120]  # Initial conditions for cell 2
+
+# y = [N1, D1, I1, N2, D2, I2]
[…]

### 2025-03-12 models/config.py

@@ -1,0 +1,30 @@
+import numpy as np
+import numba
+
+# Parameter values
+K = 2        # Hill function exponent
+NM = 10      # Maximum rate of Notch production
+DM = 10      # Maximum rate of Delta production
+N0 = 100     # Notch hill function (H+) half-max
+D0 = 100     # Delta hill function (H-) half-max
+KT = 0.0001  # Delta-Notch binding rate
+G = 0.02     # Extracellular molecule decay rate (used in dN/dt and dD/dt equations)
+GI = 0.025   # NICD decay rate
+
+# Set the simulation length and initial cell state
+TIME = 1000
+INITIAL_CELL = np.array([200, 200, 100])
+STEPS = np.linspace(0, TIME, 10000)
+
+# Define increasing hill function based on parameters above
+@numba.njit
+def H_PLUS(i):
+    return (NM * i**K) / ((N0**K) + (i**K))
+
+
+# Define decreasing hill function based on parameters below
+@numba.njit
+def H_MINUS(i):
+    return (DM * D0**K) / ((D0**K) + (i**K))
+
+

### 2025-03-12 models/domains.py

@@ -1,0 +1,123 @@
+import numpy as np
+
+
+# Linear domain
+def linear(size, boundary):
+
+    # Set the number of cell pairs
+    pairs = size
+    if boundary == "dirichlet":
+        pairs = size - 1
+
+    # Generate the cell pairs
+    output = [] 
+    for i in range(pairs):
+        output.append([i, (i+1) % size])
+        output.append([(i+1) % size, i])
+
+    return np.array(output), size
+
+
+# TEST CASES
+# print(linear(2, "dirichlet"))
+# print(linear(10, "dirichlet"))
+# print(linear(3, "periodic"))
+# print(linear(10, "periodic"))
+
+
+
+# Hexagonal domains are paralellograms, indexed from left to right, top to bottom 
+
+# EXAMPLE:
+#
+# 00 01 02 03 04 
+#  05 06 07 08 09
+#   10 11 12 13 14
+#    15 16 17 18 19
+#     20 21 22 23 24
+
+# Hexagonal domain with zero-flux boundary
+def hexagonal_dirichlet(height, width):
+
+    # Generate the parallelogram row by row
+    output = []
+    for h in range(height):
+
+        # Add all neighbours of each hexagon
+        for w in range(width):
+
+            # Get the index of the current cell
+            i = (h * width) + w
+
+            # Top left neighbour
+            if h > 0: output.append([i, i - width])
+
+            # Top right neighbour
+            if h > 0 and w < width - 1: output.append([i, i - width + 1])
+
+            # Left neighbour
+            if w > 0: output.append([i, i - 1])
+
+            # Right neighbour
+            if w < width - 1: output.append([i, i + 1])
+
+            # Bottom left neighbour
+            if h < height - 1 and w > 0: output.append([i, i + width - 1])
+
+            # Bottom right neighbour
+            if h < height - 1: output.append([i, i + width])
+
+    return np.array(output), (height * width)
+
+
+# Hexagonal domain with periodic boundary
+def hexagonal_periodic(height, width):
+
+    # Generate the parallelogram row by row
+    output = []
+    for h in range(height):
+
+        # Add all neighbours of each hexagon
+        for w in range(width):
+
+            # Get the row of this cell, and the row above/below
+            rt = ((h - 1) % height) * width
+            rm = (h * width)
+            rb = ((h + 1) % height) * width
+
+            # Top left neighbour
+            output.append([rm + w, rt + w])
+
[…]

### 2025-03-12 models/gillespie.py

@@ -1,49 +1,25 @@
+# Import libraries
 import numpy as np
 import sys
 import logging
 import time
 import numba
-import domains
-import visualization
+from scipy.interpolate import interp1d
+
+# Import configuration from config.py
+from config import H_PLUS, H_MINUS, INITIAL_CELL, KT, G, GI, TIME, STEPS
 
 # Initialize logger
 logger = logging.getLogger(__name__)
 logging.basicConfig(filename = "main.log", filemode="w", level=logging.INFO)
 
-# Parameter values
-K = 2        # Hill function exponent
-NM = 10      # Maximum rate of Notch production
-DM = 10      # Maximum rate of Delta production
-N0 = 100     # Notch hill function (H+) half-max
-D0 = 100     # Delta hill function (H-) half-max
-KT = 0.0001  # Delta-Notch binding rate
-G = 0.02     # Extracellular molecule decay rate (used in dN/dt and dD/dt equations)
-GI = 0.025   # NICD decay rate
-
-# Set the initial conditions and domain
-STEPS = 100000
-DOMAIN, SIZE = domains.linear(9, "dirichlet")
-INITIAL_CELL = np.array([200, 200, 100])
-
-
-# Increasing hill function
-@numba.njit
-def H_PLUS(i):
-    return (NM * i**K) / ((N0**K) + (i**K))
-
-
-# Decreasing hill function
-@numba.njit
-def H_MINUS(i):
-    return (DM * D0**K) / ((D0**K) + (i**K))
-
 
 # Generates the reaction rates based on the state
 @numba.njit
-def update_reaction_rates(state):
+def update_reaction_rates(state, domain):
 
     # Initialize empty rate arrays for each reaction
-    r1 = np.empty(len(DOMAIN))
+    r1 = np.empty(len(domain))
     r2 = np.empty(len(state))
     r3 = np.empty(len(state))
     r4 = np.empty(len(state))
@@ -63,7 +39,7 @@ def update_reaction_rates(state):
         r6[i] = H_MINUS(cell[2])   # Rate parameter of Delta production
 
     # Sum over the neighbour pairs to get the binding rates (reaction 1) 
-    for i, (cellA, cellB) in enumerate(DOMAIN):
+    for i, (cellA, cellB) in enumerate(domain):
         r1[i] = KT * state[cellA][0] * state[cellB][1]
 
     # Return the combined array of rates
@@ -73,35 +49,35 @@ def update_reaction_rates(state):
[…]

### 2025-03-12 models/main.py

@@ -1,0 +1,37 @@
+# Import libraries
+import numpy as np
+import sys
+import logging
+import time
+import numba
+
+# Import modules 
+import visualization
+import domains
+import gillespie 
+
+# Initialize logger
+logger = logging.getLogger(__name__)
+logging.basicConfig(filename = "main.log", filemode="w", level=logging.INFO)
+
+
+# Run all simulations
+def main():
+
+    # Two Cell Domain
+    domain, size = domains.linear(2, "dirichlet")
+    gillespie_data = gillespie.run(domain, size, samples = 20)
+    visualization.two_cell_individual(gillespie_data, "Two-Cell Gillespie")
+    
+
+    # Linear Domain (Dirichlet)
+    # domain, size = domains.linear(11, "dirichlet")
+    # v_time, v_state = gillespie.run(domain, size)
+
+    # Linear Domain (Periodic)
+    # domain, size = domains.linear(11, "periodic")
+    # v_time, v_state = gillespie.run(domain, size)
+
+
+main()
+

### 2025-03-12 models/visualization.py

@@ -1,0 +1,78 @@
+# Import libraries
+from matplotlib.colors import LinearSegmentedColormap
+import matplotlib.pyplot as plt
+import numpy as np
+
+# Import modules
+from config import STEPS
+
+
+# Static, individual visualization for two-cell models
+def two_cell_individual(data, title):
+    times, states, means = data
+    fig, axs = plt.subplots(nrows = 3, sharex = True)
+
+    # Iterate through axes (0: Notch, 1: Delta, 2: NICD)
+    for i in range(3):
+
+        # Plot the trajectories for each sample
+        for j in range(len(states)):
+            axs[i].plot(times[j], states[j][:, 0, i], color = "green", alpha = 0.1)
+            axs[i].plot(times[j], states[j][:, 1, i], color = "orange", alpha = 0.1)
+
+        # Plot the means at full opacity
+        axs[i].plot(STEPS, means[:, 0, i], color = "green")
+        axs[i].plot(STEPS, means[:, 1, i], color = "orange")
+
+    # Set axis labels
+    axs[0].set_ylabel("Notch Molecules")
+    axs[1].set_ylabel("Delta Molecules")
+    axs[2].set_ylabel("NICD Molecules")
+    axs[2].set_xlabel("Time")
+
+    # Add title and legend
+    fig.suptitle(title)
+    plt.show()
+
+
+# Static, individual visualization for linear models
+def linear_individual(v_time, v_state):
+    cells = v_state.shape[1]
+
+    # Create a custom colormap for Delta levels (blue to red)
+    delta_cmap = LinearSegmentedColormap.from_list("DeltaMap", ["blue", "white", "red"])
+
+    # --- Static visualization of Delta patterns over time ---
+    plt.figure(figsize=(12, 8))
+    plt.subplot(3, 1, 1)
+
+    # Subplot 1: Line plot of Delta levels for each cell
+    for i in range(cells):
+        plt.plot(v_time, v_state[:, i, 1], label=f'Cell {i+1}')
+
+    plt.xlabel('Time')
+    plt.ylabel('Delta Level')
+    plt.title('Delta Expression Over Time for Each Cell')
+    plt.legend(loc='upper right', ncol=2)
+
+    # Subplot 3: Cell row visualization at the final time point
+    ax = plt.subplot(3, 1, 3)
+    d_final = v_state[-1, :, 1]
[…]

### 2025-03-13 models/config.py
Some progress on /models

@@ -11,10 +11,14 @@ KT = 0.0001  # Delta-Notch binding rate
 G = 0.02     # Extracellular molecule decay rate (used in dN/dt and dD/dt equations)
 GI = 0.025   # NICD decay rate
 
+# Initial and parameter perturbations
+PI = 0.1
+PP = 0.05
+
 # Set the simulation length and initial cell state
 TIME = 1000
+STEPS = 10000
 INITIAL_CELL = np.array([200, 200, 100])
-STEPS = np.linspace(0, TIME, 10000)
 
 # Define increasing hill function based on parameters above
 @numba.njit

### 2025-03-13 models/domains.py

@@ -15,7 +15,8 @@ def linear(size, boundary):
         output.append([i, (i+1) % size])
         output.append([(i+1) % size, i])
 
-    return np.array(output), size
+    name = "Two Cell" if size == 2 else "Linear"
+    return name, np.array(output), size
 
 
 # TEST CASES
@@ -67,7 +68,7 @@ def hexagonal_dirichlet(height, width):
             # Bottom right neighbour
             if h < height - 1: output.append([i, i + width])
 
-    return np.array(output), (height * width)
+    return "Hexagonal", np.array(output), (height * width)
 
 
 # Hexagonal domain with periodic boundary
@@ -103,7 +104,7 @@ def hexagonal_periodic(height, width):
             # Bottom right neighbour
             output.append([rm + w, rb + w])
 
-    return np.array(output), (height * width)
+    return "Hexagonal", np.array(output), (height * width)
 
 
 # Master function to call the correct helper

### 2025-03-13 models/gillespie.py

@@ -16,10 +16,10 @@ logging.basicConfig(filename = "main.log", filemode="w", level=logging.INFO)
 
 # Generates the reaction rates based on the state
 @numba.njit
-def update_reaction_rates(state, domain):
+def update_reaction_rates(state, neighbours):
 
     # Initialize empty rate arrays for each reaction
-    r1 = np.empty(len(domain))
+    r1 = np.empty(len(neighbours))
     r2 = np.empty(len(state))
     r3 = np.empty(len(state))
     r4 = np.empty(len(state))
@@ -39,7 +39,7 @@ def update_reaction_rates(state, domain):
         r6[i] = H_MINUS(cell[2])   # Rate parameter of Delta production
 
     # Sum over the neighbour pairs to get the binding rates (reaction 1) 
-    for i, (cellA, cellB) in enumerate(domain):
+    for i, (cellA, cellB) in enumerate(neighbours):
         r1[i] = KT * state[cellA][0] * state[cellB][1]
 
     # Return the combined array of rates
@@ -54,30 +54,30 @@ def generate_reaction_time(rates):
 
 # Generate an event, returning a (reaction, index) tuple
 @numba.njit
-def generate_event(rates, state, domain):
+def generate_event(rates, state, neighbours):
 
     # Sample a random value in [0, 1] and find its index in the partition
     partition = np.cumsum(rates) / np.sum(rates)
     reaction = np.searchsorted(partition, np.random.rand())
 
     # Check if the event is a binding event
-    if reaction < len(domain):
+    if reaction < len(neighbours):
         return [1, reaction]
 
     # Otherwise do some modulo fuckery to get the reaction
-    reaction_type = 2 + ((reaction - len(domain)) // len(state))
-    index = (reaction - len(domain)) % len(state)
+    reaction_type = 2 + ((reaction - len(neighbours)) // len(state))
+    index = (reaction - len(neighbours)) % len(state)
     return [reaction_type, index]
 
 
 # Updates the state AND the reaciton rates given an event
 @numba.njit
-def update_state(state, event, domain):
+def update_state(state, event, neighbours):
     reaction, index = event
 
     # Binding reaction event
     if reaction == 1:
-        cellA, cellB = domain[index]
+        cellA, cellB = neighbours[index]
         state[cellA][0] -= 1 # Remove Notch from cell A
         state[cellB][1] -= 1 # Remove Delta from cell B
         state[cellA][2] += 1 # Add a NICD to cell A
[…]

### 2025-03-13 models/main.py

@@ -9,6 +9,7 @@ import numba
 import visualization
 import domains
 import gillespie 
+import ode
 
 # Initialize logger
 logger = logging.getLogger(__name__)
@@ -16,14 +17,28 @@ logging.basicConfig(filename = "main.log", filemode="w", level=logging.INFO)
 
 
 # Run all simulations
+np.random.seed(69420)
 def main():
 
-    # Two Cell Domain
-    domain, size = domains.linear(2, "dirichlet")
-    gillespie_data = gillespie.run(domain, size, samples = 20)
-    visualization.two_cell_individual(gillespie_data, "Two-Cell Gillespie")
-    
+    # Two Cell Domain 
+    domain = domains.linear(2, "dirichlet")
 
+    # Deterministic Model
+    deterministic_data = ode.ode(domain, "Deterministic", samples = 25)
+    visualization.two_cell_individual(domain, deterministic_data, "Deterministic")
+
+    # Stochastic Model
+    stochastic_data = ode.ode(domain, "Deterministic", samples = 25)
+    visualization.two_cell_individual(domain, stochastic_data, "Stochastic")
+
+    # Gillespie Model
+    gillespie_data = gillespie.gillespie(domain, samples = 25)
+    visualization.two_cell_individual(domain, gillespie_data, "Gillespie")
+
+    # Comparison Plot
+    all_data = (deterministic_data, stochastic_data, gillespie_data)
+    visualization.two_cell_comparison(all_data)
+    
     # Linear Domain (Dirichlet)
     # domain, size = domains.linear(11, "dirichlet")
     # v_time, v_state = gillespie.run(domain, size)

### 2025-03-13 models/ode.py

@@ -1,0 +1,125 @@
+# Import libraries
+import numpy as np
+import sys
+import logging
+import time
+import numba
+from scipy.integrate import solve_ivp
+
+# Import configuration from config.py
+from config import H_PLUS, H_MINUS, INITIAL_CELL, KT, G, GI, PI, PP, TIME, STEPS
+
+# Initialize logger
+logger = logging.getLogger(__name__)
+logging.basicConfig(filename = "main.log", filemode="w", level=logging.INFO)
+
+
+# Perturb a variable with a sd of n * p
+@numba.njit
+def perturb(n, p):
+    return n + np.random.normal(0, n * p)
+
+
+# Define the ODE system
+def ode_system(t, y, name, neighbours, size, mode):
+
+    # Reshape into a matrix where each row represents a cell [N, D, I]
+    state = y.reshape(size, 3)
+    
+    # Generate the derivative functions
+    derivatives = np.empty((size, 3))
+    for cell in range(size):
+
+        # Get the neighobur indices
+        mask = (neighbours[:, 0] == cell)
+        indices = neighbours[mask, 1]
+
+        # Compute the average delta signal over the neighbours
+        divisor = 1
+        if name == "Linear": divisor = 2
+        if name == "Hexagonal": divisor = 6
+        n_ext = np.sum(state[indices, 0]) / divisor
+        d_ext = np.sum(state[indices, 1]) / divisor
+
+        # Current cell values
+        n, d, i = state[cell]
+       
+        # Use fixed equations in deterministic simulations
+        dN = H_PLUS(i) - KT * n * d_ext - G * n
+        dD = H_MINUS(i) - KT * d * n_ext - G * d
+        dI = KT * n * d_ext - GI * i
+
+        # Perturb parameter values in stochastic simulations
+        if mode == "Stochastic":
+            dN = H_PLUS(i) - perturb(KT, PP) * n * d_ext - perturb(G, PP) * n
+            dD = H_MINUS(i) - perturb(KT, PP) * d * n_ext - perturb(G, PP) * d
+            dI = perturb(KT, PP) * n * d_ext - perturb(GI, PP) * i
+
+        derivatives[cell] = [dN, dD, dI]
+    
+    # Return flattened array
+    return derivatives.flatten()
+
+
+# Run a single simulation of the deterministic ODE model
+def simulate(domain, mode):
+
+    # Set the initial conditions
+    name, neighbours, size = domain
+    state = np.empty((size, 3))
+
+    for i in range(size):
+        for j in range(3):
+
[…]

### 2025-03-13 models/visualization.py

@@ -4,11 +4,13 @@ import matplotlib.pyplot as plt
 import numpy as np
 
 # Import modules
-from config import STEPS
+from config import TIME, STEPS
 
 
 # Static, individual visualization for two-cell models
-def two_cell_individual(data, title):
+def two_cell_individual(domain, data, model):
+
+    name, neighbours, size = domain
     times, states, means = data
     fig, axs = plt.subplots(nrows = 3, sharex = True)
 
@@ -21,18 +23,58 @@ def two_cell_individual(data, title):
             axs[i].plot(times[j], states[j][:, 1, i], color = "orange", alpha = 0.1)
 
         # Plot the means at full opacity
-        axs[i].plot(STEPS, means[:, 0, i], color = "green")
-        axs[i].plot(STEPS, means[:, 1, i], color = "orange")
+        default_times = np.linspace(0, TIME, STEPS)
+        axs[i].plot(default_times, means[:, 0, i], color = "green")
+        axs[i].plot(default_times, means[:, 1, i], color = "orange")
 
     # Set axis labels
     axs[0].set_ylabel("Notch Molecules")
     axs[1].set_ylabel("Delta Molecules")
     axs[2].set_ylabel("NICD Molecules")
-    axs[2].set_xlabel("Time")
 
-    # Add title and legend
-    fig.suptitle(title)
-    plt.show()
+    # Add title and labels to the graph
+    fig.supxlabel("Time")
+    fig.suptitle(f"{name} {model}")
+
+    # Save to the img/ foler
+    fname = "_".join(f"{name} {model}".lower().split(" "))
+    plt.savefig(f"img/{fname}.pdf", format = "pdf")
+
+
+# Static, comparison visualization for two-cell models
+def two_cell_comparison(all_data):
+
+    # Iterate through axes (0: Deterministic, 1: Stochastic, 2: Gillespie)
+    fig, axs = plt.subplots(nrows = 3, sharex = True)
+    for i in range(3):
+
+        # Unpack the data arrays
+        data = all_data[i]
+        times, states, means = data 
+
+        # Plot the trajectories for each sample
+        for j in range(len(states)):
+            axs[i].plot(times[j], states[j][:, 0, 0], color = "green", alpha = 0.1)
+            axs[i].plot(times[j], states[j][:, 1, 0], color = "orange", alpha = 0.1)
+
+        # Plot the means at full opacity
[…]

### 2025-03-13 code/new-ode-model/NewModel_TwoCell_SDE.py
Stochastic DE model

@@ -1,0 +1,346 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+
+def run_stochastic_delta_notch(noise_levels=None, num_simulations=1, seed=None, dt=0.1, t_max=1000, 
+                               initial_conditions=None):
+    """
+    Run multiple Delta-Notch interaction simulations with stochastic differential equations.
+    
+    Parameters:
+    -----------
+    noise_levels : dict or None
+        Dictionary with keys 'N', 'D', 'I' and values representing the
+        noise intensity (sigma) for each variable. If None, no noise is added.
+    num_simulations : int
+        Number of simulations to run with different noise realizations.
+    seed : int or None
+        Seed for random number generation for reproducibility.
+    dt : float
+        Time step for Euler-Maruyama integration.
+    t_max : float
+        Maximum simulation time.
+    initial_conditions : dict or None
+        Dictionary with keys 'N1', 'D1', 'I1', 'N2', 'D2', 'I2' specifying
+        initial conditions. If None, random initial conditions are generated.
+        
+    Returns:
+    --------
+    all_solutions : list
+        List of dictionaries containing simulation results
+    params : dict
+        Dictionary of parameter values used in the simulations
+    """
+    # Set random seed if provided
+    if seed is not None:
+        np.random.seed(seed)
+    
+    # --- Initialize Default Parameter Values ---
+    params = {
+        'k': 2,          # Hill exponent for Notch activation
+        'NM': 10,        # Maximum rates of Notch production
+        'DM': 10,        # Maximum rates of Delta production
+        'N0': 100,       # Delta Hill function
+        'D0': 100,       # Delta Hill function
+        'KT': 0.0001,    # Binding rate between extracellular Delta and Notch
+        'G': 0.02,       # Decay rate for Delta and Notch
+        'GI': 0.025      # Decay rate for NICD
+    }
+    
+    # Default noise levels if none provided
+    if noise_levels is None:
+        noise_levels = {'N': 0.0, 'D': 0.0, 'I': 0.0}
+    
+    # Ensure all noise components are defined
+    for key in ['N', 'D', 'I']:
+        if key not in noise_levels:
[…]

### 2025-03-13 models/config.py
Streamline SDE code.

@@ -12,8 +12,8 @@ G = 0.02     # Extracellular molecule decay rate (used in dN/dt and dD/dt equati
 GI = 0.025   # NICD decay rate
 
 # Initial and parameter perturbations
-PI = 0.1
-PP = 0.05
+PI = 0.05
+NOISE = 0.02
 
 # Set the simulation length and initial cell state
 TIME = 1000

### 2025-03-13 models/main.py

@@ -28,7 +28,7 @@ def main():
     visualization.two_cell_individual(domain, deterministic_data, "Deterministic")
 
     # Stochastic Model
-    stochastic_data = ode.ode(domain, "Deterministic", samples = 25)
+    stochastic_data = ode.ode(domain, "Stochastic", samples = 25)
     visualization.two_cell_individual(domain, stochastic_data, "Stochastic")
 
     # Gillespie Model

### 2025-03-13 models/ode.py

@@ -7,26 +7,19 @@ import numba
 from scipy.integrate import solve_ivp
 
 # Import configuration from config.py
-from config import H_PLUS, H_MINUS, INITIAL_CELL, KT, G, GI, PI, PP, TIME, STEPS
+from config import H_PLUS, H_MINUS, INITIAL_CELL, KT, G, GI, PI, NOISE, TIME, STEPS
 
 # Initialize logger
 logger = logging.getLogger(__name__)
 logging.basicConfig(filename = "main.log", filemode="w", level=logging.INFO)
 
 
-# Perturb a variable with a sd of n * p
-@numba.njit
-def perturb(n, p):
-    return n + np.random.normal(0, n * p)
-
-
 # Define the ODE system
-def ode_system(t, y, name, neighbours, size, mode):
+@numba.njit
+def deterministic_step(domain, state):
 
-    # Reshape into a matrix where each row represents a cell [N, D, I]
-    state = y.reshape(size, 3)
-    
-    # Generate the derivative functions
+    # Generate the increments 
+    name, neighbours, size = domain
     derivatives = np.empty((size, 3))
     for cell in range(size):
 
@@ -35,11 +28,9 @@ def ode_system(t, y, name, neighbours, size, mode):
         indices = neighbours[mask, 1]
 
         # Compute the average delta signal over the neighbours
-        divisor = 1
-        if name == "Linear": divisor = 2
-        if name == "Hexagonal": divisor = 6
-        n_ext = np.sum(state[indices, 0]) / divisor
-        d_ext = np.sum(state[indices, 1]) / divisor
+        divisors = {"Two Cell": 1, "Linear": 2, "Hexagonal": 6}
+        n_ext = np.sum(state[indices, 0]) / divisors[name]
+        d_ext = np.sum(state[indices, 1]) / divisors[name]
 
         # Current cell values
         n, d, i = state[cell]
@@ -48,66 +39,64 @@ def ode_system(t, y, name, neighbours, size, mode):
         dN = H_PLUS(i) - KT * n * d_ext - G * n
         dD = H_MINUS(i) - KT * d * n_ext - G * d
         dI = KT * n * d_ext - GI * i
-
-        # Perturb parameter values in stochastic simulations
-        if mode == "Stochastic":
-            dN = H_PLUS(i) - perturb(KT, PP) * n * d_ext - perturb(G, PP) * n
[…]

### 2025-03-13 models/visualization.py

@@ -37,6 +37,7 @@ def two_cell_individual(domain, data, model):
     fig.suptitle(f"{name} {model}")
 
     # Save to the img/ foler
+    plt.tight_layout()
     fname = "_".join(f"{name} {model}".lower().split(" "))
     plt.savefig(f"img/{fname}.pdf", format = "pdf")
 

### 2025-03-14 paper/paper.tex
Create paper draft with bibliography, generate figures with higher DPI.

--- 2025-03-12 writing/notes.tex
@@
 \usepackage[left=1.5in, right=1.5in, top=1in, bottom=1in]{geometry}
+\usepackage{graphicx}
 \usepackage{amsmath}
@@
 \usepackage{hyperref}
+
 \usepackage{graphicx}
+\graphicspath{ {./img/} }
+
+\usepackage[backend=biber]{biblatex}
+\bibliography{sources.bib}
+
+\title{Mathematical Models of Delta-Notch Signalling}  
+\author{Riley Wheadon, Edwin Huras, Ziyi Zhuang}
 
@@
 
+\maketitle
+
 \begin{flushleft}
 
-\section{Background Information}
+\section{Abstract}
 
-Sources: \href{https://www.pnas.org/doi/full/10.1073/pnas.1416287112}{Boareto et al., 2015}, \href{https://www.nature.com/articles/nrm2009}{Bray, 2006}
+\section{Introduction}
 
@@
 
-Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor.
+Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor \cite{collier_pattern_1996}.
 This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch and \emph{inhibits} Delta.
@@
 \end{itemize}
-
-\section{Ideas and Issues with Collier et al., 1996}
 
@@
 
-\newpage
-
 \subsection{Issues}
@@
 
-\section{Probabilistic Models of Increasing Complexity}
+\section{Methods}
 
-\textbf{Idea}: Our first goal is to develop a probabilistic model based on a modern understanding of the biochemistry of the delta-notch signalling network.
-Then, we will use this model to attempt to replicate qualitative results from Collier et al.
-(1996).
-
-\subsection{A First Model}
+\subsection{Parameters and Assumptions}
 
@@
 
-\subsection{Computational Implementation}
+\subsection{Kolmogorov Forward Equations}
+
+\subsection{Deterministic ODE Model}
+
+\subsection{Stochastic ODE Model}
+
+\subsection{Gillespie Algorithm}
 
@@
 
-\subsubsection{Time-Scale Analysis}
+\medskip
 
-The expected time for event $i$ to occur is given by $\mathbb{E}[X_{i}] = 1/\lambda_{i}$.
+\textbf{Remark}: The expected time for event $i$ to occur is given by $\mathbb{E}[X_{i}] = 1/\lambda_{i}$.
 Therefore, the rate at which event $i$ occurs is $\lambda_{i}$, which is equal to the rate at which the event occurs in the original ODE.
@@
 
-\includegraphics[width=\textwidth]{bifurcation-diagram}
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/bifurcation-diagram.png}
+  \caption{Approximate bifurcation diagram of the equilibrium Notch concentration of a single cell under the influence of fixed extracellular Delta from a neighbour.}
+\end{figure}
+
+\section{Results}
+
+\subsection{Two-Cell Domain}
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/two_cell_deterministic.pdf}
+  \caption{Results of the two-cell deterministic ODE model.}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/two_cell_stochastic.pdf}
+  \caption{Results of the two-cell stochastic ODE model.}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/two_cell_gillespie.pdf}
+  \caption{Results of the two-cell Gillespie model.}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/two_cell_comparison.pdf}
+  \caption{Comparison of Notch concentrations in deterministic ODE, stochastic ODE, and Gillespie models.
+All models follow roughly the same trajectory and have similar differentiation times, although the Gillespie model appears to converge slightly faster.}
+\end{figure}
+
+\subsection{Linear Domain}
+
+\subsection{Hexagonal Domain}
+
+\section{Discussion}
 
@@
 
+\nocite{*}
+\printbibliography
+
 \end{document}

### 2025-03-14 models/NewModel_CellLine_Visualization.py

@@ -1,186 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from matplotlib.colors import LinearSegmentedColormap
-from scipy.integrate import solve_ivp
-import matplotlib.animation as animation
-
-# --- Initialize Parameter Values ---
-k = 2           # Hill exponent for Notch activation
-NM = 10         # Maximum rates of Notch production
-DM = 10         # Maximum rates of Delta production
-N0 = 100        # Delta Hill function
-D0 = 100        # Delta Hill function
-KT = 0.0001     # Binding rate between estracellular Delta and Notch
-G = 0.02        # Decay rate for Delta and Notch
-GI = 0.025      # Decay rate for NICD
-
-# Number of cells in the row
-num_cells = 8
-
-# --- Hill Functions ---
-def H_plus(I):
-    """Notch activation by intracellular NICD."""
-    return (NM * I**k) / ((N0**k) + (I**k))
-
-def H_minus(I):
-    """Delta inhibited by intracellular NICD."""
-    return (DM * D0**k) / ((D0**k) + (I**k))
-
-# --- ODE System ---
-def ode_system(t, y):
-    # Reshape into a matrix where each row represents a cell [N, D, I]
-    state = y.reshape(num_cells, 3)
-    
-    # Initialize the derivatives
-    derivatives = np.zeros_like(state)
-    
-    for i in range(num_cells):
-        # Get left and right neighbor indices with periodic boundary conditions
-        left_idx = (i - 1) % num_cells
-        right_idx = (i + 1) % num_cells
-        
-        # Current cell values
-        N_i, D_i, I_i = state[i]
-        
-        # Neighbor Delta values (for Notch-Delta binding)
-        D_left = state[left_idx][1]
-        D_right = state[right_idx][1]
-        
-        # Average Delta signal from neighbors
-        D_neighbors = (D_left + D_right) / 2
-        
-        # Calculate derivatives for current cell
-        dN_dt = H_plus(I_i) - KT * N_i * D_neighbors - G * N_i
-        dD_dt = H_minus(I_i) - G * D_i
-        dI_dt = KT * N_i * D_neighbors - GI * I_i
-        
-        derivatives[i] = [dN_dt, dD_dt, dI_dt]
-    
[…]

### 2025-03-14 models/NewModel_HexGrid_Visualization.py

@@ -1,262 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from matplotlib.colors import LinearSegmentedColormap
-from matplotlib.patches import RegularPolygon
-from scipy.integrate import solve_ivp
-import matplotlib.animation as animation
-
-# --- Updated Parameter Values ---
-k = 2           # Hill exponent for Notch activation
-NM = 10         # Maximum rates of Notch production (increased from 10)
-DM = 10         # Maximum rates of Delta production
-N0 = 100        # Delta Hill function
-D0 = 100        # Delta Hill function
-KT = 0.0001     # Binding rate between estracellular Delta and Notch
-G = 0.019        # Decay rate for Delta and Notch (increased from 0.01)
-GI = 0.025      # Decay rate for NICD
-
-# Hexagonal grid size parameters
-nx = 8  # Number of cells in x direction
-ny = 8  # Number of cells in y direction
-
-# Animation settings
-save_animation = False
-
-# --- Hill Functions ---
-def H_plus(I):
-    """Notch activation by intracellular NICD."""
-    return (NM * I**k) / ((N0**k) + (I**k))
-
-def H_minus(I):
-    """Delta inhibited by intracellular NICD."""
-    return (DM * D0**k) / ((D0**k) + (I**k))
-
-# --- Hexagonal Grid Helper Functions ---
-def get_hex_neighbors(i, j, nx, ny):
-    """
-    Get the indices of the six neighboring hexagons with periodic boundary conditions.
-    In a hexagonal grid, each cell has 6 neighbors.
-    """
-    # The neighbor pattern depends on whether we're in an even or odd row
-    is_even_row = (i % 2 == 0)
-    
-    if is_even_row:
-        # Even rows: neighbors are at positions:
-        # (i-1,j-1), (i-1,j), (i,j-1), (i,j+1), (i+1,j-1), (i+1,j)
-        neighbors = [
-            ((i-1) % ny, (j-1) % nx),  # Top-left
-            ((i-1) % ny, j),           # Top-right
-            (i, (j-1) % nx),           # Left
-            (i, (j+1) % nx),           # Right
-            ((i+1) % ny, (j-1) % nx),  # Bottom-left
-            ((i+1) % ny, j)            # Bottom-right
-        ]
-    else:
-        # Odd rows: neighbors are at positions:
-        # (i-1,j), (i-1,j+1), (i,j-1), (i,j+1), (i+1,j), (i+1,j+1)
[…]

### 2025-03-14 models/NewModel_TwoCell_Deterministic.py

@@ -1,117 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from scipy.integrate import solve_ivp
-
-# --- Initialize Parameter Values ---
-k = 2           # Hill exponent for Notch activation
-NM = 15         # Maximum rates of Notch production
-DM = 10         # Maximum rates of Delta production
-N0 = 100        # Delta Hill function
-D0 = 100        # Delta Hill function
-KT = 0.0001     # Binding rate between estracellular Delta and Notch
-G = 0.02        # Decay rate for Delta and Notch
-GI = 0.025       # Decay rate for NICD
-
-# --- Hill Functions ---
-def H_plus(I):
-    """Notch activation by intracellular NICD."""
-    return (NM * I**k) / ((N0**k) + (I**k))
-
-def H_minus(I):
-    """Delta inhibited by intracellular NICD."""
-    return (DM * D0**k) / ((D0**k) + (I**k))
-
-# --- ODE System ---
-def ode_system(t, y):
-    N1, D1, I1, N2, D2, I2 = y
-    
-    # Cell 1
-    dN1_dt = H_plus(I1) - KT * N1 * D2 - G * N1
-    dD1_dt = H_minus(I1) - KT * D1 * N2 - G * D1
-    dI1_dt = KT * N1 * D2 - GI * I1
-
-    # Cell 2
-    dN2_dt = H_plus(I2) - KT * N2 * D1 - G * N2
-    dD2_dt = H_minus(I2) - KT * D2 * N1 - G * D2
-    dI2_dt = KT * N2 * D1 - GI * I2
-
-    return [dN1_dt, dD1_dt, dI1_dt, dN2_dt, dD2_dt, dI2_dt]
-
-# --- Simulation Settings ---
-t_span = (0, 1000)  
-t_eval = np.linspace(t_span[0], t_span[1], 1000)
-
-# Initial conditions 
-# C = [N, D, I]
-C1 = [280, 200, 100]  # Initial conditions for cell 1
-C2 = [300, 180, 120]  # Initial conditions for cell 2
-
-# y = [N1, D1, I1, N2, D2, I2]
[…]

### 2025-03-14 models/NewModel_TwoCell_SDE.py

@@ -1,346 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-import matplotlib.cm as cm
-
-def run_stochastic_delta_notch(noise_levels=None, num_simulations=1, seed=None, dt=0.1, t_max=1000, 
-                               initial_conditions=None):
-    """
-    Run multiple Delta-Notch interaction simulations with stochastic differential equations.
-    
-    Parameters:
-    -----------
-    noise_levels : dict or None
-        Dictionary with keys 'N', 'D', 'I' and values representing the
-        noise intensity (sigma) for each variable. If None, no noise is added.
-    num_simulations : int
-        Number of simulations to run with different noise realizations.
-    seed : int or None
-        Seed for random number generation for reproducibility.
-    dt : float
-        Time step for Euler-Maruyama integration.
-    t_max : float
-        Maximum simulation time.
-    initial_conditions : dict or None
-        Dictionary with keys 'N1', 'D1', 'I1', 'N2', 'D2', 'I2' specifying
-        initial conditions. If None, random initial conditions are generated.
-        
-    Returns:
-    --------
-    all_solutions : list
-        List of dictionaries containing simulation results
-    params : dict
-        Dictionary of parameter values used in the simulations
-    """
-    # Set random seed if provided
-    if seed is not None:
-        np.random.seed(seed)
-    
-    # --- Initialize Default Parameter Values ---
-    params = {
-        'k': 2,          # Hill exponent for Notch activation
-        'NM': 10,        # Maximum rates of Notch production
-        'DM': 10,        # Maximum rates of Delta production
-        'N0': 100,       # Delta Hill function
-        'D0': 100,       # Delta Hill function
-        'KT': 0.0001,    # Binding rate between extracellular Delta and Notch
-        'G': 0.02,       # Decay rate for Delta and Notch
-        'GI': 0.025      # Decay rate for NICD
-    }
-    
-    # Default noise levels if none provided
-    if noise_levels is None:
-        noise_levels = {'N': 0.0, 'D': 0.0, 'I': 0.0}
-    
-    # Ensure all noise components are defined
-    for key in ['N', 'D', 'I']:
-        if key not in noise_levels:
[…]

### 2025-03-14 models/NewModel_TwoCell_VaryInitialConditions.py

@@ -1,170 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from scipy.integrate import solve_ivp
-import matplotlib.cm as cm
-
-# --- Initialize Parameter Values ---
-k = 2           # Hill exponent for Notch activation
-NM = 10         # Maximum rates of Notch production
-DM = 10         # Maximum rates of Delta production
-N0 = 100        # Delta Hill function
-D0 = 100        # Delta Hill function
-KT = 0.0001     # Binding rate between extracellular Delta and Notch
-G = 0.01        # Decay rate for Delta and Notch
-GI = 0.025      # Decay rate for NICD
-
-# --- Hill Functions ---
-def H_plus(I):
-    """Notch activation by intracellular NICD."""
-    return (NM * I**k) / ((N0**k) + (I**k))
-
-def H_minus(I):
-    """Delta inhibited by intracellular NICD."""
-    return (DM * D0**k) / ((D0**k) + (I**k))
-
-# --- ODE System ---
-def ode_system(t, y):
-    N1, D1, I1, N2, D2, I2 = y
-    
-    # Cell 1
-    dN1_dt = H_plus(I1) - KT * N1 * D2 - G * N1
-    dD1_dt = H_minus(I1) - KT * D1 * N2 - G * D1
-    dI1_dt = KT * N1 * D2 - GI * I1
-
-    # Cell 2
-    dN2_dt = H_plus(I2) - KT * N2 * D1 - G * N2
-    dD2_dt = H_minus(I2) - KT * D2 * N1 - G * D2
-    dI2_dt = KT * N2 * D1 - GI * I2
-
-    return [dN1_dt, dD1_dt, dI1_dt, dN2_dt, dD2_dt, dI2_dt]
-
-# --- Simulation Settings ---
-t_span = (0, 1000)  
-t_eval = np.linspace(t_span[0], t_span[1], 1000)
-
-# Generate multiple initial conditions
-num_simulations = 20
-np.random.seed(42)  # For reproducibility
-
-# Ranges for initial values
-N_range = (100, 400)
-D_range = (100, 400)
-I_range = (50, 200)
-
-# Store all solutions
-all_solutions = []
-
-# Run simulations with different initial conditions
-for i in range(num_simulations):
-    # Generate random initial conditions
-    N = np.random.uniform(*N_range, 2)
[…]

### 2025-03-14 models/visualization.py

@@ -39,7 +39,7 @@ def two_cell_individual(domain, data, model):
     # Save to the img/ foler
     plt.tight_layout()
     fname = "_".join(f"{name} {model}".lower().split(" "))
-    plt.savefig(f"img/{fname}.pdf", format = "pdf")
+    plt.savefig(f"../paper/img/{fname}.pdf", dpi = 200, format = "pdf")
 
 
 # Static, comparison visualization for two-cell models
@@ -75,7 +75,7 @@ def two_cell_comparison(all_data):
 
     # Save to the img/ foler
     plt.tight_layout()
-    plt.savefig(f"img/two_cell_comparison.pdf", format = "pdf")
+    plt.savefig(f"../paper/img/two_cell_comparison.pdf", dpi = 200, format = "pdf")
 
 
 # Static, individual visualization for linear models

### 2025-03-14 paper/paper.tex
paper.tex

@@
 \subsection{Kolmogorov Forward Equations}
+\subsubsection{Derivation of Kolmogorov Forward Equation}
+
+Let us define:\\
+$P(n,d,i,t)$ = probability of having $n$ Notch receptors, $d$ Delta ligands, and $i$ NICD molecules at time $t$.\\
+
+The Kolmogorov forward equation describes how $P(n,d,i,t)$ changes over a small time step $\Delta t$:
+
+\begin{align*}
+P(n,d,i,t+\Delta t) &= P(n,d,i,t) \\
+&+ P(n-1,d,i,t) \cdot H^+(i) \Delta t \quad \text{(Notch production)} \\
+&+ P(n,d-1,i,t) \cdot H^-(i) \Delta t \quad \text{(Delta production)} \\
+&+ P(n+1,d,i,t) \cdot \gamma(n+1)\Delta t \quad \text{(Notch decay)} \\
+&+ P(n,d+1,i,t) \cdot \gamma(d+1)\Delta t \quad \text{(Delta decay)} \\
+&+ P(n,d,i+1,t) \cdot \gamma_I(i+1)\Delta t \quad \text{(NICD decay)} \\
+&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext}\Delta t \quad \text{(Trans-activation)} \\
+&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]\Delta t
+\end{align*}
+
+Rearranging to find the time derivative:
+
+\begin{align*}
+\frac{P(n,d,i,t+\Delta t) - P(n,d,i,t)}{\Delta t} &= P(n-1,d,i,t) \cdot H^+(i) \\
+&+ P(n,d-1,i,t) \cdot H^-(i) \\
+&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
+&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
+&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
+&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
+&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
+\end{align*}
+
+Taking the limit as $\Delta t \rightarrow 0$:
+
+\begin{align*}
+\frac{\partial P(n,d,i,t)}{\partial t} &= P(n-1,d,i,t) \cdot H^+(i) \\
+&+ P(n,d-1,i,t) \cdot H^-(i) \\
+&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
+&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
+&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
+&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
+&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
+\end{align*}
+
+\subsubsection{Matrix Formulation}
+The Kolmogorov forward equation can be expressed in matrix form as:
+\[
+\frac{d\vec{P}(t)}{dt} = \mathbf{A} \vec{P}(t)
+\]
+
+where $\vec{P}(t)$ is a column vector containing the probabilities for all possible states, and $\mathbf{A}$ is the transition rate matrix. \\
+
+To construct the matrix $\mathbf{A}$, we first establish a one-to-one mapping between the three-dimensional state space $(n,d,i)$ and a one-dimensional index $k$.
+For a system with maximum values $n_{max}$, $d_{max}$, and $i_{max}$, we define:
+\[
+k(n,d,i) = n \cdot (d_{max}+1) \cdot (i_{max}+1) + d \cdot (i_{max}+1) + i
+\]
+for $0 \leq n \leq n_{max}, \; 0 \leq d \leq d_{max}, \; 0 \leq i \leq i_{max}$.
+
+The inverse mapping gives:
+\begin{align*}
+n(k) &= \left\lfloor \frac{k}{(d_{max}+1) \cdot (i_{max}+1)} \right\rfloor \\
+d(k) &= \left\lfloor \frac{k \bmod ((d_{max}+1) \cdot (i_{max}+1))}{i_{max}+1} \right\rfloor \\
+i(k) &= k \bmod (i_{max}+1)
+\end{align*}
+
+With this mapping, we can now define the elements of matrix $\mathbf{A}$. Let $k$ and $k'$ be the indices corresponding to states $(n,d,i)$ and $(n',d',i')$ respectively:
+
+\begin{align*}
+A_{k',k} = 
+\begin{cases}
+H^+(i) & \text{if } (n',d',i') = (n+1,d,i) \quad \text{(Notch production)} \\
+H^-(i) & \text{if } (n',d',i') = (n,d+1,i) \quad \text{(Delta production)} \\
+\gamma(n) & \text{if } (n',d',i') = (n-1,d,i) \quad \text{(Notch decay)} \\
+\gamma(d) & \text{if } (n',d',i') = (n,d-1,i) \quad \text{(Delta decay)} \\
+\gamma_I(i) & \text{if } (n',d',i') = (n,d,i-1) \quad \text{(NICD decay)} \\
+k_T(n)D_{ext} & \text{if } (n',d',i') = (n-1,d,i+1) \quad \text{(Trans-activation)} \\
+-\sum_{k'' \neq k} A_{k'',k} & \text{if } k' = k \quad \text{(Diagonal terms)}
+\end{cases}
+\end{align*}
+
+Note that $A_{k',k}$ represents the transition rate from state $k$ to state $k'$. The diagonal elements $A_{k,k}$ are the negative sum of all outgoing rates from state $k$:
+
+\[
+A_{k,k} = -[H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}]
+\]
+
+where $(n,d,i)$ is the state corresponding to index $k$.\\
+
+The resulting matrix $\mathbf{A}$ is a sparse matrix with the following properties:
+\begin{itemize}
+    \item Dimension: $(n_{max}+1) \cdot (d_{max}+1) \cdot (i_{max}+1) \times (n_{max}+1) \cdot (d_{max}+1) \cdot (i_{max}+1)$
+    \item Number of non-zero elements: $\approx 7 \cdot (n_{max}+1) \cdot (d_{max}+1) \cdot (i_{max}+1)$
+\end{itemize}
+
+To illustrate, let's consider a minimal system with $n_{max} = d_{max} = i_{max} = 1$. This gives us 8 possible states:
+$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1)$
+
+The corresponding transition matrix $\mathbf{A}$ would be:
+
+\begin{align*}
+\mathbf{A} = 
+\begin{pmatrix}
+-\alpha_{000} & \gamma_I & \gamma & 0 & \gamma & 0 & 0 & 0 \\
+H^+(0) & -\alpha_{001} & 0 & \gamma & 0 & \gamma & 0 & 0 \\
+H^-(0) & 0 & -\alpha_{010} & \gamma_I & 0 & 0 & \gamma & 0 \\
+0 & H^-(0) & H^+(0) & -\alpha_{011} & 0 & 0 & 0 & \gamma \\
+H^+(0) & 0 & 0 & 0 & -\alpha_{100} & \gamma_I & \gamma & 0 \\
+0 & H^+(0) & 0 & 0 & H^+(1) & -\alpha_{101} & 0 & \gamma \\
+0 & 0 & H^+(0) & 0 & H^-(1) & 0 & -\alpha_{110} & \gamma_I \\
+0 & 0 & 0 & H^+(0) & 0 & H^-(1) & H^+(1) & -\alpha_{111}
+\end{pmatrix}
+\end{align*}
+
+where:
+\[
+\alpha_{ndi} = H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}
+\]
+
+Additionally, we need to add the terms for trans-activation, which would connect states like $(1,0,0)$ to $(0,0,1)$ with rate $k_T \cdot 1 \cdot D_{ext}$.\\
+
+For a fully labelled matrix with all transitions explicitly shown, the matrix for $n_{max} = d_{max} = i_{max} = 1$ would be:
+
+\[
+\mathbf{A} = 
+\begin{bmatrix}
+A_{11} & A_{12} & A_{13} & A_{14} & A_{15} & A_{16} & A_{17} & A_{18} \\
+B_{11} & B_{12} & B_{13} & B_{14} & B_{15} & B_{16} & B_{17} & B_{18} \\
+C_{11} & C_{12} & C_{13} & C_{14} & C_{15} & C_{16} & C_{17} & C_{18} \\
+D_{11} & D_{12} & D_{13} & D_{14} & D_{15} & D_{16} & D_{17} & D_{18} \\
+E_{11} & E_{12} & E_{13} & E_{14} & E_{15} & E_{16} & E_{17} & E_{18} \\
+F_{11} & F_{12} & F_{13} & F_{14} & F_{15} & F_{16} & F_{17} & F_{18} \\
+G_{11} & G_{12} & G_{13} & G_{14} & G_{15} & G_{16} & G_{17} & G_{18} \\
+H_{11} & H_{12} & H_{13} & H_{14} & H_{15} & H_{16} & H_{17} & H_{18}
+\end{bmatrix}
+\]
+
+where
+
+\begin{tiny}
+\begin{align*}
+A_{11} &= -(H^+(0) + H^-(0)),  & A_{12} &= \gamma_I,  \\
+A_{13} &= \gamma,  & A_{14} &= 0,  \\
+A_{15} &= \gamma,  & A_{16} &= 0,  \\
+A_{17} &= 0,  & A_{18} &= 0,  \\[5pt]
+B_{11} &= H^+(0),  & B_{12} &= -(H^+(1) + H^-(1) + \gamma_I),  \\
+B_{13} &= 0,  & B_{14} &= \gamma,  \\
+B_{15} &= k_T D_{ext},  & B_{16} &= \gamma,  \\
+B_{17} &= 0,  & B_{18} &= 0,  \\[5pt]
+C_{11} &= H^-(0),  & C_{12} &= 0,  \\
+C_{13} &= -(H^+(0) + H^-(0) + \gamma + k_T N_{ext}),  & C_{14} &= \gamma_I,  \\
+C_{15} &= 0,  & C_{16} &= 0,  \\
+C_{17} &= \gamma,  & C_{18} &= 0,  \\[5pt]
+D_{11} &= 0,  & D_{12} &= H^-(1),  \\
+D_{13} &= H^+(0),  & D_{14} &= -(H^+(1) + H^-(1) + \gamma + \gamma_I + k_T N_{ext}),  \\
+D_{15} &= 0,  & D_{16} &= k_T D_{ext},  \\
+D_{17} &= 0,  & D_{18} &= \gamma,  \\[5pt]
+E_{11} &= H^+(0),  & E_{12} &= 0,  \\
+E_{13} &= 0,  & E_{14} &= 0,  \\
+E_{15} &= -(H^+(0) + H^-(0) + \gamma + k_T D_{ext}),  & E_{16} &= \gamma_I,  \\
+E_{17} &= \gamma,  & E_{18} &= 0,  \\[5pt]
+F_{11} &= 0,  & F_{12} &= H^+(1),  \\
+F_{13} &= 0,  & F_{14} &= 0,  \\
+F_{15} &= H^+(0),  & F_{16} &= -(H^+(1) + H^-(1) + \gamma + \gamma_I + k_T D_{ext}),  \\
+F_{17} &= 0,  & F_{18} &= \gamma,  \\[5pt]
+G_{11} &= 0,  & G_{12} &= 0,  \\
+G_{13} &= H^+(0),  & G_{14} &= 0,  \\
+G_{15} &= H^-(0),  & G_{16} &= 0,  \\
+G_{17} &= -(H^+(0) + H^-(0) + \gamma + \gamma + k_T D_{ext} + k_T N_{ext}),  & G_{18} &= \gamma_I,  \\[5pt]
+H_{11} &= 0,  & H_{12} &= 0,  \\
+H_{13} &= 0,  & H_{14} &= H^+(1),  \\
+H_{15} &= 0,  & H_{16} &= H^-(1),  \\
+H_{17} &= H^+(0),  & H_{18} &= -(H^+(1) + H^-(1) + \gamma + \gamma + \gamma_I + k_T D_{ext} + k_T N_{ext}).
+\end{align*}
+\end{tiny}
+
+Each row and column corresponds to a state in the order:
+$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1)$
+
+The matrix element $A_{j,i}$ represents the transition rate from state $i$ to state $j$. The diagonal elements $A_{i,i}$ contain the negative sum of all outgoing rates from state $i$.
 
@@
 
-TBD
+\subsubsection{Setup and Notation}
+
+Let us define:
+\begin{itemize}
+  \item $\langle N \rangle = \sum_{n,d,i} n \cdot P(n,d,i,t)$ = expected number of Notch receptors
+  \item $\langle D \rangle = \sum_{n,d,i} d \cdot P(n,d,i,t)$ = expected number of Delta ligands
+  \item $\langle I \rangle = \sum_{n,d,i} i \cdot P(n,d,i,t)$ = expected number of NICD molecules
+\end{itemize}
+\subsubsection{Deriving ODEs for Expected Values}
+
+\textbf{Now we derive the ODE for the expected value of Notch:}
+
+\begin{align*}
+\frac{d\langle N \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} n \cdot P(n,d,i,t) \\
+&= \sum_{n,d,i} n \cdot \frac{\partial P(n,d,i,t)}{\partial t}
+\end{align*}
+
+Substituting the Kolmogorov forward equation:
+
+\begin{align*}
+\frac{d\langle N \rangle}{dt} &= \sum_{n,d,i} n \cdot \Big[ P(n-1,d,i,t) \cdot H^+(i) \\
+&+ P(n,d-1,i,t) \cdot H^-(i) \\
+&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
+&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
+&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
+&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
+&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \Big]
+\end{align*}
+
+Now we use index shifting to simplify.\\~\\
+
+\textbf{(A) Notch Production} \\
+Let's substitute $n' = n-1$ (so $n = n'+1$):
+\begin{align*}
+\sum_{n,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) &= \sum_{n',d,i} (n'+1) \cdot P(n',d,i,t) \cdot H^+(i) \\
+&= \sum_{n',d,i} n' \cdot P(n',d,i,t) \cdot H^+(i) + \sum_{n',d,i} P(n',d,i,t) \cdot H^+(i) \\
+&= \sum_{n,d,i} n \cdot P(n,d,i,t) \cdot H^+(i) + \sum_{n,d,i} P(n,d,i,t) \cdot H^+(i) \\
+\end{align*}
+
+The first term is $\langle N \cdot H^+(I) \rangle$.
+If we assume that $H^+(I)$ is approximately $H^+(\langle I \rangle)$, then this term becomes $\langle N \rangle \cdot H^+(\langle I \rangle)$.
+The second term is $\langle H^+(I) \rangle \approx H^+(\langle I \rangle)$.
+
+So the contribution of this term is:
+\[
+\sum_{n,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) \approx \langle N \rangle \cdot H^+(\langle I \rangle) + H^+(\langle I \rangle)
+\]
+
+Since we are indexing from 0, and $n-1$ could be negative, the sum should start from $n=1$:
+
+\begin{small}
+\begin{align*}
+\sum_{n \geq 1,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) &= \sum_{n' \geq 0,d,i} (n'+1) \cdot P(n',d,i,t) \cdot H^+(i) \\
+&= \sum_{n' \geq 0,d,i} n' \cdot P(n',d,i,t) \cdot H^+(i) + \sum_{n' \geq 0,d,i} P(n',d,i,t) \cdot H^+(i) \\
+&= \langle N \rangle \cdot H^+(\langle I \rangle) + H^+(\langle I \rangle) \\
+&= H^+(\langle I \rangle) \cdot (1 + \langle N \rangle)
+\end{align*}
+\end{small}
+
+\textbf{(B) Delta Production}\\
+This term does not affect $N$, so the prefactor $n$ remains with the original probability:
+\begin{align*}
+\sum_{n,d,i} n \cdot P(n,d-1,i,t) \cdot H^-(i) &= \sum_{n,d',i} n \cdot P(n,d',i,t) \cdot H^-(i) \\
+&= \langle N \cdot H^-(I) \rangle \\
+&\approx \langle N \rangle \cdot H^-(\langle I \rangle)
+\end{align*}
+
+\textbf{(C) Notch Decay}
+\begin{align*}
+\sum_{n,d,i} n \cdot P(n+1,d,i,t) \cdot \gamma(n+1) &= \sum_{n',d,i} (n'-1) \cdot P(n',d,i,t) \cdot \gamma n' \\
+&= \gamma \sum_{n',d,i} (n'-1)n' \cdot P(n',d,i,t) \\
+&= \gamma \sum_{n',d,i} n'^2 \cdot P(n',d,i,t) - \gamma \sum_{n',d,i} n' \cdot P(n',d,i,t) \\
+&= \gamma(\langle N^2 \rangle - \langle N \rangle)
+\end{align*}
+
+If we assume that the variance of $N$ is small compared to its mean (which is reasonable for high molecule counts), then $\langle N^2 \rangle \approx \langle N \rangle^2$. So:
+\begin{align*}
+\gamma(\langle N^2 \rangle - \langle N \rangle) &\approx \gamma(\langle N \rangle^2 - \langle N \rangle) \\
+&= \gamma \langle N \rangle (\langle N \rangle - 1)
+\end{align*}
+
+\textbf{(D) Delta Decay}
+\begin{align*}
+\sum_{n,d,i} n \cdot P(n,d+1,i,t) \cdot \gamma(d+1) &= \gamma \sum_{n,d',i} n \cdot P(n,d',i,t) \cdot d' \\
+&= \gamma \langle N \cdot D \rangle \\
+&\approx \gamma \langle N \rangle \langle D \rangle
+\end{align*}
+
+\textbf{(E) NICD Decay}
+\begin{align*}
+\sum_{n,d,i} n \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) &= \gamma_I \sum_{n,d,i'} n \cdot P(n,d,i',t) \cdot i' \\
+&= \gamma_I \langle N \cdot I \rangle \\
+&\approx \gamma_I \langle N \rangle \langle I \rangle
+\end{align*}
+
+\textbf{(F) Trans-activation}
+\begin{align*}
+\sum_{n,d,i} n \cdot P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} &= k_T D_{ext} \sum_{n',d,i'} (n'-1) \cdot P(n',d,i',t) \cdot n' \\
+&= k_T D_{ext} \sum_{n',d,i'} (n'^2 - n') \cdot P(n',d,i',t) \\
+&= k_T D_{ext} (\langle N^2 \rangle - \langle N \rangle) \\
+&\approx k_T D_{ext} \langle N \rangle (\langle N \rangle - 1)
+\end{align*}
+
+\textbf{(G) Negative Terms}
+\begin{align*}
+&-\sum_{n,d,i} n \cdot P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \\
+&= -\langle N \cdot H^+(I) \rangle - \langle N \cdot H^-(I) \rangle - \gamma \langle N^2 \rangle - \gamma \langle N \cdot D \rangle - \gamma_I \langle N \cdot I \rangle - k_T D_{ext} \langle N^2 \rangle \\
+&\approx -\langle N \rangle \cdot H^+(\langle I \rangle) - \langle N \rangle \cdot H^-(\langle I \rangle) - \gamma \langle N \rangle^2 - \gamma \langle N \rangle \langle D \rangle - \gamma_I \langle N \rangle
+\langle I \rangle - k_T D_{ext} \langle N \rangle^2
+\end{align*} \\~\\
+
+\textbf{(H) Final Synthesis of $\langle N \rangle$}\\
+
+Combining all these terms:
+
+\begin{align*}
+\frac{d\langle N \rangle}{dt} &= H^+(\langle I \rangle) \cdot (1 + \langle N \rangle) + \langle N \rangle \cdot H^-(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1) + \gamma \langle N \rangle \langle D \rangle \\
+&+ \gamma_I \langle N \rangle \langle I \rangle + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1) \\
+&- \langle N \rangle \cdot H^+(\langle I \rangle) - \langle N \rangle \cdot H^-(\langle I \rangle) - \gamma \langle N \rangle^2 - \gamma \langle N \rangle \langle D \rangle - \gamma_I \langle N \rangle \langle I \rangle - k_T D_{ext} \langle N \rangle^2 \\
+\end{align*}
+
+After cancellations:
+
+\begin{align*}
+\frac{d\langle N \rangle}{dt} &= H^+(\langle I \rangle) + \langle N \rangle \cdot H^+(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1) + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1) \\
+&- \langle N \rangle \cdot H^+(\langle I \rangle) - \gamma \langle N \rangle^2 - k_T D_{ext} \langle N \rangle^2 \\
+&= H^+(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1) - \gamma \langle N \rangle^2 + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1) - k_T D_{ext} \langle N \rangle^2 \\
+&= H^+(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1 - \langle N \rangle) + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1 - \langle N \rangle) \\
+&= H^+(\langle I \rangle) - \gamma \langle N \rangle - k_T D_{ext} \langle N \rangle \\
+\end{align*}
+
+Therefore, the ODE for $\langle N \rangle$ is:
+
+\[
+\frac{d\langle N \rangle}{dt} = H^+(\langle I \rangle) - \gamma \langle N \rangle - k_T D_{ext} \langle N \rangle
+\]
+
+Substituting the definition of $H^+(\langle I \rangle)$:
+
+\[
+\frac{d\langle N \rangle}{dt} = \frac{n_m \langle I \rangle^2}{n_0^2 + \langle I \rangle^2} - \gamma \langle N \rangle - k_T D_{ext} \langle N \rangle
+\]
+
+Which can be written as:
+
+\[
+\frac{dN}{dt} = \frac{n_m I^2}{n_0^2 + I^2} - \gamma N - k_T D_{ext} N
+\]
+
+where we use $N$, $D$, and $I$ as shorthand for the expected values $\langle N \rangle$, $\langle D \rangle$, and $\langle I \rangle$.\\~\\
+
+\textbf{Similarly, we can derive the ODE for the expected value of Delta:}
+
+\begin{align*}
+\frac{d\langle D \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} d \cdot P(n,d,i,t) \\
+&= \sum_{n,d,i} d \cdot \frac{\partial P(n,d,i,t)}{\partial t}
+\end{align*}
+
+We will analyse each term of the Kolmogorov equation, multiplied by $d$, as follows.
+
+\[
+\sum_{n,d,i} d \cdot P(n-1,d,i,t) \cdot H^+(i) = \langle D \cdot H^+(I) \rangle \approx \langle D \rangle \cdot H^+(\langle I \rangle)
+\]
+
+\[
+\sum_{n,d,i} d \cdot P(n,d-1,i,t) \cdot H^-(i) = \langle D \cdot H^-(I) \rangle + \langle H^-(I) \rangle 
+\approx \langle D \rangle \cdot H^-(\langle I \rangle) + H^-(\langle I \rangle)
+\]
+
+\[
+\sum_{n,d,i} d \cdot P(n+1,d,i,t) \cdot \gamma(n+1) \approx \gamma \langle D \rangle \langle N \rangle
+\]
+
+\[
+\sum_{n,d,i} d \cdot P(n,d+1,i,t) \cdot \gamma(d+1) \approx \gamma \langle D \rangle (\langle D \rangle - 1)
+\]
+
+\[
+\sum_{n,d,i} d \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) \approx \gamma_I \langle D \rangle \langle I \rangle
+\]
+
+\[
+\sum_{n,d,i} d \cdot P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \approx k_T D_{ext} \langle D \rangle \langle N \rangle
+\]
+
+Now we consider this term which appears in the ODEs but was not explicitly stated in our Kolmogorov equation setup.
+This is a special case where we need to account for Delta from our cell binding to Notch in neighbouring cells:
+
+\begin{align*}
+-\sum_{n,d,i} d \cdot P(n,d,i,t) \cdot k_T d N_{ext} &= -k_T N_{ext} \sum_{n,d,i} d^2 \cdot P(n,d,i,t) \\
+&= -k_T N_{ext} \langle D^2 \rangle \\
+&\approx -k_T N_{ext} \langle D \rangle^2
+\end{align*}
+
+Now we consider negative terms from original equation.
+\begin{align*}
+&-\sum_{n,d,i} d \cdot P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \\
+&\approx -\langle D \rangle \cdot H^+(\langle I \rangle) - \langle D \rangle \cdot H^-(\langle I \rangle) - \gamma \langle D \rangle \langle N \rangle - \gamma \langle D \rangle^2 - \gamma_I \langle D \rangle \langle I \rangle - k_T D_{ext} \langle D \rangle \langle N \rangle
+\end{align*}
+
+Let's synthesise all terms of $\langle D \rangle$
+
+\begin{align*}
+\frac{d\langle D \rangle}{dt} &= \langle D \rangle \cdot H^+(\langle I \rangle) + \langle D \rangle \cdot H^-(\langle I \rangle) + H^-(\langle I \rangle) + \gamma \langle D \rangle \langle N \rangle \\
+&+ \gamma \langle D \rangle (\langle D \rangle - 1) + \gamma_I \langle D \rangle \langle I \rangle + k_T D_{ext} \langle D \rangle \langle N \rangle - k_T N_{ext} \langle D \rangle^2 \\
+&- \langle D \rangle \cdot H^+(\langle I \rangle) - \langle D \rangle \cdot H^-(\langle I \rangle) - \gamma \langle D \rangle \langle N \rangle - \gamma \langle D \rangle^2 - \gamma_I \langle D \rangle \langle I \rangle - k_T D_{ext} \langle D \rangle \langle N \rangle \\
+&= H^-(\langle I \rangle) - \gamma \langle D \rangle - k_T N_{ext} \langle D \rangle^2
+\end{align*}
+
+Making a final approximation that $\langle D \rangle^2 \approx \langle D \rangle$, we get:
+
+\[
+\frac{d\langle D \rangle}{dt} = H^-(\langle I \rangle) - \gamma \langle D \rangle - k_T N_{ext} \langle D \rangle
+\]
+
+Substituting the definition of $H^-(\langle I \rangle)$:
+
+\[
+\frac{d\langle D \rangle}{dt} = \frac{d_m}{d_0^2 + \langle I \rangle^2} - \gamma \langle D \rangle - k_T N_{ext} \langle D \rangle
+\]
+
+Which can be written as:
+
+\[
+\frac{dD}{dt} = \frac{d_m}{d_0^2 + I^2} - \gamma D - k_T N_{ext} D
+\]
+
+\textbf{Finally, we derive the ODE for NICD:}
+
+\begin{align*}
+\frac{d\langle I \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} i \cdot P(n,d,i,t) \\
+&= \sum_{n,d,i} i \cdot \frac{\partial P(n,d,i,t)}{\partial t}
+\end{align*}
+
+Let's analyse each term:
+
+\[
+\sum_{n,d,i} i \cdot P(n-1,d,i,t) \cdot H^+(i) = \langle I \cdot H^+(I) \rangle \approx \langle I \rangle \cdot H^+(\langle I \rangle)
+\]
+
+\[
+\sum_{n,d,i} i \cdot P(n,d-1,i,t) \cdot H^-(i) \approx \langle I \rangle \cdot H^-(\langle I \rangle)
+\]
+
+\[
+\sum_{n,d,i} i \cdot P(n+1,d,i,t) \cdot \gamma(n+1) \approx \gamma \langle I \rangle \langle N \rangle
+\]
+
+\[
+\sum_{n,d,i} i \cdot P(n,d+1,i,t) \cdot \gamma(d+1) \approx \gamma \langle I \rangle \langle D \rangle
+\]
+
+\[
+\sum_{n,d,i} i \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) \approx \gamma_I \langle I \rangle (\langle I \rangle - 1)
+\]
+
+\[
+\sum_{n,d,i} i \cdot P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \approx k_T D_{ext} \langle I \rangle \langle N \rangle + k_T D_{ext} \langle N \rangle
+\]
+
+Negative terms can be rewritten as:
+\begin{align*}
+&-\sum_{n,d,i} i \cdot P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \\
+&\approx -\langle I \rangle \cdot H^+(\langle I \rangle) - \langle I \rangle \cdot H^-(\langle I \rangle) - \gamma \langle I \rangle \langle N \rangle - \gamma \langle I \rangle \langle D \rangle - \gamma_I \langle I \rangle^2 - k_T D_{ext} \langle I \rangle \langle N \rangle
+\end{align*}
+
+Let's combine all terms for $\langle I \rangle$
+
+\begin{align*}
+\frac{d\langle I \rangle}{dt} &= \langle I \rangle \cdot H^+(\langle I \rangle) + \langle I \rangle \cdot H^-(\langle I \rangle) + \gamma \langle I \rangle \langle N \rangle + \gamma \langle I \rangle \langle D \rangle \\
+&+ \gamma_I \langle I \rangle (\langle I \rangle - 1) + k_T D_{ext} \langle I \rangle \langle N \rangle + k_T D_{ext} \langle N \rangle \\
+&- \langle I \rangle \cdot H^+(\langle I \rangle) - \langle I \rangle \cdot H^-(\langle I \rangle) - \gamma \langle I \rangle \langle N \rangle - \gamma \langle I \rangle \langle D \rangle - \gamma_I \langle I \rangle^2 - k_T D_{ext} \langle I \rangle \langle N \rangle \\
+\end{align*}
+
+Therefore, the ODE for $\langle I \rangle$ is:
+\[
+\frac{d\langle I \rangle}{dt} = k_T D_{ext} \langle N \rangle - \gamma_I \langle I \rangle
+\]
+
+Which can be written as:
+
+\[
+\frac{dI}{dt} = k_T D_{ext} N - \gamma_I I
+\]
+
+\subsubsection{Final System of ODEs}
+
+Combining our three derived equations, we get the following system of ODEs:
+
+\begin{align*}
+\frac{dN}{dt} &= \frac{n_m I^2}{n_0^2 + I^2} - \gamma N - k_T D_{ext} N \\
+\frac{dD}{dt} &= \frac{d_m}{d_0^2 + I^2} - \gamma D - k_T N_{ext} D \\
+\frac{dI}{dt} &= k_T D_{ext} N - \gamma_I I
+\end{align*}
+
+This matches the system presented in section 3.6 of the document:
+
+\begin{align*}
+\frac{dN}{dt} &= \frac{n_m I^2}{n_0^2 + I^2} - k_T N D_{ext} - \gamma N \\
+\frac{dD}{dt} &= \frac{d_m}{d_0^2 + I^2} - k_T D N_{ext} - \gamma D \\
+\frac{dI}{dt} &= k_T N D_{ext} - \gamma_I I
+\end{align*}
+
+\subsubsection{Mapping Between Probabilities and Rates}
+
+To relate the probabilistic formulation to the deterministic ODE system, we need to understand how reaction probabilities map to reaction rates.
+In the stochastic formulation, the probability of a reaction occurring in a small time interval $\Delta t$ is:
+\begin{align*}
+P(\text{reaction}) \approx \text{rate} \times \Delta t
+\end{align*}
+
+For example:
+
+\[
+P(\emptyset \rightarrow N_A) = H^+(I) \Delta t = \frac{n_m I^2}{n_0^2 + I^2} \Delta t 
+\]
+
+In the limit as $\Delta t \rightarrow 0$, these probabilities per unit time become rates, which appear directly in the deterministic ODEs.
+The ODE system essentially tracks the expected values of these stochastic processes under the assumption that fluctuations are small compared to the means.
 

### 2025-03-15 paper/paper.tex
Add abstract to paper.tex

@@
 \section{Abstract}
+
+The Delta-Notch signalling pathway plays a critical role in determining the cell fate of initially homogeneous cells in diverse biological systems.
+Early deterministic models of Delta-Notch interactions, such as that by Collier et al.
+(1996), achieved cell differentiation in linear and 2D cell arrays.
+In this paper, we develop a simple – yet more biologically realistic – ODE model of Delta-Notch signalling and compare the deterministic behaviour with the Collier et al.
+model.
+We derive a probabilistic framework via Kolmogorov forward equations, demonstrating that, in expectation, it converges to our ODE model under appropriate limiting conditions.
+By comparing deterministic ODEs, stochastic ODEs, and Gillespie simulations, we show that natural perturbations in receptor-ligand binding dynamics are sufficient to produce differentiation, even from identical initial conditions.
+Numerical experiments across parameter regimes highlight the robustness of this behaviour and bifurcation analysis further reveals the parameter-dependent bistability.
 

### 2025-03-15 paper/paper.tex
Added SDE section to paper.tex

@@
 \subsection{Stochastic ODE Model}
+
+To capture the inherent randomness in cellular processes while maintaining computational efficiency, we implement a stochastic differential equation (SDE) model based on the chemical Langevin equation formalism.
+This approach represents an intermediate level of detail between the deterministic ODE model and the fully discrete Gillespie algorithm.
+
+Starting from the deterministic ODEs described in Section 3.3, we incorporate stochastic fluctuations by adding noise terms proportional to the square root of the reaction propensities:
+$$
+\begin{aligned}
+  dN &= \left( \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \right) dt + \sigma \cdot N \cdot dW_{N} \\ \\[5pt]
+  dD &= \left( \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \right)  dt + \sigma \cdot D \cdot dW_{D}  \\[5pt]
+  dI &= \left( k_{T}ND_{ext} - \gamma_{I}I \right) dt + \sigma \cdot I \cdot dW_{I}
+\end{aligned}
+$$
+where $dW_N$, $dW_D$, and $dW_I$ are independent Wiener processes representing white noise.
+The structure of the noise terms follows from the assumption that noise in each reaction channel is proportional to the current concentration of those molecules.
+\par
+
+For numerical implementation, we employ the Euler-Maruyama scheme with time step $dt$.
+The deterministic step is first calculated, and then the Wiener increments are sampled from a normal distribution with mean 0 and standard deviation $\sqrt{dt}$.
+Each increment is scaled by the its respective variable and added to the results of the deterministic step.
 

### 2025-03-15 paper/paper.tex
Added ODE model section to paper.tex

@@
 \subsection{Deterministic ODE Model}
+
+We implemented the following deterministic ODE model adapted from Boareto et al. (2015):
+
+$$
+\begin{aligned}
+  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\[5pt]
+  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\[5pt]
+  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
+\end{aligned}
+$$
+
+where $N$ is the number of intracellular Notch molecules, $D$ is the number of intracellular Delta molecules, and $I$ is the number of NICD molecules (see $3.1$ reference).
+This model is a simplification of the model found in Boareto et al.
+That model included an additonal equation for the Jagged molecule and kept track of cis-inhibition: where a Notch and Delta bind together inside the same cell, resulting in the degradation of both cells without contributing to the amount of NICD.
+So as to more directly compare our results with those of Collier et al., we ignore Jagged interactions and cis-inhibition.
+Including an NICD variable was necessary, however, as this additional interaction allowed for the assignment of probabilities to each reaction event and thus for the derivation of the Kolmogorov Equations and Gillespie algorithm.
 

### 2025-03-15 paper/paper.tex
Add linear domain heatmap figure to paper.tex
Added SDE heatmap figure to paper.tex

@@
 
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/ODEModel_TwoCell_Heatmap.png}
+  \caption{Results of the linear domain deterministic ODE model. Heatmap shows Notch levels in each of 8 cells over 1000 time steps.}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/SDE_CellLine_HeatMap.png}
+  \caption{Results of the linear domain SDE model. Heatmap shows Notch levels in each of 8 cells over 1000 time steps.}
+\end{figure}
+
 \subsection{Hexagonal Domain}

### 2025-03-17 paper/paper.tex
Added hex grid figure paper.tex
Differentiation time analysis for two-cell models.

@@
 
-\newpage
-
 Furthermore, we will assume that there is no cis-interaction and no Serrate ligand. This gives us the following set of chemical reactions that can occur in a single time step:
@@
 
-Let us define:\\
-$P(n,d,i,t)$ = probability of having $n$ Notch receptors, $d$ Delta ligands, and $i$ NICD molecules at time $t$.\\
-
+Let us define $P(n,d,i,t)$ to be the probability of having $n$ Notch receptors, $d$ Delta ligands, and $i$ NICD molecules at time $t$.
 The Kolmogorov forward equation describes how $P(n,d,i,t)$ changes over a small time step $\Delta t$:
@@
 
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/ODE_HexGrid.png}
+  \caption{Results of the 8x8 hexagonal domain ODE model showing final Notch values per cell.}
+\end{figure}
+
 \section{Discussion}

### 2025-03-17 models/config.py

@@ -16,7 +16,7 @@ PI = 0.05
 NOISE = 0.02
 
 # Set the simulation length and initial cell state
-TIME = 1000
+TIME = 1500
 STEPS = 10000
 INITIAL_CELL = np.array([200, 200, 100])
 

### 2025-03-17 models/gillespie.py

@@ -153,6 +153,7 @@ def gillespie(domain, samples = 1):
     data_time = []
     data_state = []
     data_mean = []
+    data_diff = []
 
     # Run 'samples' simulations
     start = time.time()
@@ -169,6 +170,11 @@ def gillespie(domain, samples = 1):
         if size == 2 and v_state[-1, 1, 0] > v_state[-1, 0, 0]:
             v_state = np.roll(v_state, 1, axis = 1)
 
+        # Compute the differentiation time for two cell simulations
+        if size == 2:
+            max_index = np.argmax(v_state[:, 1, 0] < 1)
+            data_diff.append(v_time[max_index])
+
         # Linearly interpolate the state vector
         f_interp = interp1d(v_time, v_state, axis = 0, fill_value = "extrapolate")
         v_interp = f_interp(np.linspace(0, TIME, STEPS))
@@ -179,10 +185,11 @@ def gillespie(domain, samples = 1):
         data_mean.append(v_interp)
         logger.info(f" - {time.time() - start:.4f}s: Finished simulation {i}")
 
-    # Average over the interpolated data to get the mean
+    # Average over the interpolated data to get STD and the mean
+    data_std = np.std(data_mean, axis = 0)
     data_mean = np.average(data_mean, axis = 0) 
     logger.info(f"Finished in {time.time() - start:.4f}s")
-    return data_time, data_state, data_mean
+    return data_time, data_state, data_mean, data_std, data_diff
 
 
 

### 2025-03-17 models/main.py

@@ -6,7 +6,7 @@ import time
 import numba
 
 # Import modules 
-import visualization
+import visualization as vis
 import domains
 import gillespie 
 import ode
@@ -23,21 +23,32 @@ def main():
     # Two Cell Domain 
     domain = domains.linear(2, "dirichlet")
 
-    # Deterministic Model
-    deterministic_data = ode.ode(domain, "Deterministic", samples = 25)
-    visualization.two_cell_individual(domain, deterministic_data, "Deterministic")
+    # Deterministic Model (Linear ICs)
+    linear_data = ode.ode(domain, "Deterministic-Linear", samples = 10)
+    vis.two_cell_individual(domain, linear_data, "Linear Deterministic")
+    
+    # Deterministic Model (Random ICs)
+    random_data = ode.ode(domain, "Deterministic-Random", samples = 10)
+    vis.two_cell_individual(domain, random_data, "Random Deterministic")
 
     # Stochastic Model
-    stochastic_data = ode.ode(domain, "Stochastic", samples = 25)
-    visualization.two_cell_individual(domain, stochastic_data, "Stochastic")
+    stochastic_data = ode.ode(domain, "Stochastic", samples = 10)
+    vis.two_cell_individual(domain, stochastic_data, "Stochastic")
 
     # Gillespie Model
-    gillespie_data = gillespie.gillespie(domain, samples = 25)
-    visualization.two_cell_individual(domain, gillespie_data, "Gillespie")
+    gillespie_data = gillespie.gillespie(domain, samples = 10)
+    vis.two_cell_individual(domain, gillespie_data, "Gillespie")
 
     # Comparison Plot
-    all_data = (deterministic_data, stochastic_data, gillespie_data)
-    visualization.two_cell_comparison(all_data)
+    all_data = (random_data, stochastic_data, gillespie_data)
+    vis.two_cell_comparison(all_data)
+
+    # Differentiation time plots (with additional samples)
+    linear_data = ode.ode(domain, "Deterministic-Linear", samples = 50)
+    stochastic_data = ode.ode(domain, "Stochastic", samples = 500)
+    gillespie_data = gillespie.gillespie(domain, samples = 500)
+    vis.two_cell_deterministic_differentiation(domain, linear_data)
+    vis.two_cell_stochastic_differentiation(domain, stochastic_data, gillespie_data)
     
     # Linear Domain (Dirichlet)
     # domain, size = domains.linear(11, "dirichlet")

### 2025-03-17 models/ode.py

@@ -47,9 +47,10 @@ def deterministic_step(domain, state):
 
 # Run a single simulation of the deterministic ODE model
 @numba.njit
-def simulate(domain, dt, perturbations, wiener):
+def simulate(domain, perturbations, wiener):
 
     # Initialize the states variable
+    dt = TIME / STEPS
     name, neighbours, size = domain
     states = np.empty((STEPS, size, 3))
 
@@ -70,45 +71,68 @@ def simulate(domain, dt, perturbations, wiener):
     return np.linspace(0, TIME, STEPS), states
 
 
+# Return the initial perturbations and wiener values
+def initialize_simulation(index, domain, mode, samples):
+
+    # Default to perturbation and wiener coefficients of 0
+    name, neighbours, size = domain
+    perturbations = np.zeros((size, 3))
+    wiener = np.zeros((STEPS, size, 3))
+
+    # Generate consistent initial perturbations 
+    if mode == "Deterministic-Linear" and size == 2:
+        value = INITIAL_CELL * PI * ((index + 1) / samples)
+        perturbations = np.array([value, -value])
+
+    # Generate random initial perturbations 
+    if mode == "Deterministic-Random":
+        perturbations = np.random.normal(0, INITIAL_CELL * PI, (size, 3))
+
+    # Generate wiener values 
+    if mode == "Stochastic":
+        dt = TIME / STEPS
+        wiener = np.random.normal(0, np.sqrt(dt), (STEPS, size, 3))
+
+    return perturbations, wiener
+
+
 # Run a gillespie model with given domain and size
 def ode(domain, mode, samples = 1):
     
     # Initialize the simulations
-    dt = TIME / STEPS
     name, neighbours, size = domain
-    data_time, data_state = [], []
+    data_time, data_state, data_diff = [], [], []
 
     # Run simulations simulations
     start = time.time()
     logger.info(f"Model: {mode} ODE, Samples: {samples}, Domain: {name}")
     for i in range(samples):
 
-        # Generate initial perturbations 
-        if mode == "Deterministic":
-            perturbations = np.random.normal(0, INITIAL_CELL * PI, (size, 3))
-            wiener = np.zeros((STEPS, size, 3))
-
-        # Generate wiener values 
-        if mode == "Stochastic":
-            perturbations = np.zeros((size, 3))
-            wiener = np.random.normal(0, np.sqrt(dt), (STEPS, size, 3))
-        
-        # Run the simulation
-        v_time, v_state = simulate(domain, dt, perturbations, wiener)
[…]

### 2025-03-17 models/visualization.py

@@ -1,6 +1,7 @@
 # Import libraries
 from matplotlib.colors import LinearSegmentedColormap
 import matplotlib.pyplot as plt
+import matplotlib as mpl
 import numpy as np
 
 # Import modules
@@ -11,21 +12,30 @@ from config import TIME, STEPS
 def two_cell_individual(domain, data, model):
 
     name, neighbours, size = domain
-    times, states, means = data
+    times, states, means, stds, diffs = data
     fig, axs = plt.subplots(nrows = 3, sharex = True)
 
     # Iterate through axes (0: Notch, 1: Delta, 2: NICD)
     for i in range(3):
+        
+        # Reduce the linewidth
+        mpl.rcParams['lines.linewidth'] = 1
 
         # Plot the trajectories for each sample
         for j in range(len(states)):
-            axs[i].plot(times[j], states[j][:, 0, i], color = "green", alpha = 0.1)
-            axs[i].plot(times[j], states[j][:, 1, i], color = "orange", alpha = 0.1)
+            axs[i].plot(times[j], states[j][:, 0, i], color = "green", alpha = 0.2)
+            axs[i].plot(times[j], states[j][:, 1, i], color = "orange", alpha = 0.2)
 
         # Plot the means at full opacity
-        default_times = np.linspace(0, TIME, STEPS)
-        axs[i].plot(default_times, means[:, 0, i], color = "green")
-        axs[i].plot(default_times, means[:, 1, i], color = "orange")
+        vT = np.linspace(0, TIME, STEPS)
+        axs[i].plot(vT, means[:, 0, i], color = "darkgreen")
+        axs[i].plot(vT, means[:, 1, i], color = "darkorange")
+
+        # Plot the standard deviations
+        axs[i].plot(vT, means[:, 0, i] + stds[:, 0, i], "--", color = "green")
+        axs[i].plot(vT, means[:, 0, i] - stds[:, 0, i], "--", color = "green")
+        axs[i].plot(vT, means[:, 1, i] + stds[:, 1, i], "--", color = "orange")
+        axs[i].plot(vT, means[:, 1, i] - stds[:, 1, i], "--", color = "orange")
 
     # Set axis labels
     axs[0].set_ylabel("Notch Molecules")
@@ -39,7 +49,8 @@ def two_cell_individual(domain, data, model):
     # Save to the img/ foler
     plt.tight_layout()
     fname = "_".join(f"{name} {model}".lower().split(" "))
[…]

### 2025-03-24 models/config.py
Stability analysis.

@@ -3,14 +3,14 @@ import numba
 
 # Default simulation parameters
 DEFAULT = np.array([
-    2,       # K: Hill function exponent
-    10,      # NM: Maximum rate of Notch production
-    10,      # DM: Maximum rate of Delta production
-    100,     # N0: Notch hill function (H+) half-max
-    100,     # D0: Delta hill function (H-) half-max
-    0.0001,  # KT: Delta-Notch binding rate
-    0.02,    # G: Extracellular molecule decay rate 
-    0.025,   # GI: NICD decay rate
+    2,       # 0, K: Hill function exponent
+    10,      # 1, NM: Maximum rate of Notch production
+    10,      # 2, DM: Maximum rate of Delta production
+    100,     # 3, N0: Notch hill function (H+) half-max
+    100,     # 4, D0: Delta hill function (H-) half-max
+    0.0001,  # 5, KT: Delta-Notch binding rate
+    0.02,    # 6, G: Extracellular molecule decay rate 
+    0.025,   # 7, GI: NICD decay rate
 ])
 
 # Default initial cell state

### 2025-03-24 models/experiments.py

@@ -5,7 +5,7 @@ import numba
 from tqdm import trange
 
 # Import modules 
-from config import DEFAULT_CELL
+from config import DEFAULT_CELL, DEFAULT
 from helpers import *
 import visualization as vis
 import domains
@@ -209,33 +209,64 @@ def ex12(runs = 100, shift = np.linspace(0, 100, 6)):
     print("\nExperiment Completed!\n")
 
 
-# Experiment 13: Effect of hyperparameters on cell differentiation in all three models 
-def ex13():
+# Helper function for stability analysis on two parameters i1, i2 in DEFAULT
+def two_parameter_stability(i1, i2, model, samples = 9):
 
-    # Generate the grid of parameter ranges 
-    print("\nRunning Experiment 13\n")
-    grid = np.meshgrid(
-        [2],                                          # K
-        [4, 7, 10, 13, 16],                           # NM
-        [4, 7, 10, 13, 16],                           # DM
-        [100],                                        # N0
-        [100],                                        # D0
-        [0.00004, 0.00007, 0.0001, 0.00013, 0.00016], # KT
-        [0.01, 0.015, 0.02, 0.025, 0.03],             # G
-        [0.01, 0.015, 0.02, 0.025, 0.03],             # GI
-    )
-
+    # Helper function to generate parameter space for parameter i
+    def parameter_space(i):
+        if i in [i1, i2]:
+            return DEFAULT[i] * np.logspace(-1, 1, samples)
+        else:
+            return [DEFAULT[i]]
+  
+    # Generate list of parameter arrays for each simulation (plist)
+    grid = np.meshgrid(*[parameter_space(i) for i in range(8)])
     params = np.array(grid).T.reshape(-1,8)
     success = np.empty(len(params))
 
     # Iterate through the list of parameters
     domain = domains.linear(2, "Dirichlet")
     for i in trange(len(params)):
-        vT, vS = ode.ode(domain, DEFAULT_CELL, params = params[i])
-        success[i] = 0 if differentiation(vT, vS) == vT[-1] else 1
 
-    # Plot the results
-    vis.vis13(params, success)
+        # Set the initial condition (adding 2% initial perturbation for deterministic)
+        initial = np.array([DEFAULT_CELL, DEFAULT_CELL], dtype = np.float64)
+        if model == "Deterministic":
[…]

### 2025-03-24 models/gillespie.py

@@ -114,7 +114,7 @@ def gillespie(domain, initial, params = DEFAULT):
 
     # Initialize the state and time vectors
     vT, vS = [], []
-    while t <= TIME:
+    while t <= TIME and len(vT) < 100000:
 
         # Update reaction rates
         rates = update_reaction_rates(state, neighbours, params)

### 2025-03-24 models/main.py

@@ -6,23 +6,23 @@ import visualization as vis
 def main():
 
     # Two-Cell Tests
-    ex.ex01()
-    ex.ex02()
-    ex.ex03()
+    # ex.ex01()
+    # ex.ex02()
+    # ex.ex03()
 
     # Linear Domain Tests
-    ex.ex04()
-    ex.ex05()
-    ex.ex06()
+    # ex.ex04()
+    # ex.ex05()
+    # ex.ex06()
 
     # Hexagonal Domain Tests
     # TBD
 
     # Two-Cell Experiments
-    ex.ex10()
-    ex.ex11()
-    ex.ex12()
-    # ex.ex13()
+    # ex.ex10()
+    # ex.ex11()
+    # ex.ex12()
+    ex.ex13()
 
 
 main()

### 2025-03-24 models/visualization.py

@@ -1,12 +1,14 @@
 # Import libraries
 from scipy.stats import gaussian_kde
+from scipy.spatial import ConvexHull
 from matplotlib.colors import LinearSegmentedColormap
+from matplotlib.patches import Polygon
 import matplotlib.pyplot as plt
 import matplotlib as mpl
 import numpy as np
 
 # Import modules
-from config import TIME, STEPS
+from config import TIME, STEPS, DEFAULT
 
 
 # Visualization 01-03 helper function:
@@ -172,5 +174,96 @@ def vis12(shift, fates):
     plt.close()
 
 
-def vis13(params, success):
-    return
+# Helper function for vis13
+def stability_visualization(ax, grids, pair):
+    
+    # List of parameter names in LaTeX
+    NAMES = [
+        r"$K$",
+        r"$N_M$",
+        r"$D_M$",
+        r"$N_0$",
+        r"$D_0$",
+        r"$K_T$",
+        r"$\gamma$",
+        r"$\gamma_I$"
+    ]
+
+    # Unpack parameters, set constants
+    i1, i2 = pair
+    d_grid, s_grid, g_grid = grids
+
+    # Compute the convex hull for each grid
+    d_hull = d_grid[ConvexHull(d_grid).vertices, :]
+    s_hull = s_grid[ConvexHull(s_grid).vertices, :]
+    g_hull = g_grid[ConvexHull(g_grid).vertices, :]
+
+    # Plot the convex hull for each grid
+    d_poly = Polygon(d_hull, fc='none', ec="b", label="Deterministic ODE")
+    s_poly = Polygon(s_hull, fc='none', ec="g", label="Stochastic ODE")
+    g_poly = Polygon(g_hull, fc='none', ec="r", label="Agent-Based")
+
+    # Add the patches to the axes object
+    ax.add_patch(d_poly)
+    ax.add_patch(s_poly)
+    ax.add_patch(g_poly)
+   
+    # Set tick formatters
+    ax.xaxis.set_major_formatter(mpl.ticker.StrMethodFormatter("{x:.1e}"))
+    ax.yaxis.set_major_formatter(mpl.ticker.StrMethodFormatter("{x:.1e}"))
+
+    # Set the x and y limits, titles, labels, etc.
[…]

### 2025-03-24 models/experiments.py
Add pattern finding code.

@@ -210,7 +210,7 @@ def ex12(runs = 100, shift = np.linspace(0, 100, 6)):
 
 
 # Helper function for stability analysis on two parameters i1, i2 in DEFAULT
-def two_parameter_stability(i1, i2, model, samples = 9):
+def two_parameter_stability(i1, i2, model, samples = 25):
 
     # Helper function to generate parameter space for parameter i
     def parameter_space(i):
@@ -267,6 +267,19 @@ def ex13():
         grids.append([d_grid, s_grid, g_grid])
 
     vis.vis13(grids, pairs)
+    print("\nExperiment Completed!\n")
+
+
+# Experiment 14: Linear model pattern analysis
+def ex14():
+
+    print("\nRunning Experiment 14\n")
+
+    domain = domains.linear(9, "Dirichlet")
+    vT, vS = ode.ode(domain, DEFAULT_CELL, noise = 0.02)
+    print(pattern(vS))
+
+
     print("\nExperiment Completed!\n")
 
 

### 2025-03-24 models/helpers.py

@@ -29,4 +29,13 @@ def differentiation(vT, vS):
     return vT[np.max(vI)]
 
 
+# Get the pattern for a single simulation
+def pattern(vS):
+    return np.where(vS[-1, :, 0] < 1, 0, 1)
 
+# NOTE: In the function above, vS is stores states with shape (# steps, # cells, 3)
+# I only use one index to denote cells in both linear and hexagonal domains
+# If your vS has shape (# steps, width, height, 3) you could use the code below
+
+def pattern_hexagonal(vS):
+    return np.where(vS[-1, :, :, 0] < 1, 0, 1)

### 2025-03-24 models/main.py

@@ -22,7 +22,8 @@ def main():
     # ex.ex10()
     # ex.ex11()
     # ex.ex12()
-    ex.ex13()
+    # ex.ex13()
+    ex.ex14()
 
 
 main()

### 2025-03-24 models/visualization.py

@@ -213,8 +213,6 @@ def stability_visualization(ax, grids, pair):
     ax.yaxis.set_major_formatter(mpl.ticker.StrMethodFormatter("{x:.1e}"))
 
     # Set the x and y limits, titles, labels, etc.
-    print(f"i1: {i1}, i2: {i2}")
-    print(f"X Limit: {DEFAULT[i1] * 10}, Y Limit: {DEFAULT[i2] * 10}")
     ax.xaxis.set_ticks(np.linspace(0, DEFAULT[i1] * 10, 3))
     ax.yaxis.set_ticks(np.linspace(0, DEFAULT[i2] * 10, 3))
     ax.set_xlim(0, DEFAULT[i1] * 10)

### 2025-03-24 models/helpers.py
Update pattern function

@@ -29,13 +29,13 @@ def differentiation(vT, vS):
     return vT[np.max(vI)]
 
 
-# Get the pattern for a single simulation
+# Get the pattern for a single simulation. Returns an array of 0's and 1's
+# where pattern[i] = 0 if the cell has Notch < 1 and pattern[i] = 1 otherwise
 def pattern(vS):
     return np.where(vS[-1, :, 0] < 1, 0, 1)
 
 # NOTE: In the function above, vS is stores states with shape (# steps, # cells, 3)
 # I only use one index to denote cells in both linear and hexagonal domains
 # If your vS has shape (# steps, width, height, 3) you could use the code below
-
 def pattern_hexagonal(vS):
     return np.where(vS[-1, :, :, 0] < 1, 0, 1)

### 2025-03-25 models/experiments.py
Adding hexagon visualization function and experiment

@@ -283,3 +283,32 @@ def ex14():
     print("\nExperiment Completed!\n")
 
 
+# Experiment 15: Hexagonal model pattern for visualization
+def ex15(runs = 1, pb = 0.05):
+    # Run 1 simulation of the stochastic ODE model
+    print("\nRunning15\n")
+    times = np.empty((100))
+    domain = domains.hexagonal_periodic(7,7)
+    size = domain[2]
+    states = np.empty((runs, 100, size, 3))
+    
+    for i in trange(runs):
+        # Generate initial conditions
+        # Create initial conditions for the entire domain
+        size = domain[2] 
+        perturbations = np.random.normal(0, DEFAULT_CELL * pb, (size, 3))
+        initial = DEFAULT_CELL + perturbations
+        
+        # Run the simulation, sort the cells, interpolate the results
+        vT, vS = ode.ode(domain, initial, noise = 0.02)
+        #vS = vS[:, vS[-1, :, 0].argsort(), :]
+        times, states[i] = interpolate(100, vT, vS)
+    
+    # Use the updated pattern_hexagonal function
+    cell_pattern = pattern(vS)
+    print(cell_pattern)
+    vis.visualize_pattern(cell_pattern, domain='hexagonal', ny=7, nx=7)
+    #print(vS[-1, :, 0])        
+    
+    print("\nExperiment Completed!\n")
+    return cell_pattern

### 2025-03-25 models/visualization.py

@@ -2,7 +2,7 @@
 from scipy.stats import gaussian_kde
 from scipy.spatial import ConvexHull
 from matplotlib.colors import LinearSegmentedColormap
-from matplotlib.patches import Polygon
+from matplotlib.patches import Polygon, RegularPolygon, Rectangle
 import matplotlib.pyplot as plt
 import matplotlib as mpl
 import numpy as np
@@ -263,5 +263,117 @@ def vis13(grids, pairs):
     plt.savefig(f"img/vis13.pdf", dpi = 200)
     plt.close()
 
+#  Produces a visualization of the cell pattern for either hexagonal or linear domains
+def visualize_pattern(pattern, domain='hexagonal', ny=7, nx=7, radius=1.0):
+    """
+    Visualize a pattern in either hexagonal or linear mode:
+    
+    Hexagonal Mode:
+    - pattern is a 1D numpy array of 0s and 1s
+    - ny is the number of rows
+    - nx is the number of columns
+    - radius controls the size of hexagons
+    
+    Linear Mode:
+    - pattern is a 1D numpy array of 0s and 1s
+    - Visualizes a single row of rectangular cells
+    """
+    # Create the figure and axis
+    fig, ax = plt.subplots(figsize=(10, 8))
+    
+    if domain == 'hexagonal':
+        # Ensure the pattern matches the grid size
+        assert len(pattern) == ny * nx, "Pattern length must match grid size"
+        
+        # Hexagon dimensions
+        hex_height = np.sqrt(3) * radius
+        
+        # Initialize hexagons in a parallelogram grid
+        pattern_index = 0
+        for j in range(ny):
+            for l in range(nx):
+                # Calculate hexagon center position for parallelogram grid
+                # Shift each row to right while starting from top
+                x = l * np.sqrt(3) * radius + j * 0.866 * radius  # Shifted right
+                y = (ny - 1 - j) * hex_height * 0.866  # Invert y to start from top
+                
+                # Determine color based on the grid value
+                facecolor = 'black' if pattern[pattern_index] == 0 else 'white'
+                pattern_index += 1
+                
+                # Create hexagon patch
+                hexagon = RegularPolygon(
+                    (x, y), 
+                    numVertices=6, 
+                    radius=radius,
[…]

### 2025-03-25 models/visualization.py
Update visualization.py

@@ -367,13 +367,13 @@ def visualize_pattern(pattern, domain='hexagonal', ny=7, nx=7, radius=1.0):
     plt.show()
 
     # Example usages
-    # Hexagonal mode
+    # Hexagonal domain
     # hex_pattern = np.array([1, 0, 1, 1, 0, 1, 1, 1, 0, 0, 1, 1, 0, 0, 1, 1, 1, 0, 0, 1, 
     #            1, 1, 0, 0, 1, 0, 0, 0, 1, 1, 0, 1, 1, 1, 0, 0, 1, 0, 1, 0, 
     #            1, 1, 0, 1, 1, 0, 0, 1, 0])
     # visualize_pattern(hex_pattern, domain='hexagonal', ny=7, nx=7)
 
-    # Linear mode
+    # Linear domain
     # linear_pattern = np.array([0, 0, 1, 1, 0, 1, 0, 1, 1, 1])
-    # visualize_pattern(linear_pattern, mode='linear')
+    # visualize_pattern(linear_pattern, domain='linear')
 

### 2025-03-27 paper/paper.tex
Some modifications to the paper.

@@
 
-\title{Mathematical Models of Delta-Notch Signalling}  
+\title{Comparing Models of Delta-Notch Signalling}  
 \author{Riley Wheadon, Edwin Huras, Ziyi Zhuang}
@@
 The Delta-Notch signalling pathway plays a critical role in determining the cell fate of initially homogeneous cells in diverse biological systems.
-Early deterministic models of Delta-Notch interactions, such as that by Collier et al.
-(1996), achieved cell differentiation in linear and 2D cell arrays.
-In this paper, we develop a simple – yet more biologically realistic – ODE model of Delta-Notch signalling and compare the deterministic behaviour with the Collier et al.
-model.
-We derive a probabilistic framework via Kolmogorov forward equations, demonstrating that, in expectation, it converges to our ODE model under appropriate limiting conditions.
-By comparing deterministic ODEs, stochastic ODEs, and Gillespie simulations, we show that natural perturbations in receptor-ligand binding dynamics are sufficient to produce differentiation, even from identical initial conditions.
-Numerical experiments across parameter regimes highlight the robustness of this behaviour and bifurcation analysis further reveals the parameter-dependent bistability.
+Early deterministic models of Delta-Notch interactions, such as that of Collier et al. \cite{collier_pattern_1996}, achieved cell differentiation in linear and 2D cell arrays.
+In this paper, we develop a simple yet more biologically realistic model of Delta-Notch signalling and test it using three modelling formalisms: deterministic ODE, stochastic ODE, and agent-based.
+Using the Kolmogorov forward equations, we show that the agent-based formalisms converges to the system of deterministic ODEs in expectation.
+We also demonstrate that the stochastic ODE and agent-based formalisms can produce cell differentiation even from identical initial conditions.
+Finally, we study the effect of parameters such as stochastic noise, domain size, and boundary conditions on the behaviour of our models.
+Our findings provide important insights into how different mathematical formalisms can affect simulations of the Delta-Notch signalling network with significant implications for future modelling work.
 
@@
 
+The Delta-Notch signalling network plays
+
 \medskip
 
-Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor \cite{collier_pattern_1996}.
+Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor \cite{bray_notch_2006}.
 This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch and \emph{inhibits} Delta.
+It is also possible for a Delta ligand to bind to a Notch receptor on the same cell. This phenomenon, known as \emph{cis-inhibition}, cleaves the Notch receptor but does not release NICD.
+Serrate (Jagged) is a ligand similar to Delta which can also bind to Notch.
+However, NICD promotes Serrate, which creates a double-positive feedaback loop.
 
-\begin{itemize}
-  \item The Serrate (Jagged) ligand can also bind to Notch. Unlike Delta, NICD promotes Jagged, which creates a double-positive feedback loop.
-  \item Interaction between Delta ligands and Notch receptors from the same cell (cis-interaction) removes both molecules and does not trigger the release of NICD.
-\end{itemize}
+\medskip
 

### 2025-03-27 models/config.py

@@ -14,7 +14,7 @@ DEFAULT = np.array([
 ])
 
 # Default initial cell state
-DEFAULT_CELL = np.array([200, 200, 100])
+DEFAULT_CELL = np.array([200, 200, 100], dtype = np.float64)
 
 # Default simulation length and time steps
 TIME = 1500

### 2025-03-27 models/experiments.py

@@ -270,16 +270,135 @@ def ex13():
     print("\nExperiment Completed!\n")
 
 
-# Experiment 14: Linear model pattern analysis
-def ex14():
+# Helper function to compute the "structure" of a linear pattern
+def get_structure(pattern):
 
-    print("\nRunning Experiment 14\n")
+    # We want to return a tuple of tuples (x, y) where x contains the multiset
+    # of groups sizes of consecutive 0's and y contains the multiset of group
+    # sizes of consecutive 1's
+
+    # EXAMPLE: ((), ()) means there are no consecutive 0's and no consecutive 1's
+    # EXAMPLE: ((3, 2), ()) means there is a triplet and a pair of consecutive 0's
+    # EXAMPLE: ((2, 2), (3)) means there are two pairs of 0's and one triplet of 1's
+
+    # prev is an (index, value) pair
+    prev = (0, pattern[0]) 
+    structure = [[], []]
+
+    # Add a dummy variable to pattern, then find duplicates
+    pattern.append(-1)
+    for i in range(1, len(pattern)):
+
+        # If cell is the same as prev, do nothing
+        if pattern[i] == prev[1]:
+            continue
+
+        # If prev is sufficiently old, add a duplicate to zeros or ones
+        if i - prev[0] > 1:
+            structure[prev[1]].append(i - prev[0])
+
+        # Reset prev
+        prev = (i, pattern[i])
+
+    x = sorted(structure[0], reverse = True)
+    y = sorted(structure[1], reverse = True)
+    return (tuple(x), tuple(y))
+
+
+# Test cases for get_structure
+# print(get_structure([0, 1, 0, 1, 0, 1, 0, 1, 0])) # ((), ())
+# print(get_structure([0, 0, 0, 1, 0, 0, 1, 0, 1])) # ((3, 2), ())
+# print(get_structure([0, 0, 1, 0, 1, 1, 1, 0, 0])) # ((2, 2), (3))
+# print(get_structure([1, 1, 1, 1, 1, 0, 0, 0, 0])) # ((4), (5))
+
+
+# Helper function to run structure analysis for a specific model
+def structure_analysis(domain, model, samples = 1000, file = "results/ex14.txt"):
+
+    structures = {}
+    name, neighbours, size = domain
+
+    # Run samples with the given model
+    for i in trange(samples):
+
+        # Set the initial condition (adding 2% initial perturbation for deterministic)
+        initial = np.tile(DEFAULT_CELL, (size, 1))
+        if model == "Deterministic":
[…]

### 2025-03-27 models/gillespie.py

@@ -104,7 +104,7 @@ def update_state(state, event, neighbours):
 
 # Runs a single simulation 
 @numba.njit
-def gillespie(domain, initial, params = DEFAULT):
+def gillespie(domain, initial, params = DEFAULT, stop = 10 ** 5):
 
     # Initialize the simulation
     t = 0
@@ -114,7 +114,7 @@ def gillespie(domain, initial, params = DEFAULT):
 
     # Initialize the state and time vectors
     vT, vS = [], []
-    while t <= TIME and len(vT) < 100000:
+    while t <= TIME and len(vT) < stop:
 
         # Update reaction rates
         rates = update_reaction_rates(state, neighbours, params)

### 2025-03-27 paper/paper.tex
Reformat paper.

@@
 \usepackage[left=1.5in, right=1.5in, top=1in, bottom=1in]{geometry}
+\usepackage{lscape}
 \usepackage{graphicx}
@@
 \usepackage{hyperref}
+\usepackage{array}
 
@@
 
-\section{Abstract}
+\section*{Abstract}
 
-The Delta-Notch signalling pathway plays a critical role in determining the cell fate of initially homogeneous cells in diverse biological systems.
-Early deterministic models of Delta-Notch interactions, such as that of Collier et al. \cite{collier_pattern_1996}, achieved cell differentiation in linear and 2D cell arrays.
-In this paper, we develop a simple yet more biologically realistic model of Delta-Notch signalling and test it using three modelling formalisms: deterministic ODE, stochastic ODE, and agent-based.
+The Delta-Notch signalling pathway plays a crucial role in determining the cell fate of initially homogeneous cells in a wide array biological systems.
+Early deterministic models of the Delta-Notch pathway achieved cell differentiation in linear and 2D cell arrays using highly simplified equations that abstracted away the biochemistry of the system.
+In this paper, we present a simple yet biologically realistic model of Delta-Notch signalling and test it using three modelling formalisms: deterministic ODE, stochastic ODE, and agent-based.
 Using the Kolmogorov forward equations, we show that the agent-based formalisms converges to the system of deterministic ODEs in expectation.
@@
 
-\section{Introduction}
+\section*{Introduction}
 
-The Delta-Notch signalling network plays
+Cell differentiation is essential for the creation of structure in all complex organisms.
+Surprisingly, only a small number of signalling pathways are responsible for this essential functions \cite{bray_notch_2006}.
+One of these pathways is the Delta-Notch signalling network, which is unique because its ligands Delta and Jagged are transmembrane proteins.
+This means that Delta-Notch signalling is largely restricted to neighbouring cells. 
+However, new findings suggest that cellular protrusions can allow a small fraction of Delta-Notch signals to travel longer distances \cite{berkemeier_coupling_2023}.
+The Delta-Notch signalling network has been shown to affect neurogenesis in vertebrates by determining which cells adopt neural or glial fates. 
+Furthermore, mutations in Notch activity has been shown to be a contributing factor to many cancers \cite{bray_notch_2006}.
+
+\medskip
+
+% TODO: Insert a figure here. The paragraph below should be a caption.
 
@@
 
-\subsection{Model}
+Previous models of Delta-Notch signalling \cite{collier_pattern_1996, boareto_jaggeddelta_2015} have made a number of assumptions which we also use for our models.
+First, cells interact through Delta-Notch signalling if and only if they are adjacent.
+Production of Notch is an increasing hill function of Delta in neighbouring cells.
+Production of Delta is a decreasing hill function of Notch in the same cell.
+Both Notch and Delta decay at a rate proportional to their concentration.
+Low levels of Notch cause the primary fate, high levels cause the secondary fate.
+The behaviour   of cells with a specific fate depends on the specifics of the organism and organ, which is outside the scope of these models.
 
-Collier et al. make the following five modelling assumptions:
-\begin{enumerate}
-  \item Cells interact through Delta-Notch signalling if and only if they are adjacent.
-  \item Production of Notch is an increasing (hill) function of Delta in neighbouring cells.
-  \item Production of Delta is a decreasing (hill) function of Notch in the same cell.
-  \item Production of Notch and Delta is balanced by decay proportional to concentration.
-  \item Low levels of Notch cause the primary fate, high levels cause the secondary fate.
-\end{enumerate}
+\medskip
 
-Then they define the following variables:
+The model of Delta-Notch signalling proposed by Collier et al. \cite{collier_pattern_1996} makes other simplifications which violate our current understanding of the Notch signalling pathways.
+These simplifications make it difficult to identify how individual chemical reactions are captured by the model.
+For instance, production of activated Notch (NICD) is exclusively a function of the Delta concentration in neighbouring cells, with no regard for the concentration of Notch receptors.
+This violates the mass-balance equations, since the quantity of Notch receptors changes dynamically in response to the concentration of Delta ligands in nearby cells.
+There are smaller issues, like the absence of cis-inhibition and the Serrate (Jagged)
+protein, but these aren’t necessary to capture the fundamental behaviour of the
+system.
 
-\begin{itemize}
-  \item $\tau$: time
-  \item $N_{P}$: notch activity (concentration) in cell $P$
-  \item $D_{P}$: delta activity (concentration) in cell $P$
-  \item $\overline{D}_{P}$: average delta activity in neighbours of $P$
-  \item $N_{0}$: typical notch activity (across all cells)
-  \item $D_{0}$: typical delta activity (across all cells)
-  \item $\mu$: The decay rate for notch (assumed constant)
-  \item $\rho$: The decay rate for delta (assumed constant)
-\end{itemize}
+\medskip
 
-Then, define the following system of differential equations, where $F:[0, \infty) \rightarrow [0, \infty)$ is a continuous increasing function and $G: [0, \infty)\rightarrow [0, \infty)$ is a continuous decreasing function.
+In this paper, we develop three models of the Delta-Notch signalling network based on a modern understanding of the biochemistry of the system. 
+Our first model is an agent-based simulation of individual Delta and Notch molecules. 
+While this model is more accurate to the biology than an ODE model, it is impossible to analyze analytically and slow to test numerically.
+Then, we define two ODE models. 
+The deterministic ODE model defines the behaviour of the agent-based model in expectation (see the Appendix for a proof).
+The stochastic ODE model is an extension to the deterministic ODE model which contains noise terms that are intended to simulate the randomness of the agent-based model.
+We compare these models with each other using a combination of analytical and numerical methods in a variety of different domains and boundary conditions.
+Furthermore, we show that our models qualitatively replicate the results of Collier et al. \cite{collier_pattern_1996}. 
+In particular we show that a if linear array of cells is given no-flux boundary conditions, the pattern of alternating cell fates arises on the outside edges before gradually spreading inwards towards the center of the domain.
+Additionally, both our model and the Collier et al. model exhibit different ratios of primary to secondary fated cells depending on the parameters on a two-dimensional hexagonal lattice.
+
+\section*{Methods}
+
+From the literature \cite{collier_pattern_1996, boareto_jaggeddelta_2015}, we assume Notch ($N$) receptor production has the form $H^{+}(I) = n_{m}I^2/(n_{0}^2 + I^2)$ where $I$ is the NICD concentration and $n_{m}, n_{0}$ are hill function parameters.
+Similarly, Delta ligand ($D$) production is governed by a decreasing hill function of the for production has the form $H^{-}(I) = d_{m}d_{0}^2/(d_{0}^2 + I^2)$.
+Furthermore, we will asume that Notch and Delta both have decay rate $\gamma$.
+The variable $k_{T}$ will denote the binding rate between Notch receptors and Delta ligands.
+$\gamma_{I}$ is the decay rate of NICD.
+We will also assume that there is no cis-interaction and no Serrate ligand.
+
+\begin{table}[!htp]
+\centering
+\begin{tabular}{|m{8em}|m{5em}|m{20em}|} 
+ \hline
+ Reaction & Rate & Description \\ 
+ \hline
+ $N_{A} + D_{B} \rightarrow I_{A}$ & 
+ $k_{T} N_{A}D_{B}$ &
+ A Notch receptor from cell $A$ binds to a delta ligand from cell $B$, triggering the release of NICD in cell $A$. \\
+ \hline
+ $N_{A} \rightarrow \emptyset$ & 
+ $\gamma N_{A}$ & 
+ A Notch receptor from cell $A$ decays. \\
+ \hline
+ $D_{A} \rightarrow \emptyset$ & 
+ $\gamma D_{A}$ & 
+ A Delta receptor from cell $A$ decays. \\
+ \hline
+ $I_{A} \rightarrow \emptyset$ &
+ $\gamma_{I} I_{A}$ &
+ A NICD molecule from cell $A$ decays.  \\
+ \hline
+ $\emptyset \rightarrow N_{A}$ & 
+ $H^{+}(I_{A})$ &
+ A Notch receptor in cell $A$ is produced \\
+ \hline
+ $\emptyset \rightarrow D_{A}$ &
+ $H^{-}(I_{A})$ &
+ A Delta ligand in cell $A$ is produced \\
+ \hline
+\end{tabular}
+\caption{
+  A list of possible reactions and their rates in a cell $A$ with neighbour $B$.
+  Since there are $6$ possible reactions per cell, a system with $n$ cells will have $6n$ possible reactions to keep track of.
+}
+\label{tb:reactions}
+\end{table}
+
+\medskip
+
+We will make the assumption that reaction times follow a Poisson point process.
+Let $E$ be the set of all possible reaction events (i.e. the $6$ reactions above for all cells).
+Furthermore, let $|E| = N$ be the number of possible reactions.
+For each event, let $X_{i} \sim \text{Exp}(\lambda_{i})$ denote the time until the next event of type $i$, where $1 \leq i \leq N$.
+Then, the waiting time $T$ in between \emph{any} two events is:
 
 $$
-\begin{aligned}
-  \frac{d(N_{P} / N_{0})}{d\tau} = F(\overline{D}_{P} / D_{0}) - \mu N_{P} / N_{0} \\[5pt]
-  \frac{d(D_{P} / D_{0})}{d\tau} = G(N_{P} / N_{0}) - \rho D_{P} / D_{0}
-\end{aligned}
+T = \text{min} \{ X_{1}, \dots, X_{N} \} \sim \text{Exp}\left( \sum_{i = 1}^{N} \lambda_{i} \right) 
 $$
 
-This equation is inherently nondimensional, since $N_{P} / N_{0}$ and $D_{P} / D_{0}$ have no units.
+This results follows from the definition of an exponential distribution and the memoryless property.
+Then, we can use Gillespie's algorithm to simulate the system.
+This is the \emph{agent-based} model.
+Details of the computational implementation of this model can be found in Appendix \ref{sec:gillespie}.
 
-\subsection{Issues}
+\medskip
 
-The Collier et al.
-model makes numerous simplifications that violate our current understanding of the delta-notch signalling network.
-These simplifications make it difficult to derive a probabilistic model from the Collier et al.
-model, since the biological implications of such a model would be contradictory.
-Some of the issues include:
+In expectation, the agent-based model converges to the system of ODEs shown in Equation \ref{eq:deterministic}.
+A proof of this fact can be found in the appendix.
+This is the \emph{deterministic ODE} model, which is described below for a single cell in a two-cell system.
+However, this model can be extended to arbitrary domains by defining equations $dN_{i}/dt$, $dD_{i}/dt$, and $dI_{i}/dt$ for each $i \in \{ 1, \dots, n \}$, where $n$ is the number of cells.
+Then, simply replace $N_{ext}$ and $D_{ext}$ with the average Notch and Delta concentration over all neighbours.
 
-\begin{itemize}
-  \item Production of activated Notch (NICD) is exclusively a function of the Delta concentration in neighbouring cells, with no regard for the concentration of extracellular Notch receptors.
-  \item Furthermore, the concentration of extracellular Notch isn't accounted for at all, which means there is no way to create a meaningful reaction term that captures the dynamics of the system.
-  \item There are smaller issues, like the absence of cis-inhibition and the Serrate (Jagged) protein, but these aren't necessary to capture the fundamental behaviour of the system.
-\end{itemize}
+\begin{equation}
+\begin{aligned}
+  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\[5pt]
+  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\[5pt]
+  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
+\end{aligned}
+\label{eq:deterministic}
+\end{equation}
 
-\subsection{Qualitative Results}
+This model is a simplification of Boareto et al. \cite{boareto_jaggeddelta_2015}, which included an additonal equation for the Jagged molecule and kept track of cis-inhibition.
+To more directly compare our results with those of Collier et al. \cite{collier_pattern_1996}, we ignore Jagged interactions and cis-inhibition.
+Adding NICD, which was absent from Collier et al., was necessary for the assignment of probabilities to each reaction event.
+A bifurcation analysis of the system of ODEs in Equation \ref{eq:deterministic} reveals that ...
 
-In the two-cell periodic boundary condition case, Collier et al.
-show that there always exists a homogeneous steady state, although this state is unstable if the feedback in the Delta-Notch signalling network is sufficiently strong.
-If the line of cells is given Dirichlet boundary conditions, the inhomogeneity arises on the outside edges before gradually spreading inwards to the center of the domain.
-On the two-dimensional hexagonal lattice, the system exhibits different ratios of primary to secondary fated cells depending on the parameters.
-In particular, Collier et al.
-observed ratios of $1/2$, $1/3$, $1/4$, $1/5$, and $1/6$.
+\medskip
 
-\section{Methods}
+To capture the inherent randomness in receptor-ligand interactions while maintaining computational efficiency, we implement a \emph{stochastic ODE} model based on the chemical Langevin equation formalism.
+This approach represents an intermediate level of detail between the deterministic ODE model and the fully discrete Gillespie algorithm.
+Starting from the deterministic ODE model described in Equation \ref{eq:deterministic}, we incorporate stochastic fluctuations by adding noise terms proportional to the square root of the reaction rates to get Equation \ref{eq:stochastic}.
 
-\subsection{Parameters and Assumptions}
+\begin{equation}
+\begin{aligned}
+  dN &= \left( \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \right) dt + \sigma \cdot N \cdot dW_{N} \\[5pt]
+  dD &= \left( \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \right)  dt + \sigma \cdot D \cdot dW_{D}  \\[5pt]
+  dI &= \left( k_{T}ND_{ext} - \gamma_{I}I \right) dt + \sigma \cdot I \cdot dW_{I}
+\end{aligned}
+\label{eq:stochastic}
+\end{equation}
 
-Assume \emph{a priori} (i.e. from previous research) the following, where $I$ is NICD:
+where $dW_N$, $dW_D$, and $dW_I$ are independent Wiener processes representing white noise.
+The structure of the noise terms follows from the assumption that noise in each reaction channel is proportional to the current concentration of those molecules.
 
-\begin{itemize}
-  \item Basal extracellular Notch ($N$) production has the form $H^{+}(I) = n_{m}I^2/(n_{0}^2 + I^2)$. 
-  \item Basal extracellular Delta ($D$) production has the form $H^{-}(I) = d_{m}d_{0}^2/(d_{0}^2 + I^2)$.
-  \item Extracellular Notch and Delta have the same decay rate, $\gamma$.
-\end{itemize}
+\medskip
 
-Then, define the following parameters:
+To implement these models computationally, we employed the Euler-Maruyama scheme with time step $dt$.
+The deterministic step is first calculated, and then the Wiener increments are sampled from a normal distribution with mean 0 and standard deviation $\sqrt{dt}$.
+Each increment is scaled by the its respective variable and added to the results of the deterministic step.
 
-\begin{itemize}
-  \item $N_{0}$ and $D_{0}$ are the maximum rates of Notch/Delta production.
-  \item $k_{T}$ is the binding rate between extracellular Notch and Delta.
-  \item $N_{ext}$ and $D_{ext}$ refer to the average Notch/Delta levels in neighbouring cells.
-  \item $\gamma_{I}$ is the decay rate of NICD.
-\end{itemize}
+\section*{Results}
 
-Furthermore, we will assume that there is no cis-interaction and no Serrate ligand. This gives us the following set of chemical reactions that can occur in a single time step:
+% TODO: Write some filler text to connect the different experiments.
 
-\begin{enumerate}
-  \item A Notch receptor (from cell $A$) binds to a Delta ligand (from cell $B$), triggering the release of NICD in cell $A$.
-This can be written as $N_{A} + D_{B} \rightarrow I_{A}$.
-The probability of this event is $k_{T}N_{A}D_{B}$.
-  \item A Notch receptor in cell $A$ decays ($N_{A} \rightarrow \emptyset$) with probability $\gamma N_{A}$.
-  \item A Delta ligand in cell $A$ decays ($D_{A} \rightarrow \emptyset$) with probability $\gamma D_{A}$.
-  \item A NICD molecule in cell $A$ decays ($I_{A} \rightarrow \emptyset$) with probability $\gamma_{I}I_{A}$.
-  \item A Notch receptor in cell $A$ is produced ($\emptyset \rightarrow N_{A}$) with probability $H^{+}(I)$.
-  \item A Delta ligand in cell $A$ is produced ($\emptyset \rightarrow D_{A}$) with probability $H^{-}(I)$.
-\end{enumerate}
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/vis11.png}
+  \caption{A plot of the KDE approximation of the distribution of differentiation times for the agent-based model and the stochastic ODE model with different levels of noise.
+All simulations were done on a two-cell domain with periodic boundary conditions.
+These simulations show that higher values of the noise parameter significantly decrease the average time it takes for the two cells to adopt different fates.
+Notably, a noise parameter of $2\%$ causes the stochastic ODE model to align most closely with the results of agent-based simulations.
+}
+\end{figure}
 
-\subsection{Kolmogorov Forward Equations}
-\subsubsection{Derivation of Kolmogorov Forward Equation}
+\begin{figure}
+  \includegraphics[width=\textwidth]{img/vis12.png}
+  \caption{A \emph{Primary Outcome} occurs when the cell with the lower initial Notch concentration ends up adopting the primary (low Notch) fate.
+This plot shows the relationship between initial perturbation and the sample proportion of primary outcomes for the agent-based model on a two-cell domain.
+Prior to the perturbation, each cell has $200$ Notch molecules.
+A perturbation of $x$ molecules removes $x/2$ Notch molecules from the first cell and adds $x/2$ Notch molecules to the second.
+Unsurprisingly, larger initial perturbations result in a higher proportion of primary outcomes, although this effect is not as extreme as we expected.
+Even in the largest perturbation we tested ($\pm 25\%$), only $80\%$ of the simulations had the primary outcome.
+This suggests that it is difficult to reliably control cell fate by perturbation alone.}
+\end{figure}
+
+\begin{figure}[!ht]
+   \makebox[\textwidth][c]{\includegraphics[width=1.3\textwidth]{img/vis13.png}}
+   \caption{Results of two-at-a-time stability analysis of the deterministic, stochastic, and agent-based modelling formalisms on a two-cell domain with periodic boundary conditions.
+The regions enclosed by the red, blue, and green lines are regions in which cell differentiation occurs for each of the three models.
+Regions were obtained by simulating a $25 \times 25$ grid of test points and taking the convex hull of the points at which cell differentiation occurred.
+This analysis suggests that the deterministic and stochastic ODE models are on average less sensitive to changes in parameter values than the agent-based model.
+In particular, the agent-based model does not achieve cell differentiation for large values of the Notch and Delta production parameters $n_{m}$ and $d_{m}$.}
+\end{figure}
+
+\begin{table}[!htp]
+\centering
+\begin{tabular}{|c|c|c|c|} 
+ \hline
+ Pattern & Frequency (Odd Domain) & Frequency (Even Domain)  \\
+ \hline
+ $0: [], \, 1: []$ & $87$ & $403$ \\
+ $0: [], \, 1: [2]$ & $801$ & $123$ \\
+ $0: [], \, 1: [2, 2]$ & $27$ & $424$ \\
+ $0: [], \, 1: [2, 2, 2]$ & $48$ & $1$ \\
+ \hline
+\end{tabular}
+\caption{
+  Frequency of selected cell patterns in $1000$ simulations of the agent-based model on $9$-cell (even) and $10$-cell (odd) domains with zero-flux boundary conditions.
+A \emph{pattern} describes the contiguous groups of cells which share a fate.
+The pattern $0: [], \, 1: []$ means that cells have alternating fates.
+The pattern $0: [], \, 1: [2, 2]$ means there are two adjacent pairs of cells with the secondary fate $1$.
+We observe significant variations in patterns accross even and odd domains.
+Similarly drastic variation is observed accross different modelling formalisms and boundary conditions (not shown).
+}
+\label{tb:patterns}
+\end{table}
+
+% TODO: Add results on hexagonal domains.
+
+\section*{Discussion}
+
+% TODO: Write this section. Should only need 2-3 paragraphs.
+
+\nocite{*}
+\printbibliography
+
+\appendix
+
+\section{Gillespie's Algorithm} \label{sec:gillespie}
+
+Gillespie's algorithm allows us to simulate many types reactions simultaneously using random time steps.
+Recall that $T_{i}$ is a random variable that represents the distribution of the next reaction of any type.
+The random variables $X_{1}, \dots, X_{N}$ are the waiting time distributions for each reaction $1, \dots, N$, where $N = (6 \times \text{\# Cells})$.
+
+\medskip
+
+\textbf{Step 1}: Sample a value $t_{i}$ from $T$ using inverse transform sampling.
+
+\medskip
+
+\textbf{Step 2}: Increment the current time $t$ by $t_{i}$.
+
+\medskip
+
+\textbf{Step 3}: Determine which event took place. To do this, recall that:
+
+$$P(X_{k} = \text{min} \{  X_{1}, \dots, X_{N} \}) = \frac{\lambda_{k}}{\lambda_{1} + \dots + \lambda_{N}}$$
+
+Use this fact to partition the interval $[0, 1]$. Then sample $y_{i}$ from a $\text{Unif}(0, 1)$ distribution, identify its partition, and find the corresponding event.
+
+\medskip
+
+\textbf{Step 4}: Update the state based on the event determined in the previous step.
+
+\medskip
+
+\textbf{Remark}: The expected time for reaction $i$ to occur is given by $\mathbb{E}[X_{i}] = 1/\lambda_{i}$.
+Therefore, the rate at which event $i$ occurs is $\lambda_{i}$, which is equal to the rate at which the event occurs in the original ODE.
+This means that the trajectories generated from the Gillespie's algorithm implementation of the agent-based  model can be compared directly to the trajectories generated from the ODE itself, without any need for scaling.
+
+\section{Kolmogorov Forward Equations}
+\label{sec:kfe}
+
+% TODO: Revise this section to incorporate Eric's feedback.
+
+\subsection*{Derivation of Kolmogorov Forward Equation}
 
@@
 
-\subsubsection{Matrix Formulation}
+\subsection*{Matrix Formulation}
 The Kolmogorov forward equation can be expressed in matrix form as:
@@
 
-\subsection{Deterministic ODE Model}
+\section{Theoretical Justification}
+\label{sec:theoretical-justification}
 
-We implemented the following deterministic ODE model adapted from Boareto et al. (2015):
-
-$$
-\begin{aligned}
-  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\[5pt]
-  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\[5pt]
-  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
-\end{aligned}
-$$
-
-where $N$ is the number of intracellular Notch molecules, $D$ is the number of intracellular Delta molecules, and $I$ is the number of NICD molecules (see $3.1$ reference).
-This model is a simplification of the model found in Boareto et al.
-That model included an additonal equation for the Jagged molecule and kept track of cis-inhibition: where a Notch and Delta bind together inside the same cell, resulting in the degradation of both cells without contributing to the amount of NICD.
-So as to more directly compare our results with those of Collier et al., we ignore Jagged interactions and cis-inhibition.
-Including an NICD variable was necessary, however, as this additional interaction allowed for the assignment of probabilities to each reaction event and thus for the derivation of the Kolmogorov Equations and Gillespie algorithm.
-
-\subsection{Stochastic ODE Model}
-
-To capture the inherent randomness in cellular processes while maintaining computational efficiency, we implement a stochastic differential equation (SDE) model based on the chemical Langevin equation formalism.
-This approach represents an intermediate level of detail between the deterministic ODE model and the fully discrete Gillespie algorithm.
-
-Starting from the deterministic ODEs described in Section 3.3, we incorporate stochastic fluctuations by adding noise terms proportional to the square root of the reaction propensities:
-$$
-\begin{aligned}
-  dN &= \left( \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \right) dt + \sigma \cdot N \cdot dW_{N} \\ \\[5pt]
-  dD &= \left( \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \right)  dt + \sigma \cdot D \cdot dW_{D}  \\[5pt]
-  dI &= \left( k_{T}ND_{ext} - \gamma_{I}I \right) dt + \sigma \cdot I \cdot dW_{I}
-\end{aligned}
-$$
-where $dW_N$, $dW_D$, and $dW_I$ are independent Wiener processes representing white noise.
-The structure of the noise terms follows from the assumption that noise in each reaction channel is proportional to the current concentration of those molecules.
-\par
-
-For numerical implementation, we employ the Euler-Maruyama scheme with time step $dt$.
-The deterministic step is first calculated, and then the Wiener increments are sampled from a normal distribution with mean 0 and standard deviation $\sqrt{dt}$.
-Each increment is scaled by the its respective variable and added to the results of the deterministic step.
-
-\subsection{Gillespie Algorithm}
-
-We will make the assumption that reaction times follow a Poisson point process.
-Let $E$ be the set of all possible reaction events (i.e.
-reactions 1-6 above for all cells).
-Furthermore, let $|E| = N$ be the number of possible reactions.
-For each event, let $X_{i} \sim \text{Exp}(\lambda_{i})$ denote the time until the next event of type $i$, where $1 \leq i \leq N$.
-Then, the waiting time $T$ in between \emph{any} two events is:
-
-$$
-T = \text{min} \{ X_{1}, \dots, X_{N} \} \sim \text{Exp}\left( \sum_{i = 1}^{N} \lambda_{i} \right) 
-$$
-
-This results follows from the definition of an exponential distribution and the memoryless property. Then, we can use Gillespie's algorithm to simulate the system as follows:
-
-\medskip
-
-\textbf{Step 1}: Sample a value $t_{i}$ from $T$ using inverse transform sampling.
-
-\medskip
-
-\textbf{Step 2}: Increment the current time $t$ by $t_{i}$.
-
-\medskip
-
-\textbf{Step 3}: Determine which event took place. To do this, recall that:
-
-$$P(X_{k} = \text{min} \{  X_{1}, \dots, X_{N} \}) = \frac{\lambda_{k}}{\lambda_{1} + \dots + \lambda_{N}}$$
-
-Use this fact to partition the interval $[0, 1]$. Then sample $y_{i}$ from a $\text{Unif}(0, 1)$ distribution, identify its partition, and find the corresponding event.
-
-\medskip
-
-\textbf{Step 4}: Update the state based on the event determined in the previous step.
-
-\medskip
-
-\textbf{Remark}: The expected time for event $i$ to occur is given by $\mathbb{E}[X_{i}] = 1/\lambda_{i}$.
-Therefore, the rate at which event $i$ occurs is $\lambda_{i}$, which is equal to the rate at which the event occurs in the original ODE.
-This means that the trajectories generated from the Gillespie's algorithm implementation of the model can be compared directly to the trajectories generated from the ODE itself, without any need for scaling.
-
-\subsection{Theoretical Justification}
+% TODO: Revise this section to incorporate Eric's feedback.
 
@@
 
-\subsubsection{Setup and Notation}
+\subsection*{Setup and Notation}
 
@@
 \end{itemize}
-\subsubsection{Deriving ODEs for Expected Values}
+
+\subsection*{Deriving ODEs for Expected Values}
 
@@
 
-\subsubsection{Final System of ODEs}
+\subsection*{Final System of ODEs}
 
@@
 
-\subsubsection{Mapping Between Probabilities and Rates}
+\subsection*{Mapping Between Probabilities and Rates}
 
@@
 
-\subsection{Bifurcation Analysis}
+\section{Bifurcation Analysis}
+\label{sec:bifurcation}
+
+% TODO: Revise this section to incorporate Eric's feedback.
 
@@
 
-\section{Results}
-
-\subsection{Two-Cell Domain}
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{img/two_cell_deterministic.pdf}
-  \caption{Results of the two-cell deterministic ODE model.}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{img/two_cell_stochastic.pdf}
-  \caption{Results of the two-cell stochastic ODE model.}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{img/two_cell_gillespie.pdf}
-  \caption{Results of the two-cell Gillespie model.}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{img/two_cell_comparison.pdf}
-  \caption{Comparison of Notch concentrations in deterministic ODE, stochastic ODE, and Gillespie models.
-All models follow roughly the same trajectory and have similar differentiation times, although the Gillespie model appears to converge slightly faster.}
-\end{figure}
-
-\subsection{Linear Domain}
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{img/ODEModel_TwoCell_Heatmap.png}
-  \caption{Results of the linear domain deterministic ODE model. Heatmap shows Notch levels in each of 8 cells over 1000 time steps.}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{img/SDE_CellLine_HeatMap.png}
-  \caption{Results of the linear domain SDE model. Heatmap shows Notch levels in each of 8 cells over 1000 time steps.}
-\end{figure}
-
-\subsection{Hexagonal Domain}
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{img/ODE_HexGrid.png}
-  \caption{Results of the 8x8 hexagonal domain ODE model showing final Notch values per cell.}
-\end{figure}
-
-\section{Discussion}
-
 \end{flushleft}
 
-\nocite{*}
-\printbibliography
-
 \end{document}

### 2025-03-27 models/experiments.py

@@ -181,7 +181,7 @@ def ex11(runs = 500, noise = [0, 0.001, 0.002, 0.005, 0.01, 0.02, 0.05]):
 
 
 # Experiment 12: Effect of initial perturbation on cell fates in agent-based model
-def ex12(runs = 100, shift = np.linspace(0, 100, 6)):
+def ex12(runs = 250, shift = np.linspace(0, 100, 11)):
 
     # Run simulations of the agent-based model
     print("\nRunning Experiment 12\n")
@@ -195,7 +195,7 @@ def ex12(runs = 100, shift = np.linspace(0, 100, 6)):
         initial = np.array([DEFAULT_CELL + S, DEFAULT_CELL - S])
 
         # Run the simulation
-        vT, vS = gillespie.gillespie(domain, initial)
+        vT, vS = gillespie.gillespie(domain, initial, stop = 10 ** 6)
         vT, vS = np.array(vT), np.array(vS)
 
         # Append 0 to fates if the first cell had 0 Notch, 1 otherwise

### 2025-03-27 models/main.py

@@ -15,17 +15,14 @@ def main():
     # ex.ex05()
     # ex.ex06()
 
-    # Hexagonal Domain Tests
-    ex.ex15()
-
     # Two-Cell Experiments
     # ex.ex10()
     # ex.ex11()
-    # ex.ex12()
+    ex.ex12()
     # ex.ex13()
     # ex.ex14()
-    
-    
 
+    # Hexagonal Domain Tests
+    # ex.ex15()
 
 main()

### 2025-03-27 models/visualization.py

@@ -46,6 +46,7 @@ def vis_01_03(times, states, means, stds, name, title):
     # Save to the img/ foler
     plt.tight_layout()
     plt.savefig(f"img/{name}.pdf", dpi = 200)
+    plt.savefig(f"img/{name}.png", dpi = 300)
     plt.close(fig)
 
 
@@ -112,6 +113,7 @@ def vis_04_06(vT, vS, name, title):
 
     # Display the plot
     plt.savefig(f"img/{name}.pdf", dpi = 200)
+    plt.savefig(f"img/{name}.png", dpi = 300)
     plt.close()  
 
 
@@ -137,6 +139,7 @@ def vis10(shifts, dtimes):
     # Save to the img/ folder
     plt.tight_layout()
     plt.savefig(f"img/vis10.pdf", dpi = 200)
+    plt.savefig(f"img/vis10.png", dpi = 300)
     plt.close()
 
 
@@ -157,6 +160,7 @@ def vis11(noise, dtimes):
     # Save to the img/ folder
     plt.tight_layout()
     plt.savefig(f"img/vis11.pdf", dpi = 200)
+    plt.savefig(f"img/vis11.png", dpi = 300)
     plt.close()
 
 
@@ -171,6 +175,7 @@ def vis12(shift, fates):
     # Save to the img/ folder
     plt.tight_layout()
     plt.savefig(f"img/vis12.pdf", dpi = 200)
+    plt.savefig(f"img/vis12.png", dpi = 300)
     plt.close()
 
 
@@ -261,6 +266,7 @@ def vis13(grids, pairs):
     # Save to the img/ folder
     plt.tight_layout()
     plt.savefig(f"img/vis13.pdf", dpi = 200)
+    plt.savefig(f"img/vis13.png", dpi = 300)
     plt.close()
 
 #  Produces a visualization of the cell pattern for either hexagonal or linear domains

### 2025-03-29 paper/paper.tex
Fix bifurcation analysis using Eric's feedback.

@@
 \usepackage[left=1.5in, right=1.5in, top=1in, bottom=1in]{geometry}
+
+\usepackage{tikz}
 \usepackage{lscape}
@@
 However, this model can be extended to arbitrary domains by defining equations $dN_{i}/dt$, $dD_{i}/dt$, and $dI_{i}/dt$ for each $i \in \{ 1, \dots, n \}$, where $n$ is the number of cells.
-Then, simply replace $N_{ext}$ and $D_{ext}$ with the average Notch and Delta concentration over all neighbours.
+In a domain of arbitrary size, $N_{ext}$ and $D_{ext}$ denote the average Notch and Delta concentration over all neighbours.
 
@@
 Adding NICD, which was absent from Collier et al., was necessary for the assignment of probabilities to each reaction event.
-A bifurcation analysis of the system of ODEs in Equation \ref{eq:deterministic} reveals that ...
+An analysis of the system of ODEs in Equation \ref{eq:deterministic} for fixed values of $N_{ext}, D_{ext}$ reveals an $N = 0$ steady state (the \emph{primary fate}) for all parameter values.
+There is also a high Notch steady state (the \emph{secondary fate}) which emerges from a fold bifurcation when $D_{ext}$ is sufficiently large.
+See Appendix \ref{sec:bifurcation} for more information.
 
@@
 
-% TODO: Revise this section to incorporate Eric's feedback.
+\subsection*{One-Cell System}
 
-We perform a bifurcation analysis for the two-cell system, where $D_{ext}$ and $N_{ext}$ are treated as parameters.
-In other words, we will examine the steady states for one cell as a function of the Notch and Delta levels of the other cell.
-For simplicity, assume all parameters other than $D_{ext}, N_{ext}$ are fixed.
+First, we consider the one-cell system under the influence of a second, static cell.
+In this model, the Notch and Delta concentrations in the second cell ($N_{ext}$ and $D_{ext}$) are treated as parameters.
+Equations \ref{eq:one-cell-notch}, \ref{eq:one-cell-delta}, and \ref{eq:one-cell-nicd} define this system.
 
-$$
+\begin{align}
+  \label{eq:one-cell-notch}
+  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - (k_{T}D_{ext} + \gamma)N \\[5pt]
+  \label{eq:one-cell-delta}
+  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - (k_{T}N_{ext} + \gamma)D \\[5pt]
+  \label{eq:one-cell-nicd}
+  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
+\end{align}
+
+Observe that $dN/dt$ and $dI/dt$ do not depend on $D$.
+Therefore, we can analyze the $N-I$ phase plane to learn more about the system.
+First, we find the nullclines of Equations \ref{eq:one-cell-notch} and \ref{eq:one-cell-nicd} as functions of $I$.
+The results of this calculation are shown in Equations \ref{eq:one-cell-notch-nullcline} and \ref{eq:one-cell-nicd-nullcline}.
+
+\begin{align}
+  \label{eq:one-cell-notch-nullcline}
+  f_{N}(I) &= \frac{1}{\gamma + k_{T}D_{ext}}\left(\frac{n_{m}I^2}{n_{0}^2 + I^2}\right) \\[5pt]
+  \label{eq:one-cell-nicd-nullcline}
+  f_{I}(I) &= \frac{\gamma_{I}I}{k_{T}D_{ext}}
+\end{align}
+
+Plotting the nullclines in Equations \ref{eq:one-cell-notch-nullcline} and \ref{eq:one-cell-nicd-nullcline} on the $N-I$ plane for small and large values of $D_{ext}$ (assuming the other parameters are suitably chosen) reveals a fold bifurcation.
+
+\begin{figure}[!htp]
+  \label{fig:one-cell-phase-plane}
+  \centering
+  \begin{tikzpicture}[scale=1.75]
+    \node[draw] at (1.5,2.5) {Low $D_{ext}$};
+    \draw[->] (0, 0) -- (3, 0) node[right] {$I$};
+    \draw[->] (0, 0) -- (0, 2) node[above] {$N$};
+    \draw[domain= 0:2, smooth, variable=\x, blue] 
+      plot ({\x}, {\x});
+    \draw[domain= 0:3, smooth, variable=\x, red] 
+      plot ({\x}, {(2*\x*\x)/(2 + \x*\x)});
+    \draw (0, 0) node[circle,fill,scale=0.5] {};
+  \end{tikzpicture}
+  \begin{tikzpicture}[scale=1.75]
+    \node[draw] at (1.5,2.5) {High $D_{ext}$};
+    \draw[->] (0, 0) -- (3, 0) node[right] {$I$};
+    \draw[->] (0, 0) -- (0, 2) node[above] {$N$};
+    \draw[domain= 0:3, smooth, variable=\x, blue] 
+      plot ({\x}, {0.45 * \x});
+    \draw[domain= 0:3, smooth, variable=\x, red] 
+      plot ({\x}, {(1.5*\x*\x)/(2 + \x*\x)});
+    \draw (0, 0) node[circle,fill,scale=0.5] {};
+    \draw (0.78475, 0.35314) node[circle,draw=black,fill=white,thick,scale=0.5] {};
+    \draw (2.54858, 1.14686) node[circle,fill,scale=0.5] {};
+  \end{tikzpicture}
+  \caption{Approximate $N-I$ phase planes for the one-cell ODE model under the influence of low and high Delta from a static neighbour.
+The $N$-nullcline is shown in red and the $I$-nullcline is shown in blue.
+Below a critical value of $D_{ext}$, the system undergoes a fold bifurcation which results in the annihalation of the high Notch steady state.
+The high Notch steady state is also absent for qualitativley high values of $\gamma_{I}$, low values of $k_{T}$, and low values of $n_{m}$.}
+\end{figure}
+
+\subsection*{Two-Cell System}
+
+Now, we will consider the coupled two-cell system in which both cells respond to the Delta-Notch signalling of the other.
+Let $N_{1}, D_{1}, I_{1}$ and $N_{2}, D_{2}, I_{2}$ denote the Notch, Delta, and NICD concentrations in the first and second cells respectively.
+Our previous analysis reveals that the $N_{1} = I_{1} = 0$ steady state exists for all values of $D_{2}$.
+The steady state concentrations $\{ N_{1}^{*}, D_{1}^{*}, I_{1}^{*}, N_{2}^{*}, D_{2}^{*}, I_{2}^{*} \}$ for this case are given in Equation \ref{eq:two-cell-steady-states}.
+
+\begin{equation}
+\label{eq:two-cell-steady-states}
 \begin{aligned}
-  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - (k_{T}D_{ext} + \gamma)N \\[5pt]
-  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - (k_{T}N_{ext} + \gamma)D \\[5pt]
-  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I \\[5pt]
+  N_{1}^{*} &= 0  &
+  N_{2}^{*} &= \frac{1}{\gamma + k_{T}D_{1}^{*}}\left( \frac{n_{m}(I_{2}^{*})^2}{n_{0}^2 + (I_{2}^{*})^2} \right)  \\[5pt]
+  D_{1}^{*} &= \frac{d_{m}}{\gamma + k_{T}N_{2}^{*}} &
+  D_{2}^{*} &= \frac{1}{\gamma} \left( \frac{d_{m}d_{0}^2}{d_{0}^2 + (I_{2}^{*})^2} \right)  \\[5pt]
+  I_{1}^{*} &= 0 &
+  I_{2}^{*} &= \frac{k_{T}N_{2}^{*}D_{1}^{*}}{\gamma_{I}}
 \end{aligned}
-$$
+\end{equation}
 
-Solving for the nullclines gives us the following:
-
-$$
-\begin{aligned}
-  N &= \frac{1}{\gamma + k_{T}D_{ext}}\left(\frac{n_{m}I^2}{n_{0}^2 + I^2}\right) \\[5pt]
-  D &= \frac{1}{\gamma + k_{T}N_{ext}}\left( \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} \right) \\[5pt]
-  I &= \frac{k_{T}ND_{ext}}{\gamma_{I}}
-\end{aligned}
-$$
-
-Note that each nullcline defines a surface in $\mathbb{R}^3$.
-To express different cell fates, we need at least two stable steady states.
-We know that $I^{*} = k_{T}ND_{ext} / \gamma_{I}$ is the unique steady state value of $I$.
-Furthermore, by substituting $I^{*}$ into the $D$-nullcline, we get the unique steady-state value of $D$.
-There also exists a trivial stable steady state at $N = I = 0$.
-
-\medskip
-
-It remains to find the other steady-state values of $N$ as a function of $D_{ext}$.
-To do this, we use the computer algebra system sympy.
-Substituting the equation for $I^{*}$ into the $N$-nullcline gives us the following nonzero steady-state values of $N$:
-
-\medskip
-
-$$
-N^{*} = \frac{n_{m}}{2(D_{ext}k_{T} + \gamma)} \pm \sqrt{\frac{(D_{ext}k_{T}n_{m})^2 - 4(D_{ext}\gamma_{I}k_{T}n_{0} + \gamma \gamma_{I}n_{0})^2}{2D_{ext}k_{T}(D_{ext}k_{T} + \gamma)}}
-$$
-
-\textbf{Remark}: As mentioned above, we need two non-zero stable steady states for $N$ in order to represent the two different cell fates. This requires that the discriminant is positive:
-
-$$D_{ext}k_{T}n_{m} > 2D_{ext}\gamma_{I}k_{T}n_{0} + 2\gamma \gamma_{I} n_{0} \Rightarrow D_{ext}k_{T}(n_{m} - 2\gamma_{I}n_{0}) - 2\gamma \gamma_{I} n_{0} > 0$$
-
-As a general rule of thumb, increasing $n_{m}$ is a reliable way to recover these steady states.
-If we replace the $>$ sign with $=$ in the equation above, we find the location of a fold bifurcation.
-The lower branch corresponds to the $N = N_{ext}$, $D = D_{ext}$ unstable steady state.
-If the system is in this state, any perturbation will cause one cell to collapse to the high notch steady state and the other to collapse to the $N = 0$ steady state.
-Shown below is a rough sketch of the bifurcation diagram, including the corresponding cell fates.
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{img/bifurcation-diagram.png}
-  \caption{Approximate bifurcation diagram of the equilibrium Notch concentration of a single cell under the influence of fixed extracellular Delta from a neighbour.}
-\end{figure}
+There is a second steady state occurs when $N_{1}^{*} = N_{2}^{*}$, $D_{1}^{*} = D_{2}^{*}$, and $I_{1}^{*} = I_{2}^{*}$.
+While it is difficult to analytically determine the steady state concentrations, we can verify that this steady state exists and is unstable using computer simulations.
 

### 2025-04-07 paper/paper.tex
First draft of snacks presentation and edits to the paper.

@@
 
-The Delta-Notch signalling pathway plays a crucial role in determining the cell fate of initially homogeneous cells in a wide array biological systems.
-Early deterministic models of the Delta-Notch pathway achieved cell differentiation in linear and 2D cell arrays using highly simplified equations that abstracted away the biochemistry of the system.
-In this paper, we present a simple yet biologically realistic model of Delta-Notch signalling and test it using three modelling formalisms: deterministic ODE, stochastic ODE, and agent-based.
-Using the Kolmogorov forward equations, we show that the agent-based formalisms converges to the system of deterministic ODEs in expectation.
-We also demonstrate that the stochastic ODE and agent-based formalisms can produce cell differentiation even from identical initial conditions.
+The Delta-Notch signalling pathway plays a crucial role in determining the fate of initially homogeneous cells in a wide array of biological systems.
+In this paper, we present a simple yet biologically realistic representation of Delta-Notch signalling and test it using three models: ODE, SDE, and agent-based.
+Using the Kolmogorov forward equations, we show that the agent-based model converges to the ODE model in expectation.
+We also demonstrate that the SDE and agent-based models can produce cell differentiation even from identical initial conditions.
 Finally, we study the effect of parameters such as stochastic noise, domain size, and boundary conditions on the behaviour of our models.
-Our findings provide important insights into how different mathematical formalisms can affect simulations of the Delta-Notch signalling network with significant implications for future modelling work.
+Our findings provide insight into how modelling formalisms can affect simulations of the Delta-Notch signalling network with significant implications for future research.
 
@@
 
-Cell differentiation is essential for the creation of structure in all complex organisms.
-Surprisingly, only a small number of signalling pathways are responsible for this essential functions \cite{bray_notch_2006}.
-One of these pathways is the Delta-Notch signalling network, which is unique because its ligands Delta and Jagged are transmembrane proteins.
+Cell differentiation is essential for the formation of structure in complex organisms.
+But surprisingly, only a small number of signalling pathways are responsible for this important function \cite{bray_notch_2006}.
+One of these pathways is the Delta-Notch signalling network, which has been shown to affect neurogenesis in vertebrates by determining which cells adopt neural or glial fates. 
+Furthermore, mutations in Notch activity has been shown to be a contributing factor to many cancers \cite{bray_notch_2006}.
+The Delta-Notch signalling network is unique because its ligands Delta and Jagged are transmembrane proteins.
 This means that Delta-Notch signalling is largely restricted to neighbouring cells. 
 However, new findings suggest that cellular protrusions can allow a small fraction of Delta-Notch signals to travel longer distances \cite{berkemeier_coupling_2023}.
-The Delta-Notch signalling network has been shown to affect neurogenesis in vertebrates by determining which cells adopt neural or glial fates. 
-Furthermore, mutations in Notch activity has been shown to be a contributing factor to many cancers \cite{bray_notch_2006}.
 
@@
 First, cells interact through Delta-Notch signalling if and only if they are adjacent.
-Production of Notch is an increasing hill function of Delta in neighbouring cells.
-Production of Delta is a decreasing hill function of Notch in the same cell.
+Production of Notch is an increasing hill function of NICD.
+Production of Delta is a decreasing hill function of NICD.
 Both Notch and Delta decay at a rate proportional to their concentration.
-Low levels of Notch cause the primary fate, high levels cause the secondary fate.
-The behaviour   of cells with a specific fate depends on the specifics of the organism and organ, which is outside the scope of these models.
+Low levels of Notch cause the primary cell fate, while high levels cause the secondary cell fate.
+The behaviour of cells with a specific fate depends on the organism and organ, which is outside the scope of these models.
 
@@
 
-The model of Delta-Notch signalling proposed by Collier et al. \cite{collier_pattern_1996} makes other simplifications which violate our current understanding of the Notch signalling pathways.
+The model of Delta-Notch signalling proposed by Collier et al. \cite{collier_pattern_1996} makes other simplifications which violate the modern understanding of the signalling pathway.
 These simplifications make it difficult to identify how individual chemical reactions are captured by the model.
@@
 There are smaller issues, like the absence of cis-inhibition and the Serrate (Jagged)
-protein, but these aren’t necessary to capture the fundamental behaviour of the
+protein, but these omissions are not necessary to capture the behaviour of the
 system.
@@
 In this paper, we develop three models of the Delta-Notch signalling network based on a modern understanding of the biochemistry of the system. 
-Our first model is an agent-based simulation of individual Delta and Notch molecules. 
-While this model is more accurate to the biology than an ODE model, it is impossible to analyze analytically and slow to test numerically.
-Then, we define two ODE models. 
-The deterministic ODE model defines the behaviour of the agent-based model in expectation (see the Appendix for a proof).
-The stochastic ODE model is an extension to the deterministic ODE model which contains noise terms that are intended to simulate the randomness of the agent-based model.
-We compare these models with each other using a combination of analytical and numerical methods in a variety of different domains and boundary conditions.
+Our first model is an agent-based simulation of individual Delta, Notch, and NICD molecules. 
+While this model precisely simulates the reaction kinetics under our assumptions, it is impossible to analyze analytically and slow to simulate computationally.
+Then, we define two models based on differential equations. 
+The ordinary differential equation (ODE) model is equivalent the expected behaviour of the agent-based model (see Appendix \ref{sec:theoretical-justification} for a proof).
+The stochastic differential equation (SDE) model is an extension to the ODE model which contains additional noise terms.
+These terms are intended to simulate the stochasticity of the agent-based model.
+We compare these models with each other using a combination of analytical and computational methods in a variety of different domains.
 Furthermore, we show that our models qualitatively replicate the results of Collier et al. \cite{collier_pattern_1996}. 
-In particular we show that a if linear array of cells is given no-flux boundary conditions, the pattern of alternating cell fates arises on the outside edges before gradually spreading inwards towards the center of the domain.
+In particular we show that if a linear array of cells is given zero-flux boundary conditions, the pattern of alternating cell fates arises on the outside edges before gradually spreading inwards towards the center of the domain.
 Additionally, both our model and the Collier et al. model exhibit different ratios of primary to secondary fated cells depending on the parameters on a two-dimensional hexagonal lattice.
@@
 Similarly, Delta ligand ($D$) production is governed by a decreasing hill function of the for production has the form $H^{-}(I) = d_{m}d_{0}^2/(d_{0}^2 + I^2)$.
-Furthermore, we will asume that Notch and Delta both have decay rate $\gamma$.
+Furthermore, we will assume that Notch and Delta both have decay rate $\gamma$.
 The variable $k_{T}$ will denote the binding rate between Notch receptors and Delta ligands.
 $\gamma_{I}$ is the decay rate of NICD.
-We will also assume that there is no cis-interaction and no Serrate ligand.
+We will also assume that there is no cis-inhibition and no Serrate ligand.
 
@@
   A list of possible reactions and their rates in a cell $A$ with neighbour $B$.
-  Since there are $6$ possible reactions per cell, a system with $n$ cells will have $6n$ possible reactions to keep track of.
+  Since there are $6$ possible reactions per cell, a system with $n$ cells will have $6n$ possible reactions.
 }
@@
 
-We will make the assumption that reaction times follow a Poisson point process.
-Let $E$ be the set of all possible reaction events (i.e. the $6$ reactions above for all cells).
+We will assume that reaction times follow a Poisson point process.
+Suppose $E$ is the set of all possible reaction events (the $6$ reactions above for all cells).
 Furthermore, let $|E| = N$ be the number of possible reactions.
 For each event, let $X_{i} \sim \text{Exp}(\lambda_{i})$ denote the time until the next event of type $i$, where $1 \leq i \leq N$.
-Then, the waiting time $T$ in between \emph{any} two events is:
+Then, the waiting time $T$ in between any two events is:
 
@@
 This results follows from the definition of an exponential distribution and the memoryless property.
-Then, we can use Gillespie's algorithm to simulate the system.
+We can use Gillespie's algorithm to simulate the system.
 This is the \emph{agent-based} model.
-Details of the computational implementation of this model can be found in Appendix \ref{sec:gillespie}.
+Details of the computational implementation can be found in Appendix \ref{sec:gillespie}.
 
@@
 
-In expectation, the agent-based model converges to the system of ODEs shown in Equation \ref{eq:deterministic}.
+In expectation, the agent-based model converges to the system of ordinary differential equations shown in Equation \ref{eq:deterministic}.
 A proof of this fact can be found in the appendix.
-This is the \emph{deterministic ODE} model, which is described below for a single cell in a two-cell system.
+This is the \emph{ODE} model, which is described below for a single cell in a two-cell system.
 However, this model can be extended to arbitrary domains by defining equations $dN_{i}/dt$, $dD_{i}/dt$, and $dI_{i}/dt$ for each $i \in \{ 1, \dots, n \}$, where $n$ is the number of cells.
@@
 
-This model is a simplification of Boareto et al. \cite{boareto_jaggeddelta_2015}, which included an additonal equation for the Jagged molecule and kept track of cis-inhibition.
-To more directly compare our results with those of Collier et al. \cite{collier_pattern_1996}, we ignore Jagged interactions and cis-inhibition.
+This model is a simplification of a model by Boareto et al. \cite{boareto_jaggeddelta_2015}.
+However, to more directly compare our results with those of Collier et al. \cite{collier_pattern_1996}, we chose to ignore Serrate and cis-inhibition.
 Adding NICD, which was absent from Collier et al., was necessary for the assignment of probabilities to each reaction event.
@@
 
-To capture the inherent randomness in receptor-ligand interactions while maintaining computational efficiency, we implement a \emph{stochastic ODE} model based on the chemical Langevin equation formalism.
-This approach represents an intermediate level of detail between the deterministic ODE model and the fully discrete Gillespie algorithm.
-Starting from the deterministic ODE model described in Equation \ref{eq:deterministic}, we incorporate stochastic fluctuations by adding noise terms proportional to the square root of the reaction rates to get Equation \ref{eq:stochastic}.
+To capture the inherent randomness in receptor-ligand interactions while maintaining computational efficiency, we implement a stochastic differential equation \emph{SDE} model based on the Langevin equation formalism.
+This approach represents an intermediate level of detail between the ODE model and the fully discrete agent-based model.
+Starting from the ODE model described in Equation \ref{eq:deterministic}, we incorporate stochastic fluctuations by adding noise terms proportional to the square root of the reaction rates to get Equation \ref{eq:stochastic}.
 
@@
 
-where $dW_N$, $dW_D$, and $dW_I$ are independent Wiener processes representing white noise.
+In Equation \ref{eq:stochastic}, $dW_N$, $dW_D$, and $dW_I$ are independent Wiener processes representing white noise.
 The structure of the noise terms follows from the assumption that noise in each reaction channel is proportional to the current concentration of those molecules.
@@
 
-To implement these models computationally, we employed the Euler-Maruyama scheme with time step $dt$.
-The deterministic step is first calculated, and then the Wiener increments are sampled from a normal distribution with mean 0 and standard deviation $\sqrt{dt}$.
-Each increment is scaled by the its respective variable and added to the results of the deterministic step.
+To implement these models computationally, we employed the Euler-Maruyama method with time step $dt$.
+The deterministic increment calculated first, and then the Wiener increments are sampled from a normal distribution with mean $0$ and standard deviation $\sqrt{dt}$.
+Then, each Wiener increment is scaled by its respective variable and added to the deterministic increment.
 
@@
 
-\begin{figure}
-  \includegraphics[width=\textwidth]{img/vis11.png}
-  \caption{A plot of the KDE approximation of the distribution of differentiation times for the agent-based model and the stochastic ODE model with different levels of noise.
-All simulations were done on a two-cell domain with periodic boundary conditions.
-These simulations show that higher values of the noise parameter significantly decrease the average time it takes for the two cells to adopt different fates.
-Notably, a noise parameter of $2\%$ causes the stochastic ODE model to align most closely with the results of agent-based simulations.
-}
+\subsection*{Two-Cell Domains}
+
+Collier et al. \cite{collier_pattern_1996} observed cell differentiation in their ODE model under small initial perturbations on a two-cell domain.
+In our first experiment, we attempted to replicate this result under all three modelling formalisms.
+Since the SDE and agent-based models are stochastic, we did not apply an initial perturbation to them.
+For the ODE model, which is deterministic, we perturbed the initial Notch concentration by a factor of $X$, where $X$ is sampled from a $\mathcal{N}(1, 0.1)$ random variable.
+The results of this experiment are shown in Figure \ref{fig:two-cell-simulations}.
+
+\begin{figure}[!htp]
+  \includegraphics[width=\textwidth]{img/vis123.png}
+  \caption{Results of ten simulations of each modelling formalism (ODE, SDE, agent-based) on a two-cell domain with periodic boundary conditions.
+An initial perturbation was applied to the ODE model to move the system off of an unstable steady state.
+In all simulations, the system eventually converged to a differentiated state.
+The number of Notch molecules in the primary-fated and secondary-fated cells  are shown in green and yellow respectively.
+The bold line denotes the mean number of Notch molecules over all simulations.}
+  \label{fig:two-cell-simulations}
 \end{figure}
 
-\begin{figure}
+\medskip
+
+Using stochastic formalisms such as the SDE and agent-based models, we can investigate more nuanced questions about the behaviour of the Delta-Notch signalling network. 
+For our next experiment, we attempted to quantify how easy it would be force the two-cell system to adopt a certain outcome by perturbing the initial Notch concentration.
+We predicted that the cell with the lower initial number of Notch molecules would adopt the primary (low Notch) fate, while the other cell would adopt the secondary (high Notch) fate.
+While our prediction was generally true, it turns out the randomness of the agent-based model makes difficult to influence the outcome of a simulation using an initial perturbation.
+The results of this experiment are shown in Figure \ref{fig:outcome-forcing}.
+
+\begin{figure}[!htp]
   \includegraphics[width=\textwidth]{img/vis12.png}
@@
 This suggests that it is difficult to reliably control cell fate by perturbation alone.}
+  \label{fig:outcome-forcing}
 \end{figure}
 
-\begin{figure}[!ht]
+\medskip
+
+When we defined the SDE model in Equation \ref{eq:stochastic}, we introduced a parameter $\sigma$ that represents the amount of noise.
+However, with the methods we know, it is difficult to estimate the value of this parameter analytically or numerically.
+Here, we present a computational method to get a qualitative estimate of $\sigma$ which uses the differentiation time of the two-cell system.
+We define the differentiation time $t^{*}$ to be the time at which one cell adopts the primary fate ($N < 1$).
+Through our experiments, we noticed that higher levels of noise lower the average differentiation time by rapidly perturbing the system into more ``extreme'' states.
+Using this relationship, we estimated that $\sigma = 0.02$ by comparing simulations of the SDE model with the agent-based model.
+See Figure \ref{fig:noise-estimate} for a summary of this result.
+
+\begin{figure}[!htp]
+  \includegraphics[width=\textwidth]{img/vis11.png}
+  \caption{A plot of the KDE approximation of the distribution of differentiation times for the agent-based model and the stochastic ODE model with different levels of noise.
+All simulations were done on a two-cell domain with periodic boundary conditions.
+These simulations show that higher values of the noise parameter significantly decrease the average time it takes for the two cells to adopt different fates.
+Notably, a noise parameter of $2\%$ causes the stochastic ODE model to align most closely with the results of agent-based simulations.
+}
+  \label{fig:noise-estimate}
+\end{figure}
+
+\medskip
+
+All of the previous experiments were run using a fixed set of parameter values.
+However, in order to evaluate the quality of our model, we wanted to see if we could replicate the key result of cell differentiation under different parameter regimes.
+We identified five main parameters which effect whether the two-cell system differentiates.
+These parameters are the maximum rates of Notch and Delta production $n_{m}$ and $d_{m}$, the binding rate $k_{T}$, the Notch and Delta decay rate $\gamma$, and the NICD decay rate $\gamma_{I}$.
+Due to computational limitations, we could not perform a complete grid search of the $5$-dimensional parameter space.
+Instead, we opted to explore all $10$ of the $2$-dimensional subspaces of our parameter space.
+The results of our stability analysis are shown in Figure \ref{fig:stability-analysis}.
+
+\begin{figure}[!htp]
    \makebox[\textwidth][c]{\includegraphics[width=1.3\textwidth]{img/vis13.png}}
@@
 In particular, the agent-based model does not achieve cell differentiation for large values of the Notch and Delta production parameters $n_{m}$ and $d_{m}$.}
+   \label{fig:stability-analysis}
 \end{figure}
+
+\subsection*{Linear Domains}
+
+A linear domain is an array of cells in which each cell receives and emits Delta-Notch signals to $2$ neighbours ($1$ if the cell is on the boundary of domain).
+Collier et al. \cite{collier_pattern_1996} tested their model on linear domains of up to $40$ cells with zero-flux boundary conditions.
+They observed that cells on the boundary of the domain adopted the primary fate first.
+Then, the alternating pattern of primary and secondary fate cells gradually spread inwards towards the center of the domain.
+In our first experiment on a linear domain, we replicated this result as shown in Figure.
+
+% TODO: Add linear domain figure here.
+
+\medskip
+
+We also attempted to broadly classify the types of patterns we observed on linear domains.
+In order to do this, we first need to define the notion of a \emph{pattern}.
+Since the ``default'' or ``regular'' outcome of a simulation on a linear domain is an alternating array of cells with each fate, we will call this pattern $0: [],\, 1: []$.
+Then, each contiguous group of cells with the same fate (where $0$ is the primary fate and $1$ is the secondary fate) will add entries to the pattern arrays.
+For example, if there are $2$ contiguous pairs of cells with the primary fate and $1$ contiguous triplet of cells with the secondary fate, the simulation will have the pattern $0: [2, 2], \, 1: [3]$. 
+There is a lot more nuance related to pattern formation that we didn't have time to explore, but one interesting result is shown in Table \ref{tb:patterns}.
 
@@
   Frequency of selected cell patterns in $1000$ simulations of the agent-based model on $9$-cell (even) and $10$-cell (odd) domains with zero-flux boundary conditions.
-A \emph{pattern} describes the contiguous groups of cells which share a fate.
-The pattern $0: [], \, 1: []$ means that cells have alternating fates.
-The pattern $0: [], \, 1: [2, 2]$ means there are two adjacent pairs of cells with the secondary fate $1$.
-We observe significant variations in patterns accross even and odd domains.
-Similarly drastic variation is observed accross different modelling formalisms and boundary conditions (not shown).
+We observe significant variations in patterns across even and odd domains.
+Similarly drastic variation is observed across different modelling formalisms and boundary conditions (not shown).
 }
@@
 \end{table}
+
+\subsection*{Hexagonal Domains}
 
@@
 Plotting the nullclines in Equations \ref{eq:one-cell-notch-nullcline} and \ref{eq:one-cell-nicd-nullcline} on the $N-I$ plane for small and large values of $D_{ext}$ (assuming the other parameters are suitably chosen) reveals a fold bifurcation.
+This is shown in Figure \ref{fig:one-cell-phase-plane}.
 
@@
 The $N$-nullcline is shown in red and the $I$-nullcline is shown in blue.
-Below a critical value of $D_{ext}$, the system undergoes a fold bifurcation which results in the annihalation of the high Notch steady state.
-The high Notch steady state is also absent for qualitativley high values of $\gamma_{I}$, low values of $k_{T}$, and low values of $n_{m}$.}
+Below a critical value of $D_{ext}$, the system undergoes a fold bifurcation which results in the annihilation of the high Notch steady state.
+The high Notch steady state is also absent for qualitatively high values of $\gamma_{I}$, low values of $k_{T}$, and low values of $n_{m}$.}
 \end{figure}

### 2025-04-07 snacks-presentation/main.tex
First draft of snacks presentation and edits to the paper.

\documentclass[aspectratio=169]{beamer}

% BEAMER SETTINGS
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{enumerate items}[default]
\setbeamertemplate{footline}[frame number]

% CREATE SECTION SLIDES
\AtBeginSection[]{
  \begin{frame}
    \centering
    \Huge \insertsection
  \end{frame}
}

% PRESENTATION METADATA
\title{Comparing Models of Delta-Notch Signalling}
\author{Edwin Huras, Riley Wheadon, Ziyi Zhuang}
\institute{University of British Columbia}
\date{April 9, 2025}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

% SECTION 1: RILEY (10m)
\section{Introduction}

\begin{frame}
  \frametitle{Introduction}
  % Pictures of organisms/systems that use delta notch signalling
\end{frame}

\begin{frame}
  \frametitle{The Signalling Pathway}
  % Diagram of Delta-Notch signalling pathway

  \begin{columns}

    \begin{column}{0.48\textwidth}

    \end{column}

    \begin{column}{0.48\textwidth}
      \begin{itemize}
        \item Ligands bind to a Notch receptor.
        \item This releases NICD in the cell.
        \item NICD promotes Notch, inhibits Delta, and promotes Serrate.
        \item \emph{Cis-inhibition} occurs between molecules from the same cell.
      \end{itemize}
    \end{column}

  \end{columns}

\end{frame}

\begin{frame}
  \frametitle{Previous Models}
  % Two columns -- compare Collier et al. to Boareto et al.
  \begin{columns}

      \onslide<1->{\begin{column}{0.48\textwidth}
        \textbf{Collier et al. (1996)}

        \begin{itemize}
          \item Is a deterministic ODE model.
          \item Not accurate to the biochemistry.
          \item Uses computational simulations.
          \item Looks at multiple 1D/2D domains.
        \end{itemize}
    \end{column}}

      \onslide<2->{\begin{column}{0.48\textwidth}
        \textbf{Boareto et al. (2015)}

        \begin{itemize}
          \item Is a deterministic ODE model.
          \item Accurate to the biochemistry.
          \item Uses bifurcation theory.
          \item Only analyzes two-cell domains.
        \end{itemize}
    \end{column}}

  \end{columns}
  \vspace{2em}

  \onslide<3->{We investigate how the results of these models change under the \textbf{Stochastic Differential Equation} (SDE) and \textbf{Agent-Based} formalisms.}

\end{frame}

\begin{frame}
  \frametitle{Our Assumptions}
  \begin{itemize}
    \item All ligands are Delta molecules (so we ignore Serrate).
    \item Delta molecules only bind to Notch receptors from neighbouring cells. 
    \item Reaction events follow a Poisson point process.
    \item Notch and Delta production are hill functions of NICD.
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Our Chemical Equations}
  Consider a cell $A$ with neighbour $B$. The Notch, Delta, and NICD concentrations of cell $A$ are given by $N_{A}$, $D_{A}$, and $I_{A}$ respectively (and similarly for cell $B$).
  \begin{enumerate} 
    \item $N_{A} + D_{B} \rightarrow I_{A}$: Binding event. 
    \item $N_{A} \rightarrow \emptyset$: Notch decay.
    \item $D_{A} \rightarrow \emptyset$: Delta decay.
    \item $I_{A} \rightarrow \emptyset$: NICD decay.
    \item $\emptyset \rightarrow N_{A}$: Notch production.
    \item $\emptyset \rightarrow D_{A}$: Delta production.
  \end{enumerate}
\end{frame}

\begin{frame}

  \frametitle{Our Models}

  \onslide<1->{In the \textbf{Agent-Based} model, we simulate the chemical reactions directly using the \emph{Gillespie Algorithm}. Under our assumptions, this is the most accurate model.}

  \bigskip

  \onslide<2->{In the \textbf{ODE} model, we approximate the reaction kinetics by simulating the agent-based model \emph{in expectation}. To ensure cell differentiation, we apply an \emph{initial perturbation} to all simulations.}

  \bigskip

  \onslide<3->{In the \textbf{SDE} model, we reintroduce randomness by adding an independent \emph{Wiener Process} to each equation in the ODE model. This recreates some the stochasticity of the agent-based model.}

\end{frame}

% NOTE: I can extend this section by talking about the Gillespie algorithm.

% SECTION 2: DORA (15m?)
\section{Derivation}

\begin{frame}{Preliminaries and Assumptions}
\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{image0.png}
\end{figure}
    Table 1: A list of possible reactions and their rates in a cell $A$ with neighbour $B$.
\end{frame}

\begin{frame}{Preliminaries and Assumptions}
Since there are $6$ possible reactions per cell, a system with $k$ cells will have $6k$ possible reactions to keep track of. \\
\hfill\break
In the following derivations, we focus on tracking a single cell (cell $A$) for simplicity.
\end{frame}

\begin{frame}{Derivation of Kolmogorov Forward Equation}
Let us define
\begin{itemize}
    \small \item $P(n,d,i,t)$: the probability of having $n$ Notch receptors, $d$ Delta ligands, and $i$ NICD molecules at time $t$.
    \small \item $D_{ext}$: the average Delta concentration over all neighbours of cell $A$.
\end{itemize}
The KFE describes how $P(n,d,i,t)$ changes over a small time step $\Delta t$:
\small \begin{align*}
P(n,d,i,t+\Delta t) &= P(n,d,i,t) \\
&+ P(n-1,d,i,t) \cdot H^+(i) \Delta t \quad \text{(Notch production)} \\
&+ P(n,d-1,i,t) \cdot H^-(i) \Delta t \quad \text{(Delta production)} \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1)\Delta t \quad \text{(Notch decay)} \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1)\Delta t \quad \text{(Delta decay)} \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1)\Delta t \quad \text{(NICD decay)} \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext}\Delta t \quad \text{(Trans-activation)} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]\Delta t
\end{align*}
\end{frame}

\begin{frame}{Derivation of Kolmogorov Forward Equation}
Rearranging to find the time derivative and taking the limit as $\Delta t \rightarrow 0$:

\begin{align*}
\frac{\partial P(n,d,i,t)}{\partial t} &= P(n-1,d,i,t) \cdot H^+(i) \\
&+ P(n,d-1,i,t) \cdot H^-(i) \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
\end{align*}

\end{frame}

\begin{frame}{Matrix Formulation}
The Kolmogorov forward equation can be expressed in matrix form as:
\[
\frac{d\vec{P}(t)}{dt} = \mathbf{A} \vec{P}(t)
\]

where $\vec{P}(t)$ is a column vector containing the probabilities for all possible states, and $\mathbf{A}$ is the transition rate matrix.\\
Each entry $A_{i,j}$ represents the transition rate from state $i$ to state $j$. The diagonal elements $A_{i,i}$ contain the negative sum of all outgoing rates from state $i$.

\end{frame}

\begin{frame}{Matrix Formulation}
Each state in our system is described by ($n$,$d$,$i$), but the matrix form needs a single index.
\vfill
\pause
We flatten the 3D state space into a 1D list.\\
\begin{itemize}
    \item We encode ($n$,$d$,$i$) into a single index — just like how multidimensional data (e.g., images, simulation grids) are stored in computer memory.
\end{itemize}
\end{frame}

\begin{frame}{Matrix Formulation}
To construct the matrix $\mathbf{A}$, we first establish a one-to-one mapping between the three-dimensional state space $(n,d,i)$ and a one-dimensional index $k$. \\~\\
We set $n_{max}$, $d_{max}$, and $i_{max}$ as large enough bounds for Notch, Delta, and NICD in cell $A$. These define the size of our 3D ``box" of possible states. \\
\textit{Remark:} These are not strict limits but chosen large enough to technically capture all relevant dynamics.
\end{frame}

\begin{frame}{Matrix Formulation}
For a system with ``maximum" values $n_{max}$, $d_{max}$, and $i_{max}$, we define:
\[
k(n,d,i) = n \cdot (d_{max}+1) \cdot (i_{max}+1) + d \cdot (i_{max}+1) + i
\]
for $0 \leq n \leq n_{max}, \; 0 \leq d \leq d_{max}, \; 0 \leq i \leq i_{max}$. \\
The inverse mapping gives:
\begin{align*}
n(k) &= \left\lfloor \frac{k}{(d_{max}+1) \cdot (i_{max}+1)} \right\rfloor \\
d(k) &= \left\lfloor \frac{k \bmod ((d_{max}+1) \cdot (i_{max}+1))}{i_{max}+1} \right\rfloor \\
i(k) &= k \bmod (i_{max}+1)
\end{align*}
\end{frame}

\begin{frame}{Matrix Formulation}
With this mapping, we can now define the elements of matrix $\mathbf{A}$. Let $k$ and $k'$ be the indices corresponding to states $(n,d,i)$ and $(n',d',i')$ respectively:

\begin{align*}
A_{k,k'} = 
\begin{cases}
H^+(i) & \text{if } (n',d',i') = (n+1,d,i) \quad \text{(Notch production)} \\
H^-(i) & \text{if } (n',d',i') = (n,d+1,i) \quad \text{(Delta production)} \\
\gamma(n) & \text{if } (n',d',i') = (n-1,d,i) \quad \text{(Notch decay)} \\
\gamma(d) & \text{if } (n',d',i') = (n,d-1,i) \quad \text{(Delta decay)} \\
\gamma_I(i) & \text{if } (n',d',i') = (n,d,i-1) \quad \text{(NICD decay)} \\
k_T(n)D_{ext} & \text{if } (n',d',i') = (n-1,d,i+1) \quad \text{(Trans-activation)} \\
-v & \text{if } k' = k \quad \text{(Diagonal terms)}
\end{cases}
\end{align*}
\end{frame}

\begin{frame}{Matrix Formulation}
Note that $A_{k,k'}$ represents the transition rate from state $k$ to state $k'$. The diagonal elements $A_{k,k}$ are the negative sum of all outgoing rates from state $k$:

\[
A_{k,k} = -v=-[H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}]
\]

where $(n,d,i)$ is the state corresponding to index $k$.\\

The resulting matrix $\mathbf{A}$ is a sparse matrix with the following properties:
\begin{itemize}
    \item Dimension: $(n_{max}+1) \cdot (d_{max}+1) \cdot (i_{max}+1) \times (n_{max}+1) \cdot (d_{max}+1) \cdot (i_{max}+1)$
    \item Number of non-zero elements: $\approx 7 \cdot (n_{max}+1) \cdot (d_{max}+1) \cdot (i_{max}+1)$
\end{itemize}
\end{frame}

\begin{frame}{Matrix Formulation}
The corresponding transition matrix $\mathbf{A}$ would be:

\begin{align*}
\mathbf{A} = 
\begin{pmatrix}
-v_{000} & \alpha_{000,001} & \alpha_{000,010} & \alpha_{000,011} & \alpha_{000,100} & \alpha_{000,101} & \alpha_{000,110} & \alpha_{000,111} \\
\alpha_{001,000} & -v_{001} & \alpha_{001,010} & \alpha_{001,011} & \alpha_{001,100} & \alpha_{001,101} & \alpha_{001,110} & \alpha_{001,111} \\
\alpha_{010,000} & \alpha_{010,001} & -v_{010} & \alpha_{010,011} & \alpha_{010,100} & \alpha_{010,101} & \alpha_{010,110} & \alpha_{010,111} \\
\alpha_{011,000} & \alpha_{011,001} & \alpha_{011,010} & -v_{011} & \alpha_{011,100} & \alpha_{011,101} & \alpha_{011,110} & \alpha_{011,111} \\
\alpha_{100,000} & \alpha_{100,001} & \alpha_{100,010} & \alpha_{100,011} & -v_{100} & \alpha_{100,101} & \alpha_{100,110} & \alpha_{100,111} \\
\alpha_{101,000} & \alpha_{101,001} & \alpha_{101,010} & \alpha_{101,011} & \alpha_{101,100} & -v_{101} & \alpha_{101,110} & \alpha_{101,111} \\
\alpha_{110,000} & 0 & H^+(0) & 0 & H^-(1) & 0 & -v_{110} & \gamma_I \\
\alpha_{111,000} & 0 & 0 & H^+(0) & 0 & H^-(1) & H^+(1) & -v_{111}
\end{pmatrix}
\end{align*}

where:
\[
\alpha_{ndi,n'd'i'} = H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}
\]

\end{frame}

\begin{frame}{Remarks}
Let us consider

\begin{itemize}
    \item A one-cell system: $P(n,d,i,t+\Delta t)$ \\
    A $k$-cell system: $P(n_1,d_1,i_1,n_2,d_2,i_2,...,n_k,d_k,i_k,t+\Delta t)$\\
    \item $D_{ext}$ (the average Delta concentration over all neighbours of cell $A$).
\end{itemize}
\vfill
\pause
The complexity of a single cell highlights the need for numerical methods to model interactions between cells.
\end{frame}

% SECTION 3: EDWIN (20m?)
\section{Results}

\begin{frame}
  \frametitle{Comparing Two-Cell Models}
\end{frame}

\begin{frame}
  \frametitle{Noise in the SDE Model}
\end{frame}

\begin{frame}
  \frametitle{Stability Analysis}

  \onslide<1->{
    We tested model stability by perturbing $5$ parameters: $N_{m}$, $D_{m}$, $K_{T}$, $\gamma$, and $\gamma_{I}$.
    \begin{itemize}
      \item $N_{m}$, $D_{m}$ are the maximum rates of Notch/Delta production.
      \item $K_{T}$ is the binding rate.
      \item $\gamma$, $\gamma_{I}$ are the Notch/Delta and NICD decay rates.
    \end{itemize}
  }

  \bigskip

  \onslide<2->{
    Due to computational limitations, we investigated the $2$-dimensional subspaces of our $5$-dimensional parameter space using the following algorithm:
    \begin{itemize}
      \item Run the two-cell model on a $25 \times 25$ grid of points spanning $2$ OOMs.
      \item Take the convex hull of the points at which the cells differentiated.
    \end{itemize}
}

\end{frame}

\begin{frame}
  \frametitle{Stability Analysis}
\end{frame}

\begin{frame}
  \frametitle{Simulations on Linear Domains}
\end{frame}

\begin{frame}
  \frametitle{Simulations on Linear Domains}
\end{frame}

\begin{frame}
  \frametitle{Patterns on Linear Domains}
\end{frame}

\begin{frame}
  \frametitle{Simulations on Hexagonal Domains}
\end{frame}

\begin{frame}
  \frametitle{Patterns on Hexagonal Domains}
\end{frame}

\begin{frame}
  \frametitle{Patterns on Hexagonal Domains}
\end{frame}

\begin{frame}
  \frametitle{Patterns on Hexagonal Domains}
\end{frame}

\begin{frame}
  \frametitle{Patterns on Hexagonal Domains}
\end{frame}

\begin{frame}
  \frametitle{Patterns on Hexagonal Domains}
\end{frame}

\end{document}

### 2025-04-07 models/experiments.py

@@ -34,6 +34,7 @@ def ex01(runs = 10, pb = 0.05):
     # Compute summary statistics, plot the results
     means, stds = summarize(states)
     vis.vis01(times, states, means, stds)
+    return times, states, means
     print("\nExperiment Completed!\n")
 
 
@@ -55,6 +56,7 @@ def ex02(runs = 10, noise = 0.02):
     # Compute summary statistics, plot the results
     means, stds = summarize(states)
     vis.vis02(times, states, means, stds)
+    return times, states, means
     print("\nExperiment Completed!\n")
 
 
@@ -78,7 +80,16 @@ def ex03(runs = 10):
     # Compute summary statistics, plot the results
     means, stds = summarize(states)
     vis.vis03(times, states, means, stds)
+    return times, states, means
     print("\nExperiment Completed!\n")
+
+
+# Experiment 123: Comparison of two-cell models
+def ex123(runs = 10):
+    ode_data = ex01(runs = runs)
+    sde_data = ex02(runs = runs)
+    abm_data = ex03(runs = runs)
+    vis.vis123(ode_data, sde_data, abm_data)
 
 
 # Experiment 04: Deterministic ODE model, linear domain
@@ -403,18 +414,21 @@ def ex14(file = "results/ex14.txt"):
 
 
 # Experiment 15: Hexagonal simulations with different boundary conditions
-def ex15(nx = 11, ny = 11):
+def ex15(g = 0.02, nx = 11, ny = 11):
 
     print("\nRunning Experiment 15\n")
 
+    # Set the parameters based on g
+    params = DEFAULT
+
     # Run the stochastic ODE model with periodic BCs
     domain = domains.hexagonal(nx, ny, "periodic")
-    vT, vS = ode.ode(domain, DEFAULT_CELL, noise = 0.02)
+    vT, vS = ode.ode(domain, DEFAULT_CELL, params = params, noise = 0.02)
     periodic_pattern = pattern(vS)
 
     # Run the stochastic ODE model with dirichlet BCs
     domain = domains.hexagonal(nx, ny, "dirichlet")
-    vT, vS = ode.ode(domain, DEFAULT_CELL, noise = 0.02)
+    vT, vS = ode.ode(domain, DEFAULT_CELL, params = params, noise = 0.02)
     dirichlet_pattern = pattern(vS)
     
     # Use the updated pattern_hexagonal function

### 2025-04-07 models/main.py

@@ -9,6 +9,7 @@ def main():
     # ex.ex01()
     # ex.ex02()
     # ex.ex03()
+    # ex.ex123()
 
     # Linear Domain Tests
     # ex.ex04()
@@ -19,10 +20,10 @@ def main():
     # ex.ex10()
     # ex.ex11()
     # ex.ex12()
-    # ex.ex13()
+    ex.ex13()
     # ex.ex14()
 
     # Hexagonal Domain Tests
-    ex.ex15()
+    # ex.ex15()
 
 main()

### 2025-04-07 models/visualization.py

@@ -60,6 +60,47 @@ def vis03(times, states, means, stds):
     vis_01_03(times, states, means, stds, "vis03", "Agent-Based Model")
 
 
+# Visualization 123 (comparison of two-cell models)
+def vis123(ode_data, sde_data, abm_data):
+
+    # Set default line width and font size
+    mpl.rcParams['lines.linewidth'] = 2
+    mpl.rcParams.update({'font.size': 20}) 
+
+    # Iterate through axes (0: ODE, 1: SDE, 2: agent-based)
+    fig, axs = plt.subplots(nrows = 3, sharex = True, figsize = (16, 9))
+    for i, data in enumerate([ode_data, sde_data, abm_data]):
+
+        # Unpack the data for the current model
+        times, states, means = data
+
+        # Plot the trajectories for each sample
+        for j in range(len(states)):
+            axs[i].plot(times, states[j, :, 0, 0], color = "green", alpha = 0.2)
+            axs[i].plot(times, states[j, :, 1, 0], color = "orange", alpha = 0.2)
+
+        # Plot the means at full opacity
+        axs[i].plot(times, means[:, 0, 0], color = "darkgreen")
+        axs[i].plot(times, means[:, 1, 0], color = "darkorange")
+        axs[i].set_xlim(0, 1500)
+
+    # Set axis labels
+    axs[0].set_ylabel("ODE")
+    axs[1].set_ylabel("SDE")
+    axs[2].set_ylabel("Agent-Based")
+
+    # Add title and labels to the graph
+    fig.supxlabel("Time")
+    fig.supylabel("Notch Molecules")
+    fig.suptitle("Comparison of Models on Two-Cell Domain", fontsize = 24)
+
+    # Save to the img/ foler
+    plt.tight_layout()
+    plt.savefig(f"img/vis123.pdf", dpi = 300)
+    plt.savefig(f"img/vis123.png", dpi = 300)
+    plt.close(fig)
+
+
 # Visualization 04-06 helper function:
 def vis_04_06(vT, vS, name, title):
 
@@ -213,15 +254,27 @@ def stability_visualization(ax, grids, pair):
     ax.add_patch(s_poly)
     ax.add_patch(g_poly)
    
[…]

### 2025-04-08 snacks-presentation/main.tex
Add a bunch of .jpg slides.

@@
 \section{Derivation}
-
 \begin{frame}{Preliminaries and Assumptions}
@@
 To construct the matrix $\mathbf{A}$, we first establish a one-to-one mapping between the three-dimensional state space $(n,d,i)$ and a one-dimensional index $k$. \\~\\
+
 We set $n_{max}$, $d_{max}$, and $i_{max}$ as large enough bounds for Notch, Delta, and NICD in cell $A$. These define the size of our 3D ``box" of possible states. \\
@@
 \[
-A_{k,k} = -v=-[H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}]
+A_{k,k} = -v_k =-[H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}]
 \]
@@
 where $(n,d,i)$ is the state corresponding to index $k$.\\
+\textit{Recall:} We assume that reaction times follow a Poisson point process.\\~\\
+
+\pause
 
@@
 
-\begin{align*}
+\small \begin{align*}
 \mathbf{A} = 
@@
 \alpha_{101,000} & \alpha_{101,001} & \alpha_{101,010} & \alpha_{101,011} & \alpha_{101,100} & -v_{101} & \alpha_{101,110} & \alpha_{101,111} \\
-\alpha_{110,000} & 0 & H^+(0) & 0 & H^-(1) & 0 & -v_{110} & \gamma_I \\
-\alpha_{111,000} & 0 & 0 & H^+(0) & 0 & H^-(1) & H^+(1) & -v_{111}
+\alpha_{110,000} & \alpha_{110,001} & \alpha_{110,010} & \alpha_{110,011} & \alpha_{110,100} & \alpha_{110,101} & -v_{110} & \alpha_{110,111} \\
+\alpha_{111,000} & \alpha_{111,001} & \alpha_{111,010} & \alpha_{111,011} & \alpha_{111,100} & \alpha_{111,101} & \alpha_{111,110} & -v_{111}
 \end{pmatrix}
@@
 where:
-\[
-\alpha_{ndi,n'd'i'} = H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}
-\]
+\begin{align*}
+\alpha_{ndi,n'd'i'} = 
+\begin{cases}
+H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext} & \text{if } (n,d,i) \rightarrow (n',d',i') \\
+0 & \text{otherwise}
+\end{cases}
+\end{align*}
 
@@
     A $k$-cell system: $P(n_1,d_1,i_1,n_2,d_2,i_2,...,n_k,d_k,i_k,t+\Delta t)$\\
-    \item $D_{ext}$ (the average Delta concentration over all neighbours of cell $A$).
+    \item $D_{ext}$ (the average Delta concentration over all neighbours of cell $A$)\\ can vary depending on the choice of domain.
 \end{itemize}

### 2025-04-08 models/experiments.py

@@ -434,16 +434,3 @@ def ex15(g = 0.02, nx = 11, ny = 11):
     # Use the updated pattern_hexagonal function
     vis.vis15(periodic_pattern, dirichlet_pattern, nx, ny)
     print("\nExperiment Completed!\n")
-
-
-# Experiment 16: Hexagonal simulation with animation
-def ex16(nx = 7, ny = 7):
-    
-    print("\nRunning Experiment 16\n")
-
-    # Run the stochastic ODE model with dirichlet BCs
-    domain = domains.hexagonal(nx, ny, "dirichlet")
-    vT, vS = ode.ode(domain, DEFAULT_CELL, noise = 0.02)
-
-    vis.animation_hexagonal(vT, vS, nx, ny, animation_title=None)
-    print("\nExperiment Completed!\n")

### 2025-04-11 paper/paper.tex
Point to new vis11 figure paper.tex
paper.tex
Add a note to Table 1 description; Update KFE; Update Theoretical Justification

@@
   Since there are $6$ possible reactions per cell, a system with $n$ cells will have $6n$ possible reactions.
+  \textbf{Note:} In the Reaction column, $N_A$, $D_B$, and $I_A$ represent one Notch, Delta, and NICD, respectively.
+In the Rate column, $N_A$, $D_B$, and $I_A$ refer to the number of Notch molecules in cell A, Delta in cell B, and NICD in cell A, respectively.
 }
@@
 \begin{figure}[!htp]
-  \includegraphics[width=\textwidth]{img/vis11.png}
+  \includegraphics[width=\textwidth]{img/vis11_V2.png}
   \caption{A plot of the KDE approximation of the distribution of differentiation times for the agent-based model and the stochastic ODE model with different levels of noise.
@@
 
-% TODO: Revise this section to incorporate Eric's feedback.
+\subsection*{Derivation of Kolmogorov Forward Equation}
+For simplicity, we consider the dynamics of a single cell (cell $A$) in the following derivations.
 
-\subsection*{Derivation of Kolmogorov Forward Equation}
+Let us define:
+\begin{itemize}
+    \item $P(n,d,i,t)$: the probability of having $n$ Notch receptors, $d$ Delta ligands, and $i$ NICD molecules in cell $A$ at time $t$.
+    \item $D_{ext}$: the average number of Delta molecules in the neighbourhood of cell $A$.
+\end{itemize}
 
-Let us define $P(n,d,i,t)$ to be the probability of having $n$ Notch receptors, $d$ Delta ligands, and $i$ NICD molecules at time $t$.
 The Kolmogorov forward equation describes how $P(n,d,i,t)$ changes over a small time step $\Delta t$:
-
 \begin{align*}
@@
 
-Rearranging to find the time derivative:
-
-\begin{align*}
-\frac{P(n,d,i,t+\Delta t) - P(n,d,i,t)}{\Delta t} &= P(n-1,d,i,t) \cdot H^+(i) \\
-&+ P(n,d-1,i,t) \cdot H^-(i) \\
-&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
-&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
-&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
-&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
-&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
-\end{align*}
-
-Taking the limit as $\Delta t \rightarrow 0$:
-
+Rearranging to find the time derivative and taking the limit as $\Delta t \rightarrow 0$:
 \begin{align*}
@@
 \[
-\frac{d\vec{P}(t)}{dt} = \mathbf{A} \vec{P}(t)
+\frac{d\vec{P}(t)}{dt} = \vec{P}(t)\mathbf{A}
 \]
-
-where $\vec{P}(t)$ is a column vector containing the probabilities for all possible states, and $\mathbf{A}$ is the transition rate matrix. \\
+where $\vec{P}(t)$ is a row vector containing the probabilities for all possible states, and $\mathbf{A}$ is the transition rate matrix with $A_{i,j}$ giving the transition rate from state $i$ to $j$.
+The diagonal elements $A_{i,i}$ contain the negative sum of all outgoing rates from state $i$.
 
 To construct the matrix $\mathbf{A}$, we first establish a one-to-one mapping between the three-dimensional state space $(n,d,i)$ and a one-dimensional index $k$.
-For a system with maximum values $n_{max}$, $d_{max}$, and $i_{max}$, we define:
+We set $n_{max}$, $d_{max}$, and $i_{max}$ as sufficiently large upper bounds for the number of Notch, Delta, and NICD molecules in cell $A$, respectively.
+These define the size of our 3D ``box" of possible states.
+
+\textbf{Remark}: These are not strict limits but chosen large enough to technically capture all relevant dynamics.
+
+For a system with sufficiently large values $n_{max}$, $d_{max}$, and $i_{max}$, we now define:
+
 \[
@@
 \begin{align*}
-A_{k',k} = 
+A_{k,k'} = 
 \begin{cases}
@@
 k_T(n)D_{ext} & \text{if } (n',d',i') = (n-1,d,i+1) \quad \text{(Trans-activation)} \\
--\sum_{k'' \neq k} A_{k'',k} & \text{if } k' = k \quad \text{(Diagonal terms)}
+-v_k & \text{if } k' = k \quad \text{(Diagonal terms)}
 \end{cases}
@@
 
-Note that $A_{k',k}$ represents the transition rate from state $k$ to state $k'$. The diagonal elements $A_{k,k}$ are the negative sum of all outgoing rates from state $k$:
+Note that $A_{k,k'}$ represents the transition rate from state $k$ to state $k'$. The diagonal elements $A_{k,k}$ are the negative sum of all outgoing rates from state $k$:
 
 \[
-A_{k,k} = -[H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}]
+A_{k,k} = -v_k =-[H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
 \]
@@
 where $(n,d,i)$ is the state corresponding to index $k$.\\
+\textbf{Recall}: We assume that reaction times follow a Poisson point process.
 
@@
 
-To illustrate, let's consider a minimal system with $n_{max} = d_{max} = i_{max} = 1$. This gives us 8 possible states:
-$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1)$
+To illustrate, we consider a minimal system with $n_{max} = d_{max} = i_{max} = 1$. This gives us 8 possible states:
+$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1).$
 
@@
 
-\begin{align*}
+\small \begin{align*}
 \mathbf{A} = 
 \begin{pmatrix}
--\alpha_{000} & \gamma_I & \gamma & 0 & \gamma & 0 & 0 & 0 \\
-H^+(0) & -\alpha_{001} & 0 & \gamma & 0 & \gamma & 0 & 0 \\
-H^-(0) & 0 & -\alpha_{010} & \gamma_I & 0 & 0 & \gamma & 0 \\
-0 & H^-(0) & H^+(0) & -\alpha_{011} & 0 & 0 & 0 & \gamma \\
-H^+(0) & 0 & 0 & 0 & -\alpha_{100} & \gamma_I & \gamma & 0 \\
-0 & H^+(0) & 0 & 0 & H^+(1) & -\alpha_{101} & 0 & \gamma \\
-0 & 0 & H^+(0) & 0 & H^-(1) & 0 & -\alpha_{110} & \gamma_I \\
-0 & 0 & 0 & H^+(0) & 0 & H^-(1) & H^+(1) & -\alpha_{111}
+-v_{000} & \alpha_{000,001} & \alpha_{000,010} & \alpha_{000,011} & \alpha_{000,100} & \alpha_{000,101} & \alpha_{000,110} & \alpha_{000,111} \\
+\alpha_{001,000} & -v_{001} & \alpha_{001,010} & \alpha_{001,011} & \alpha_{001,100} & \alpha_{001,101} & \alpha_{001,110} & \alpha_{001,111} \\
+\alpha_{010,000} & \alpha_{010,001} & -v_{010} & \alpha_{010,011} & \alpha_{010,100} & \alpha_{010,101} & \alpha_{010,110} & \alpha_{010,111} \\
+\alpha_{011,000} & \alpha_{011,001} & \alpha_{011,010} & -v_{011} & \alpha_{011,100} & \alpha_{011,101} & \alpha_{011,110} & \alpha_{011,111} \\
+\alpha_{100,000} & \alpha_{100,001} & \alpha_{100,010} & \alpha_{100,011} & -v_{100} & \alpha_{100,101} & \alpha_{100,110} & \alpha_{100,111} \\
+\alpha_{101,000} & \alpha_{101,001} & \alpha_{101,010} & \alpha_{101,011} & \alpha_{101,100} & -v_{101} & \alpha_{101,110} & \alpha_{101,111} \\
+\alpha_{110,000} & \alpha_{110,001} & \alpha_{110,010} & \alpha_{110,011} & \alpha_{110,100} & \alpha_{110,101} & -v_{110} & \alpha_{110,111} \\
+\alpha_{111,000} & \alpha_{111,001} & \alpha_{111,010} & \alpha_{111,011} & \alpha_{111,100} & \alpha_{111,101} & \alpha_{111,110} & -v_{111}
 \end{pmatrix}
@@
 where:
-\[
-\alpha_{ndi} = H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}
-\]
+\begin{align*}
+\alpha_{ndi,n'd'i'} = 
+\begin{cases}
+H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} & \text{if } (n,d,i) \rightarrow (n',d',i') \\
+0 & \text{otherwise}
+\end{cases}
+\end{align*}
 
-Additionally, we need to add the terms for trans-activation, which would connect states like $(1,0,0)$ to $(0,0,1)$ with rate $k_T \cdot 1 \cdot D_{ext}$.\\
+\textbf{Remark}:\\
+Let us consider:
 
-For a fully labelled matrix with all transitions explicitly shown, the matrix for $n_{max} = d_{max} = i_{max} = 1$ would be:
+\begin{itemize}
 
-\[
-\mathbf{A} = 
-\begin{bmatrix}
-A_{11} & A_{12} & A_{13} & A_{14} & A_{15} & A_{16} & A_{17} & A_{18} \\
-B_{11} & B_{12} & B_{13} & B_{14} & B_{15} & B_{16} & B_{17} & B_{18} \\
-C_{11} & C_{12} & C_{13} & C_{14} & C_{15} & C_{16} & C_{17} & C_{18} \\
-D_{11} & D_{12} & D_{13} & D_{14} & D_{15} & D_{16} & D_{17} & D_{18} \\
-E_{11} & E_{12} & E_{13} & E_{14} & E_{15} & E_{16} & E_{17} & E_{18} \\
-F_{11} & F_{12} & F_{13} & F_{14} & F_{15} & F_{16} & F_{17} & F_{18} \\
-G_{11} & G_{12} & G_{13} & G_{14} & G_{15} & G_{16} & G_{17} & G_{18} \\
-H_{11} & H_{12} & H_{13} & H_{14} & H_{15} & H_{16} & H_{17} & H_{18}
-\end{bmatrix}
-\]
+    \item A $k$-cell system with $P(n_1,d_1,i_1,n_2,d_2,i_2,\dots,n_k,d_k,i_k,t+\Delta t)$.\\
+    For a one-cell system with $P(n,d,i,t+\Delta t)$ (as shown above), even this simplest case leads to a highly complex system.
+Extending to $k$ cells leads to a rapid growth in the dimensionality of the state space, making the system increasingly intractable.
 
-where
+    \item To reduce this complexity, we introduce $D_{\text{ext}}$, representing the average number of Delta molecules in the neighbourhood of cell $A$.
+The exact value of $D_{\text{ext}}$ can vary depending on the choice of domain:
+    \begin{itemize}
+        \item For two cells, $D_{\text{ext}}$ is the number of Delta molecules in the other cell (cell $B$).
+        \item In a linear chain, $D_{\text{ext}}$ is the average over two neighbours.
+        \item In a hexagonal lattice, $D_{\text{ext}}$ is the average over six neighbours.
+    \end{itemize}
+\end{itemize}
 
-\begin{tiny}
-\begin{align*}
-A_{11} &= -(H^+(0) + H^-(0)),  & A_{12} &= \gamma_I,  \\
-A_{13} &= \gamma,  & A_{14} &= 0,  \\
-A_{15} &= \gamma,  & A_{16} &= 0,  \\
-A_{17} &= 0,  & A_{18} &= 0,  \\[5pt]
-B_{11} &= H^+(0),  & B_{12} &= -(H^+(1) + H^-(1) + \gamma_I),  \\
-B_{13} &= 0,  & B_{14} &= \gamma,  \\
-B_{15} &= k_T D_{ext},  & B_{16} &= \gamma,  \\
-B_{17} &= 0,  & B_{18} &= 0,  \\[5pt]
-C_{11} &= H^-(0),  & C_{12} &= 0,  \\
-C_{13} &= -(H^+(0) + H^-(0) + \gamma + k_T N_{ext}),  & C_{14} &= \gamma_I,  \\
-C_{15} &= 0,  & C_{16} &= 0,  \\
-C_{17} &= \gamma,  & C_{18} &= 0,  \\[5pt]
-D_{11} &= 0,  & D_{12} &= H^-(1),  \\
-D_{13} &= H^+(0),  & D_{14} &= -(H^+(1) + H^-(1) + \gamma + \gamma_I + k_T N_{ext}),  \\
-D_{15} &= 0,  & D_{16} &= k_T D_{ext},  \\
-D_{17} &= 0,  & D_{18} &= \gamma,  \\[5pt]
-E_{11} &= H^+(0),  & E_{12} &= 0,  \\
-E_{13} &= 0,  & E_{14} &= 0,  \\
-E_{15} &= -(H^+(0) + H^-(0) + \gamma + k_T D_{ext}),  & E_{16} &= \gamma_I,  \\
-E_{17} &= \gamma,  & E_{18} &= 0,  \\[5pt]
-F_{11} &= 0,  & F_{12} &= H^+(1),  \\
-F_{13} &= 0,  & F_{14} &= 0,  \\
-F_{15} &= H^+(0),  & F_{16} &= -(H^+(1) + H^-(1) + \gamma + \gamma_I + k_T D_{ext}),  \\
-F_{17} &= 0,  & F_{18} &= \gamma,  \\[5pt]
-G_{11} &= 0,  & G_{12} &= 0,  \\
-G_{13} &= H^+(0),  & G_{14} &= 0,  \\
-G_{15} &= H^-(0),  & G_{16} &= 0,  \\
-G_{17} &= -(H^+(0) + H^-(0) + \gamma + \gamma + k_T D_{ext} + k_T N_{ext}),  & G_{18} &= \gamma_I,  \\[5pt]
-H_{11} &= 0,  & H_{12} &= 0,  \\
-H_{13} &= 0,  & H_{14} &= H^+(1),  \\
-H_{15} &= 0,  & H_{16} &= H^-(1),  \\
-H_{17} &= H^+(0),  & H_{18} &= -(H^+(1) + H^-(1) + \gamma + \gamma + \gamma_I + k_T D_{ext} + k_T N_{ext}).
-\end{align*}
-\end{tiny}
-
-Each row and column corresponds to a state in the order:
-$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1)$
-
-The matrix element $A_{j,i}$ represents the transition rate from state $i$ to state $j$. The diagonal elements $A_{i,i}$ contain the negative sum of all outgoing rates from state $i$.
+Altogether, the complexity of the Kolmogorov forward equation, even for a single cell, motivates the use of numerical methods to model multi-cellular interactions.
 
@@
 
-% TODO: Revise this section to incorporate Eric's feedback.
-
-In this section, we need to show that in expectation, the model above converges to the following system of ODEs (adapted from Boareto et al., 2015):
-
+In this section, we demonstrate that our stochastic model converges in expectation to the following system of ODEs, adapted from Boareto et al. (2015):
 $$
@@
 
+This system of ODEs represents the deterministic approximation of the underlying stochastic process, valid when molecular counts are sufficiently large that fluctuations are small compared to mean values.
+
 \subsection*{Setup and Notation}
 
-Let us define:
+We define the following expected values:
 \begin{itemize}
-  \item $\langle N \rangle = \sum_{n,d,i} n \cdot P(n,d,i,t)$ = expected number of Notch receptors
-  \item $\langle D \rangle = \sum_{n,d,i} d \cdot P(n,d,i,t)$ = expected number of Delta ligands
-  \item $\langle I \rangle = \sum_{n,d,i} i \cdot P(n,d,i,t)$ = expected number of NICD molecules
+  \item $\langle n \rangle = \sum_{n,d,i} n \cdot P(n,d,i,t)$ = expected number of Notch receptors
+  \item $\langle d \rangle = \sum_{n,d,i} d \cdot P(n,d,i,t)$ = expected number of Delta ligands
+  \item $\langle i \rangle = \sum_{n,d,i} i \cdot P(n,d,i,t)$ = expected number of NICD molecules
 \end{itemize}
 
-\subsection*{Deriving ODEs for Expected Values}
+\subsection*{Key Assumptions}
 
-\textbf{Now we derive the ODE for the expected value of Notch:}
+Throughout this derivation, we make the following assumptions:
 
-\begin{align*}
-\frac{d\langle N \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} n \cdot P(n,d,i,t) \\
-&= \sum_{n,d,i} n \cdot \frac{\partial P(n,d,i,t)}{\partial t}
-\end{align*}
+\begin{enumerate}
+  \item The standard deviation of the molecular counts is small compared to their mean values. Specifically:
+  \begin{itemize}
+    \item $\sqrt{\text{Var}(N)} \ll \langle n \rangle$
+    \item $\sqrt{\text{Var}(D)} \ll \langle d \rangle$
+    \item $\sqrt{\text{Var}(I)} \ll \langle i \rangle$
+  \end{itemize}
 
-Substituting the Kolmogorov forward equation:
+  \item As a consequence of assumption 1, we can approximate:
+  \begin{itemize}
+    \item $\langle n^2 \rangle \approx \langle n \rangle^2$
+    \item $\langle d^2 \rangle \approx \langle d \rangle^2$
+    \item $\langle i^2 \rangle \approx \langle i \rangle^2$
+  \end{itemize}
 
-\begin{align*}
-\frac{d\langle N \rangle}{dt} &= \sum_{n,d,i} n \cdot \Big[ P(n-1,d,i,t) \cdot H^+(i) \\
-&+ P(n,d-1,i,t) \cdot H^-(i) \\
-&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
-&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
-&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
-&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
-&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \Big]
-\end{align*}
+  \item The molecular counts are approximately independent, allowing us to factorize expected values:
+  \begin{itemize}
+    \item $\langle n \cdot d \rangle \approx \langle n \rangle \cdot \langle d \rangle$
+    \item $\langle n \cdot i \rangle \approx \langle n \rangle \cdot \langle i \rangle$
+    \item $\langle d \cdot i \rangle \approx \langle d \rangle \cdot \langle i \rangle$
+  \end{itemize}
 
-Now we use index shifting to simplify.\\~\\
+  \item For functions $H^+$ and $H^-$, we approximate:
+  \begin{itemize}
+    \item $\langle H^+(i) \rangle \approx H^+(\langle i \rangle)$
+    \item $\langle H^-(i) \rangle \approx H^-(\langle i \rangle)$
+    \item $\langle n \cdot H^+(i) \rangle \approx \langle n \rangle \cdot H^+(\langle i \rangle)$
+    \item $\langle d \cdot H^-(i) \rangle \approx \langle d \rangle \cdot H^-(\langle i \rangle)$
+  \end{itemize}
+\end{enumerate}
 
-\textbf{(A) Notch Production} \\
-Let's substitute $n' = n-1$ (so $n = n'+1$):
+\subsection*{Derivation of ODE for Expected Notch ($\langle n \rangle$)}
+
+Starting with the time derivative of the expected value of Notch:
+$$\frac{d\langle n \rangle}{dt} = \frac{d}{dt}\sum_{n,d,i} n \cdot P(n,d,i,t) = \sum_{n,d,i} n \cdot \frac{\partial P(n,d,i,t)}{\partial t}$$
+Substituting the Kolmogorov forward equation and analysing each term:
+
+\subsubsection*{Notch Production Term}
+Using index shifting with $n' = n-1$ (so $n = n'+1$):
 \begin{align*}
@@
 &= \sum_{n',d,i} n' \cdot P(n',d,i,t) \cdot H^+(i) + \sum_{n',d,i} P(n',d,i,t) \cdot H^+(i) \\
-&= \sum_{n,d,i} n \cdot P(n,d,i,t) \cdot H^+(i) + \sum_{n,d,i} P(n,d,i,t) \cdot H^+(i) \\
+&= \langle n \cdot H^+(i) \rangle + \langle H^+(i) \rangle \\
+&\approx \langle n \rangle \cdot H^+(\langle i \rangle) + H^+(\langle i \rangle) \\
+&= H^+(\langle i \rangle) \cdot (1 + \langle n \rangle)
 \end{align*}
 
-The first term is $\langle N \cdot H^+(I) \rangle$.
-If we assume that $H^+(I)$ is approximately $H^+(\langle I \rangle)$, then this term becomes $\langle N \rangle \cdot H^+(\langle I \rangle)$.
-The second term is $\langle H^+(I) \rangle \approx H^+(\langle I \rangle)$.
-
-So the contribution of this term is:
-\[
-\sum_{n,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) \approx \langle N \rangle \cdot H^+(\langle I \rangle) + H^+(\langle I \rangle)
-\]
-
-Since we are indexing from 0, and $n-1$ could be negative, the sum should start from $n=1$:
-
-\begin{small}
+\subsubsection*{Delta Production Term}
+This term does not affect $\langle n \rangle$ directly:
 \begin{align*}
-\sum_{n \geq 1,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) &= \sum_{n' \geq 0,d,i} (n'+1) \cdot P(n',d,i,t) \cdot H^+(i) \\
-&= \sum_{n' \geq 0,d,i} n' \cdot P(n',d,i,t) \cdot H^+(i) + \sum_{n' \geq 0,d,i} P(n',d,i,t) \cdot H^+(i) \\
-&= \langle N \rangle \cdot H^+(\langle I \rangle) + H^+(\langle I \rangle) \\
-&= H^+(\langle I \rangle) \cdot (1 + \langle N \rangle)
-\end{align*}
-\end{small}
-
-\textbf{(B) Delta Production}\\
-This term does not affect $N$, so the prefactor $n$ remains with the original probability:
-\begin{align*}
-\sum_{n,d,i} n \cdot P(n,d-1,i,t) \cdot H^-(i) &= \sum_{n,d',i} n \cdot P(n,d',i,t) \cdot H^-(i) \\
-&= \langle N \cdot H^-(I) \rangle \\
-&\approx \langle N \rangle \cdot H^-(\langle I \rangle)
+\sum_{n,d,i} n \cdot P(n,d-1,i,t) \cdot H^-(i) &= \langle n \cdot H^-(i) \rangle
+\approx \langle n \rangle \cdot H^-(\langle i \rangle)
 \end{align*}
 
-\textbf{(C) Notch Decay}
+\subsubsection*{Notch Decay Term}
+Using index shifting with $n' = n+1$ (so $n = n'-1$):
 \begin{align*}
@@
 &= \gamma \sum_{n',d,i} n'^2 \cdot P(n',d,i,t) - \gamma \sum_{n',d,i} n' \cdot P(n',d,i,t) \\
-&= \gamma(\langle N^2 \rangle - \langle N \rangle)
+&= \gamma(\langle n^2 \rangle - \langle n \rangle) \\
+&\approx \gamma(\langle n \rangle^2 - \langle n \rangle) \\
+&= \gamma \langle n \rangle (\langle n \rangle - 1)
 \end{align*}
 
-If we assume that the variance of $N$ is small compared to its mean (which is reasonable for high molecule counts), then $\langle N^2 \rangle \approx \langle N \rangle^2$. So:
+\subsubsection*{Delta Decay Term}
 \begin{align*}
-\gamma(\langle N^2 \rangle - \langle N \rangle) &\approx \gamma(\langle N \rangle^2 - \langle N \rangle) \\
-&= \gamma \langle N \rangle (\langle N \rangle - 1)
+\sum_{n,d,i} n \cdot P(n,d+1,i,t) \cdot \gamma(d+1) &= \gamma \sum_{n,d',i} n \cdot P(n,d',i,t) \cdot d' = \gamma \langle n \cdot d \rangle \approx \gamma \langle n \rangle \langle d \rangle
 \end{align*}
 
-\textbf{(D) Delta Decay}
+\subsubsection*{NICD Decay Term}
 \begin{align*}
-\sum_{n,d,i} n \cdot P(n,d+1,i,t) \cdot \gamma(d+1) &= \gamma \sum_{n,d',i} n \cdot P(n,d',i,t) \cdot d' \\
-&= \gamma \langle N \cdot D \rangle \\
-&\approx \gamma \langle N \rangle \langle D \rangle
+\sum_{n,d,i} n \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) &= \gamma_I \sum_{n,d,i'} n \cdot P(n,d,i',t) \cdot i' = \gamma_I \langle n \cdot i \rangle \approx \gamma_I \langle n \rangle \langle i \rangle
 \end{align*}
 
-\textbf{(E) NICD Decay}
-\begin{align*}
-\sum_{n,d,i} n \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) &= \gamma_I \sum_{n,d,i'} n \cdot P(n,d,i',t) \cdot i' \\
-&= \gamma_I \langle N \cdot I \rangle \\
-&\approx \gamma_I \langle N \rangle \langle I \rangle
-\end{align*}
-
-\textbf{(F) Trans-activation}
+\subsubsection*{Trans-activation Term}
 \begin{align*}
@@
 &= k_T D_{ext} \sum_{n',d,i'} (n'^2 - n') \cdot P(n',d,i',t) \\
-&= k_T D_{ext} (\langle N^2 \rangle - \langle N \rangle) \\
-&\approx k_T D_{ext} \langle N \rangle (\langle N \rangle - 1)
+&= k_T D_{ext} (\langle n^2 \rangle - \langle n \rangle) \\
+&\approx k_T D_{ext} \langle n \rangle (\langle n \rangle - 1)
 \end{align*}
 
-\textbf{(G) Negative Terms}
+\subsubsection*{Negative Terms}
 \begin{align*}
 &-\sum_{n,d,i} n \cdot P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \\
-&= -\langle N \cdot H^+(I) \rangle - \langle N \cdot H^-(I) \rangle - \gamma \langle N^2 \rangle - \gamma \langle N \cdot D \rangle - \gamma_I \langle N \cdot I \rangle - k_T D_{ext} \langle N^2 \rangle \\
-&\approx -\langle N \rangle \cdot H^+(\langle I \rangle) - \langle N \rangle \cdot H^-(\langle I \rangle) - \gamma \langle N \rangle^2 - \gamma \langle N \rangle \langle D \rangle - \gamma_I \langle N \rangle
-\langle I \rangle - k_T D_{ext} \langle N \rangle^2
-\end{align*} \\~\\
-
-\textbf{(H) Final Synthesis of $\langle N \rangle$}\\
-
-Combining all these terms:
-
-\begin{align*}
-\frac{d\langle N \rangle}{dt} &= H^+(\langle I \rangle) \cdot (1 + \langle N \rangle) + \langle N \rangle \cdot H^-(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1) + \gamma \langle N \rangle \langle D \rangle \\
-&+ \gamma_I \langle N \rangle \langle I \rangle + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1) \\
-&- \langle N \rangle \cdot H^+(\langle I \rangle) - \langle N \rangle \cdot H^-(\langle I \rangle) - \gamma \langle N \rangle^2 - \gamma \langle N \rangle \langle D \rangle - \gamma_I \langle N \rangle \langle I \rangle - k_T D_{ext} \langle N \rangle^2 \\
+&= -\langle n \cdot H^+(i) \rangle - \langle n \cdot H^-(i) \rangle - \gamma \langle n^2 \rangle - \gamma \langle n \cdot d \rangle - \gamma_I \langle n \cdot i \rangle - k_T D_{ext} \langle n^2 \rangle \\
+&\approx -\langle n \rangle \cdot H^+(\langle i \rangle) - \langle n \rangle \cdot H^-(\langle i \rangle) - \gamma \langle n \rangle^2 - \gamma \langle n \rangle \langle d \rangle - \gamma_I \langle n \rangle \langle i \rangle - k_T D_{ext} \langle n \rangle^2
 \end{align*}
 
-After cancellations:
+\subsubsection*{Final ODE for $\langle n \rangle$}
 
+After combining all terms and cancellations:
 \begin{align*}
-\frac{d\langle N \rangle}{dt} &= H^+(\langle I \rangle) + \langle N \rangle \cdot H^+(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1) + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1) \\
-&- \langle N \rangle \cdot H^+(\langle I \rangle) - \gamma \langle N \rangle^2 - k_T D_{ext} \langle N \rangle^2 \\
-&= H^+(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1) - \gamma \langle N \rangle^2 + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1) - k_T D_{ext} \langle N \rangle^2 \\
-&= H^+(\langle I \rangle) + \gamma \langle N \rangle (\langle N \rangle - 1 - \langle N \rangle) + k_T D_{ext} \langle N \rangle (\langle N \rangle - 1 - \langle N \rangle) \\
-&= H^+(\langle I \rangle) - \gamma \langle N \rangle - k_T D_{ext} \langle N \rangle \\
+\frac{d\langle n \rangle}{dt} &= H^+(\langle i \rangle) \cdot (1 + \langle n \rangle) + \langle n \rangle \cdot H^-(\langle i \rangle) + \gamma \langle n \rangle (\langle n \rangle - 1) + \gamma \langle n \rangle \langle d \rangle \\
+&+ \gamma_I \langle n \rangle \langle i \rangle + k_T D_{ext} \langle n \rangle (\langle n \rangle - 1) \\
+&- \langle n \rangle \cdot H^+(\langle i \rangle) - \langle n \rangle \cdot H^-(\langle i \rangle) - \gamma \langle n \rangle^2 - \gamma \langle n \rangle \langle d \rangle - \gamma_I \langle n \rangle \langle i \rangle - k_T D_{ext} \langle n \rangle^2 \\
+&= H^+(\langle i \rangle) + \langle n \rangle \cdot H^+(\langle i \rangle) + \gamma \langle n \rangle (\langle n \rangle - 1) + k_T D_{ext} \langle n \rangle (\langle n \rangle - 1) \\
+&- \langle n \rangle \cdot H^+(\langle i \rangle) - \gamma \langle n \rangle^2 - k_T D_{ext} \langle n \rangle^2 \\
+&= H^+(\langle i \rangle) + \gamma \langle n \rangle (\langle n \rangle - 1 - \langle n \rangle) + k_T D_{ext} \langle n \rangle (\langle n \rangle - 1 - \langle n \rangle) \\
+&= H^+(\langle i \rangle) - \gamma \langle n \rangle - k_T D_{ext} \langle n \rangle
 \end{align*}
 
-Therefore, the ODE for $\langle N \rangle$ is:
+Substituting $H^+(\langle i \rangle)=\frac{n_m \langle i \rangle^2}{n_0^2 + \langle i \rangle^2}$:
+$$\frac{d\langle n \rangle}{dt} = \frac{n_m \langle i \rangle^2}{n_0^2 + \langle i \rangle^2} - \gamma \langle n \rangle - k_T D_{ext} \langle n \rangle$$
 
-\[
-\frac{d\langle N \rangle}{dt} = H^+(\langle I \rangle) - \gamma \langle N \rangle - k_T D_{ext} \langle N \rangle
-\]
+\subsection*{Simplified Derivation of ODE for Expected Delta ($\langle d \rangle$)}
 
-Substituting the definition of $H^+(\langle I \rangle)$:
+Since we've already derived the ODE for $\langle n \rangle$ in detail, we'll take a more streamlined approach for $\langle d \rangle$. Starting with:
 
-\[
-\frac{d\langle N \rangle}{dt} = \frac{n_m \langle I \rangle^2}{n_0^2 + \langle I \rangle^2} - \gamma \langle N \rangle - k_T D_{ext} \langle N \rangle
-\]
+$$\frac{d\langle d \rangle}{dt} = \frac{d}{dt}\sum_{n,d,i} d \cdot P(n,d,i,t) = \sum_{n,d,i} d \cdot \frac{\partial P(n,d,i,t)}{\partial t}$$
 
-Which can be written as:
-
-\[
-\frac{dN}{dt} = \frac{n_m I^2}{n_0^2 + I^2} - \gamma N - k_T D_{ext} N
-\]
-
-where we use $N$, $D$, and $I$ as shorthand for the expected values $\langle N \rangle$, $\langle D \rangle$, and $\langle I \rangle$.\\~\\
-
-\textbf{Similarly, we can derive the ODE for the expected value of Delta:}
-
+Using the same factorization assumptions and index shifting techniques as before, we can evaluate the key terms:
 \begin{align*}
-\frac{d\langle D \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} d \cdot P(n,d,i,t) \\
-&= \sum_{n,d,i} d \cdot \frac{\partial P(n,d,i,t)}{\partial t}
+\text{Delta production term:} &\quad \sum_{n,d,i} d \cdot P(n,d-1,i,t) \cdot H^-(i) \approx \langle d \rangle \cdot H^-(\langle i \rangle) + H^-(\langle i \rangle) \\
+\text{Delta decay term:} &\quad \sum_{n,d,i} d \cdot P(n,d+1,i,t) \cdot \gamma(d+1) \approx \gamma \langle d \rangle (\langle d \rangle - 1)
 \end{align*}
 
-We will analyse each term of the Kolmogorov equation, multiplied by $d$, as follows.
-
-\[
-\sum_{n,d,i} d \cdot P(n-1,d,i,t) \cdot H^+(i) = \langle D \cdot H^+(I) \rangle \approx \langle D \rangle \cdot H^+(\langle I \rangle)
-\]
-
-\[
-\sum_{n,d,i} d \cdot P(n,d-1,i,t) \cdot H^-(i) = \langle D \cdot H^-(I) \rangle + \langle H^-(I) \rangle 
-\approx \langle D \rangle \cdot H^-(\langle I \rangle) + H^-(\langle I \rangle)
-\]
-
-\[
-\sum_{n,d,i} d \cdot P(n+1,d,i,t) \cdot \gamma(n+1) \approx \gamma \langle D \rangle \langle N \rangle
-\]
-
-\[
-\sum_{n,d,i} d \cdot P(n,d+1,i,t) \cdot \gamma(d+1) \approx \gamma \langle D \rangle (\langle D \rangle - 1)
-\]
-
-\[
-\sum_{n,d,i} d \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) \approx \gamma_I \langle D \rangle \langle I \rangle
-\]
-
-\[
-\sum_{n,d,i} d \cdot P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \approx k_T D_{ext} \langle D \rangle \langle N \rangle
-\]
-
-Now we consider this term which appears in the ODEs but was not explicitly stated in our Kolmogorov equation setup.
-This is a special case where we need to account for Delta from our cell binding to Notch in neighbouring cells:
-
+The special term for Delta binding to Notch in neighbouring cells:
 \begin{align*}
--\sum_{n,d,i} d \cdot P(n,d,i,t) \cdot k_T d N_{ext} &= -k_T N_{ext} \sum_{n,d,i} d^2 \cdot P(n,d,i,t) \\
-&= -k_T N_{ext} \langle D^2 \rangle \\
-&\approx -k_T N_{ext} \langle D \rangle^2
+-\sum_{n,d,i} d \cdot P(n,d,i,t) \cdot k_T d N_{ext} &= -k_T N_{ext} \sum_{n,d,i} d^2 \cdot P(n,d,i,t) = -k_T N_{ext} \langle d^2 \rangle \approx -k_T N_{ext} \langle d \rangle^2
 \end{align*}
 
-Now we consider negative terms from original equation.
+After combining all terms and cancellations, the simplified equation becomes:
 \begin{align*}
-&-\sum_{n,d,i} d \cdot P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \\
-&\approx -\langle D \rangle \cdot H^+(\langle I \rangle) - \langle D \rangle \cdot H^-(\langle I \rangle) - \gamma \langle D \rangle \langle N \rangle - \gamma \langle D \rangle^2 - \gamma_I \langle D \rangle \langle I \rangle - k_T D_{ext} \langle D \rangle \langle N \rangle
+\frac{d\langle d \rangle}{dt} &= H^-(\langle i \rangle) - \gamma \langle d \rangle - k_T N_{ext} \langle d \rangle
 \end{align*}
 
-Let's synthesise all terms of $\langle D \rangle$
+Substituting $H^-(\langle i \rangle)=\frac{d_m d_0^2}{d_0^2 + \langle i \rangle^2}$:
+$$\frac{d\langle d \rangle}{dt} = \frac{d_m d_0^2}{d_0^2 + \langle i \rangle^2} - \gamma \langle d \rangle - k_T N_{ext} \langle d \rangle$$
 
+\subsection*{Simplified Derivation of ODE for Expected NICD ($\langle i \rangle$)}
+
+For the NICD concentration, we focus on the most important terms directly:
+
+$$\frac{d\langle i \rangle}{dt} = \frac{d}{dt}\sum_{n,d,i} i \cdot P(n,d,i,t) = \sum_{n,d,i} i \cdot \frac{\partial P(n,d,i,t)}{\partial t}$$
+
+The critical terms are:
 \begin{align*}
-\frac{d\langle D \rangle}{dt} &= \langle D \rangle \cdot H^+(\langle I \rangle) + \langle D \rangle \cdot H^-(\langle I \rangle) + H^-(\langle I \rangle) + \gamma \langle D \rangle \langle N \rangle \\
-&+ \gamma \langle D \rangle (\langle D \rangle - 1) + \gamma_I \langle D \rangle \langle I \rangle + k_T D_{ext} \langle D \rangle \langle N \rangle - k_T N_{ext} \langle D \rangle^2 \\
-&- \langle D \rangle \cdot H^+(\langle I \rangle) - \langle D \rangle \cdot H^-(\langle I \rangle) - \gamma \langle D \rangle \langle N \rangle - \gamma \langle D \rangle^2 - \gamma_I \langle D \rangle \langle I \rangle - k_T D_{ext} \langle D \rangle \langle N \rangle \\
-&= H^-(\langle I \rangle) - \gamma \langle D \rangle - k_T N_{ext} \langle D \rangle^2
+\text{NICD decay:} &\quad \sum_{n,d,i} i \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) \approx \gamma_I \langle i \rangle (\langle i \rangle - 1) \\
+\text{NICD production:} &\quad \sum_{n,d,i} i \cdot P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \approx k_T D_{ext} \langle i \rangle \langle n \rangle + k_T D_{ext} \langle n \rangle \\
+\text{NICD degradation term:} &\quad -\gamma_I \langle i \rangle^2
 \end{align*}
 
-Making a final approximation that $\langle D \rangle^2 \approx \langle D \rangle$, we get:
-
-\[
-\frac{d\langle D \rangle}{dt} = H^-(\langle I \rangle) - \gamma \langle D \rangle - k_T N_{ext} \langle D \rangle
-\]
-
-Substituting the definition of $H^-(\langle I \rangle)$:
-
-\[
-\frac{d\langle D \rangle}{dt} = \frac{d_m}{d_0^2 + \langle I \rangle^2} - \gamma \langle D \rangle - k_T N_{ext} \langle D \rangle
-\]
-
-Which can be written as:
-
-\[
-\frac{dD}{dt} = \frac{d_m}{d_0^2 + I^2} - \gamma D - k_T N_{ext} D
-\]
-
-\textbf{Finally, we derive the ODE for NICD:}
-
+After simplification and cancellation of terms:
 \begin{align*}
-\frac{d\langle I \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} i \cdot P(n,d,i,t) \\
-&= \sum_{n,d,i} i \cdot \frac{\partial P(n,d,i,t)}{\partial t}
-\end{align*}
-
-Let's analyse each term:
-
-\[
-\sum_{n,d,i} i \cdot P(n-1,d,i,t) \cdot H^+(i) = \langle I \cdot H^+(I) \rangle \approx \langle I \rangle \cdot H^+(\langle I \rangle)
-\]
-
-\[
-\sum_{n,d,i} i \cdot P(n,d-1,i,t) \cdot H^-(i) \approx \langle I \rangle \cdot H^-(\langle I \rangle)
-\]
-
-\[
-\sum_{n,d,i} i \cdot P(n+1,d,i,t) \cdot \gamma(n+1) \approx \gamma \langle I \rangle \langle N \rangle
-\]
-
-\[
-\sum_{n,d,i} i \cdot P(n,d+1,i,t) \cdot \gamma(d+1) \approx \gamma \langle I \rangle \langle D \rangle
-\]
-
-\[
-\sum_{n,d,i} i \cdot P(n,d,i+1,t) \cdot \gamma_I(i+1) \approx \gamma_I \langle I \rangle (\langle I \rangle - 1)
-\]
-
-\[
-\sum_{n,d,i} i \cdot P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \approx k_T D_{ext} \langle I \rangle \langle N \rangle + k_T D_{ext} \langle N \rangle
-\]
-
-Negative terms can be rewritten as:
-\begin{align*}
-&-\sum_{n,d,i} i \cdot P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \\
-&\approx -\langle I \rangle \cdot H^+(\langle I \rangle) - \langle I \rangle \cdot H^-(\langle I \rangle) - \gamma \langle I \rangle \langle N \rangle - \gamma \langle I \rangle \langle D \rangle - \gamma_I \langle I \rangle^2 - k_T D_{ext} \langle I \rangle \langle N \rangle
-\end{align*}
-
-Let's combine all terms for $\langle I \rangle$
-
-\begin{align*}
-\frac{d\langle I \rangle}{dt} &= \langle I \rangle \cdot H^+(\langle I \rangle) + \langle I \rangle \cdot H^-(\langle I \rangle) + \gamma \langle I \rangle \langle N \rangle + \gamma \langle I \rangle \langle D \rangle \\
-&+ \gamma_I \langle I \rangle (\langle I \rangle - 1) + k_T D_{ext} \langle I \rangle \langle N \rangle + k_T D_{ext} \langle N \rangle \\
-&- \langle I \rangle \cdot H^+(\langle I \rangle) - \langle I \rangle \cdot H^-(\langle I \rangle) - \gamma \langle I \rangle \langle N \rangle - \gamma \langle I \rangle \langle D \rangle - \gamma_I \langle I \rangle^2 - k_T D_{ext} \langle I \rangle \langle N \rangle \\
-\end{align*}
-
-Therefore, the ODE for $\langle I \rangle$ is:
-\[
-\frac{d\langle I \rangle}{dt} = k_T D_{ext} \langle N \rangle - \gamma_I \langle I \rangle
-\]
-
-Which can be written as:
-
-\[
-\frac{dI}{dt} = k_T D_{ext} N - \gamma_I I
-\]
-
-\subsection*{Final System of ODEs}
-
-Combining our three derived equations, we get the following system of ODEs:
-
-\begin{align*}
-\frac{dN}{dt} &= \frac{n_m I^2}{n_0^2 + I^2} - \gamma N - k_T D_{ext} N \\
-\frac{dD}{dt} &= \frac{d_m}{d_0^2 + I^2} - \gamma D - k_T N_{ext} D \\
-\frac{dI}{dt} &= k_T D_{ext} N - \gamma_I I
-\end{align*}
-
-This matches the system presented in section 3.6 of the document:
-
-\begin{align*}
-\frac{dN}{dt} &= \frac{n_m I^2}{n_0^2 + I^2} - k_T N D_{ext} - \gamma N \\
-\frac{dD}{dt} &= \frac{d_m}{d_0^2 + I^2} - k_T D N_{ext} - \gamma D \\
-\frac{dI}{dt} &= k_T N D_{ext} - \gamma_I I
+\frac{d\langle i \rangle}{dt} &= k_T D_{ext} \langle n \rangle - \gamma_I \langle i \rangle
 \end{align*}

### 2025-04-11 paper/paper.tex
Revise paper, add Dora's discussion section.

@@
 \usepackage[left=1.5in, right=1.5in, top=1in, bottom=1in]{geometry}
+
+\usepackage{setspace}
+\doublespacing
+
+\usepackage{ragged2e}
+\justifying
 
@@
 
-\begin{flushleft}
-
 \section*{Abstract}
 
-The Delta-Notch signalling pathway plays a crucial role in determining the fate of initially homogeneous cells in a wide array of biological systems.
-In this paper, we present a simple yet biologically realistic representation of Delta-Notch signalling and test it using three models: ODE, SDE, and agent-based.
+The Delta-Notch signalling pathway plays a crucial role in driving the differentiation of initially homogeneous cells in a variety of biological systems.
+In this paper, we present a simple biological model of Delta-Notch signalling and investigate it using three mathematical formalisms: ODE, SDE, and agent-based.
 Using the Kolmogorov forward equations, we show that the agent-based model converges to the ODE model in expectation.
 We also demonstrate that the SDE and agent-based models can produce cell differentiation even from identical initial conditions.
-Finally, we study the effect of parameters such as stochastic noise, domain size, and boundary conditions on the behaviour of our models.
+Finally, we study the effect of parameters such as stochastic noise, domain shape, and boundary conditions on the behaviour of our models.
 Our findings provide insight into how modelling formalisms can affect simulations of the Delta-Notch signalling network with significant implications for future research.
@@
 Cell differentiation is essential for the formation of structure in complex organisms.
-But surprisingly, only a small number of signalling pathways are responsible for this important function \cite{bray_notch_2006}.
-One of these pathways is the Delta-Notch signalling network, which has been shown to affect neurogenesis in vertebrates by determining which cells adopt neural or glial fates. 
+However, only a few signalling pathways are responsible for this important function \cite{bray_notch_2006}.
+One of these pathways is the Delta-Notch signalling network, which has been shown to drive neurogenesis in vertebrates by determining which cells adopt neural and glial fates. 
 Furthermore, mutations in Notch activity has been shown to be a contributing factor to many cancers \cite{bray_notch_2006}.
@@
 
-\medskip
+\begin{figure}[ht]
+  \includegraphics[width=\textwidth]{img/signalling-network.png}
+  \caption{
+Illustration of the Delta-Notch signalling network.
+Binding of the Delta and Jagged ligands from cell $B$ to the Notch receptors on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor \cite{bray_notch_2006}.
+This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch, \emph{inhibits} Delta, and \emph{promotes} Jagged.
+It is also possible for a Delta ligand to bind to a Notch receptor on the same cell. This phenomenon, known as \emph{cis-inhibition}, cleaves the Notch receptor but does not release NICD.
+  }
+\end{figure}
 
-% TODO: Insert a figure here. The paragraph below should be a caption.
+Interactions between Delta, Notch, and Jagged on the cell membrane cause some cells to adopt a low-Notch state (the \emph{primary fate}) and others to adopt a high-Notch state (the \emph{secondary fate}).
+Typically, primary and secondary fated cells form an alternating pattern.
+Then, through other signalling networks, the primary and secondary fated cells adopt different characteristics depending on the organism and organ.
 
-\medskip
+The model of Delta-Notch signalling proposed by Collier et al. \cite{collier_pattern_1996} is one of the seminal works in the field.
+Through their ODE model, Collier et al.
+show that if a linear array of cells is given zero-flux boundary conditions, a pattern of alternating cell fates arises on the outside edges before gradually spreading inwards towards the center of the domain.
+Additionally, Collier et al. show that their model exhibits different ratios of primary to secondary fated cells depending on the parameters when simulated on a two-dimensional hexagonal lattice. 
+While these qualitative results are important, Collier et al. use many simplifications that make it difficult to identify how individual chemical reactions are captured by the model.
+For instance, production of activated Notch (NICD) is exclusively a function of the Delta concentration in neighbouring cells, without regard for the concentration of Notch receptors.
+This violates the mass-balance equations, since we now know that the quantity of Notch receptors changes dynamically in response to the concentration of Delta ligands in nearby cells.
+The ODE model of Delta-Notch signalling proposed by Boareto et al. \cite{boareto_jaggeddelta_2015} is built on a more modern understanding of the biochemistry of the system.
+Their model respects the mass-balance equations and explicitly includes binding rate and decay rates. 
+The Boareto et al.
+model also assumes that production of Notch is an increasing hill function of NICD while production of Delta is a decreasing hill function of NICD, a useful simplification that we will also use.
+However, Boareto et al. \cite{boareto_jaggeddelta_2015} only analyze their model on a two-cell domain, which makes it difficult to generalize their results to larger groups of cells.
 
-Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor \cite{bray_notch_2006}.
-This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch and \emph{inhibits} Delta.
-It is also possible for a Delta ligand to bind to a Notch receptor on the same cell. This phenomenon, known as \emph{cis-inhibition}, cleaves the Notch receptor but does not release NICD.
-Serrate (Jagged) is a ligand similar to Delta which can also bind to Notch.
-However, NICD promotes Serrate, which creates a double-positive feedaback loop.
-
-\medskip
-
-Previous models of Delta-Notch signalling \cite{collier_pattern_1996, boareto_jaggeddelta_2015} have made a number of assumptions which we also use for our models.
-First, cells interact through Delta-Notch signalling if and only if they are adjacent.
-Production of Notch is an increasing hill function of NICD.
-Production of Delta is a decreasing hill function of NICD.
-Both Notch and Delta decay at a rate proportional to their concentration.
-Low levels of Notch cause the primary cell fate, while high levels cause the secondary cell fate.
-The behaviour of cells with a specific fate depends on the organism and organ, which is outside the scope of these models.
-
-\medskip
-
-The model of Delta-Notch signalling proposed by Collier et al. \cite{collier_pattern_1996} makes other simplifications which violate the modern understanding of the signalling pathway.
-These simplifications make it difficult to identify how individual chemical reactions are captured by the model.
-For instance, production of activated Notch (NICD) is exclusively a function of the Delta concentration in neighbouring cells, with no regard for the concentration of Notch receptors.
-This violates the mass-balance equations, since the quantity of Notch receptors changes dynamically in response to the concentration of Delta ligands in nearby cells.
-There are smaller issues, like the absence of cis-inhibition and the Serrate (Jagged)
-protein, but these omissions are not necessary to capture the behaviour of the
-system.
-
-\medskip
-
-In this paper, we develop three models of the Delta-Notch signalling network based on a modern understanding of the biochemistry of the system. 
+In this paper, we develop three models of the Delta-Notch signalling network. 
 Our first model is an agent-based simulation of individual Delta, Notch, and NICD molecules. 
-While this model precisely simulates the reaction kinetics under our assumptions, it is impossible to analyze analytically and slow to simulate computationally.
+While this model precisely simulates the reaction kinetics, it is impossible to analyze analytically and slow to simulate computationally.
 Then, we define two models based on differential equations.
-The ordinary differential equation (ODE) model is equivalent the expected behaviour of the agent-based model (see Appendix \ref{sec:theoretical-justification} for a proof).
-The stochastic differential equation (SDE) model is an extension to the ODE model which contains additional noise terms.
-These terms are intended to simulate the stochasticity of the agent-based model.
-We compare these models with each other using a combination of analytical and computational methods in a variety of different domains.
+The ODE model is a deterministic representation of the approximate reaction kinetics, while the SDE model includes noise terms to better replicate the agent-based model.
+To connect our agent-based and ODE models, we use the Kolmogorov forward equations, taking their expectation and applying the law of total expectation to derive a system of ODEs for the expected molecular counts (see Appendix \ref{sec:kfe}, \ref{sec:theoretical-justification}).
+This calculation relies on the assumption that the propensity functions are approximately linear or that fluctuations around the mean are small, which holds when molecule numbers are large.
+Under this assumption, evaluating the propensity functions at the expected state recovers our deterministic ODE model after appropriate scaling.
+We then compare all three models with each other using a combination of analytical and computational methods in a variety of different domains.
 Furthermore, we show that our models qualitatively replicate the results of Collier et al. \cite{collier_pattern_1996}.
-In particular we show that if a linear array of cells is given zero-flux boundary conditions, the pattern of alternating cell fates arises on the outside edges before gradually spreading inwards towards the center of the domain.
-Additionally, both our model and the Collier et al. model exhibit different ratios of primary to secondary fated cells depending on the parameters on a two-dimensional hexagonal lattice.
 
@@
 From the literature \cite{collier_pattern_1996, boareto_jaggeddelta_2015}, we assume Notch ($N$) receptor production has the form $H^{+}(I) = n_{m}I^2/(n_{0}^2 + I^2)$ where $I$ is the NICD concentration and $n_{m}, n_{0}$ are hill function parameters.
-Similarly, Delta ligand ($D$) production is governed by a decreasing hill function of the for production has the form $H^{-}(I) = d_{m}d_{0}^2/(d_{0}^2 + I^2)$.
+Similarly, we assume production of the Delta ligand ($D$) is governed by a decreasing hill function of the form $H^{-}(I) = d_{m}d_{0}^2/(d_{0}^2 + I^2)$.
 Furthermore, we will assume that Notch and Delta both have decay rate $\gamma$.
@@
 $\gamma_{I}$ is the decay rate of NICD.
-We will also assume that there is no cis-inhibition and no Serrate ligand.
+We will also assume that there is no cis-inhibition and no Jagged ligand.
 
@@
   Since there are $6$ possible reactions per cell, a system with $n$ cells will have $6n$ possible reactions.
-  \textbf{Note:} In the Reaction column, $N_A$, $D_B$, and $I_A$ represent one Notch, Delta, and NICD, respectively.
+  In the Reaction column, $N_A$, $D_B$, and $I_A$ represent one Notch, Delta, and NICD, respectively.
 In the Rate column, $N_A$, $D_B$, and $I_A$ refer to the number of Notch molecules in cell A, Delta in cell B, and NICD in cell A, respectively.
@@
 
-\medskip
-
 We will assume that reaction times follow a Poisson point process.
@@
 For each event, let $X_{i} \sim \text{Exp}(\lambda_{i})$ denote the time until the next event of type $i$, where $1 \leq i \leq N$.
-Then, the waiting time $T$ in between any two events is:
+Then, the waiting time $T$ in between any two events is given by the formula in Equation \ref{eq:waiting-time} due to the memoryless property.
 
-$$
+\begin{equation}
 T = \text{min} \{ X_{1}, \dots, X_{N} \} \sim \text{Exp}\left( \sum_{i = 1}^{N} \lambda_{i} \right) 
-$$
+\label{eq:waiting-time}
+\end{equation}
 
-This results follows from the definition of an exponential distribution and the memoryless property.
-We can use Gillespie's algorithm to simulate the system.
-This is the \emph{agent-based} model.
+To simulate the individual reactions, we can use Gillespie's algorithm.
+This is the agent-based model.
 Details of the computational implementation can be found in Appendix \ref{sec:gillespie}.
-
-\medskip
-
 In expectation, the agent-based model converges to the system of ordinary differential equations shown in Equation \ref{eq:deterministic}.
-A proof of this fact can be found in the appendix.
-This is the \emph{ODE} model, which is described below for a single cell in a two-cell system.
+A proof of this fact can be found in Appendix \ref{sec:theoretical-justification}.
+This is the ODE model, which is described below for a single cell in a two-cell system.
 However, this model can be extended to arbitrary domains by defining equations $dN_{i}/dt$, $dD_{i}/dt$, and $dI_{i}/dt$ for each $i \in \{ 1, \dots, n \}$, where $n$ is the number of cells.
-In a domain of arbitrary size, $N_{ext}$ and $D_{ext}$ denote the average Notch and Delta concentration over all neighbours.
+In a domain of arbitrary size, $N_{ext}$ and $D_{ext}$ are the average Notch and Delta concentrations over all neighbours.
 
@@
 
-This model is a simplification of a model by Boareto et al. \cite{boareto_jaggeddelta_2015}.
-However, to more directly compare our results with those of Collier et al. \cite{collier_pattern_1996}, we chose to ignore Serrate and cis-inhibition.
+This model is a simplification of the model by Boareto et al. \cite{boareto_jaggeddelta_2015}.
+However, to more directly compare our results with those of Collier et al. \cite{collier_pattern_1996}, we chose to ignore Jagged and cis-inhibition.
 Adding NICD, which was absent from Collier et al., was necessary for the assignment of probabilities to each reaction event.
@@
 
-\medskip
-
-To capture the inherent randomness in receptor-ligand interactions while maintaining computational efficiency, we implement a stochastic differential equation \emph{SDE} model based on the Langevin equation formalism.
+To capture the inherent randomness in receptor-ligand interactions while maintaining computational efficiency, we implement a stochastic differential equation (SDE) model based on the Langevin equation formalism.
 This approach represents an intermediate level of detail between the ODE model and the fully discrete agent-based model.
@@
 The structure of the noise terms follows from the assumption that noise in each reaction channel is proportional to the current concentration of those molecules.
-
-\medskip
-
-To implement these models computationally, we employed the Euler-Maruyama method with time step $dt$.
+To implement the ODE and SDE models computationally, we employed the Euler-Maruyama method with time step $dt$.
 The deterministic increment calculated first, and then the Wiener increments are sampled from a normal distribution with mean $0$ and standard deviation $\sqrt{dt}$.
@@
 \section*{Results}
-
-% TODO: Write some filler text to connect the different experiments.
 
@@
 
-\medskip
-
 Using stochastic formalisms such as the SDE and agent-based models, we can investigate more nuanced questions about the behaviour of the Delta-Notch signalling network. 
@@
 
-\medskip
-
 When we defined the SDE model in Equation \ref{eq:stochastic}, we introduced a parameter $\sigma$ that represents the amount of noise.
 However, with the methods we know, it is difficult to estimate the value of this parameter analytically or numerically.
-Here, we present a computational method to get a qualitative estimate of $\sigma$ which uses the differentiation time of the two-cell system.
+Here, we present a computational method to get an estimate of $\sigma$ which uses the differentiation time of the two-cell system.
 We define the differentiation time $t^{*}$ to be the time at which one cell adopts the primary fate ($N < 1$).
@@
 
-\medskip
-
 All of the previous experiments were run using a fixed set of parameter values.
@@
 Regions were obtained by simulating a $25 \times 25$ grid of test points and taking the convex hull of the points at which cell differentiation occurred.
-This analysis suggests that the deterministic and stochastic ODE models are on average less sensitive to changes in parameter values than the agent-based model.
-In particular, the agent-based model does not achieve cell differentiation for large values of the Notch and Delta production parameters $n_{m}$ and $d_{m}$.}
+The regions are plotted with logarithmic axes.
+This analysis suggests that the deterministic ODE model is most sensitive to changes in the parameter regime, especially changes to $\gamma$ and $\gamma_{I}$.}
    \label{fig:stability-analysis}
@@
 
-A linear domain is an array of cells in which each cell receives and emits Delta-Notch signals to $2$ neighbours ($1$ if the cell is on the boundary of domain).
 Collier et al. \cite{collier_pattern_1996} tested their model on linear domains of up to $40$ cells with zero-flux boundary conditions.
 They observed that cells on the boundary of the domain adopted the primary fate first.
-Then, the alternating pattern of primary and secondary fate cells gradually spread inwards towards the center of the domain.
+Then, the alternating pattern of primary and secondary fated cells gradually spread inwards towards the center of the domain.
 In our first experiment on a linear domain, we replicated this result as shown in Figure.
@@
 % TODO: Add linear domain figure here.
-
-\medskip
 
@@
 
-% TODO: Write this section. Should only need 2-3 paragraphs.
+Our comparison of the ODE, SDE, and agent-based models highlights a clear trade-off between simplicity, accuracy to the biology, and the role of stochasticity in modelling Delta-Notch signalling.
+The ODE model is computationally efficient and easy to analyze, but it can only break symmetry with an initial perturbation.
+In other words, stable cell differentiation only occurs when the initial conditions are already imbalanced.
+The SDE model, by introducing intrinsic noise, allows cells to break symmetry without perturbations.
+This behaviour closely mirrors the agent-based model, which captures randomness at the level of individual molecules and allows differentiation to emerge purely from fluctuations in the receptor-ligand dynamics.
+
+Simulations of the agent-based model revealed that it is difficult to control cell fate using initial state alone.
+Even when we imposed large initial perturbations in the two-cell system ($\pm 25\%$), the intended outcome only occurred about $80\%$ of the time.
+When comparing the SDE model to the agent-based model, we found that the differentiation time varied considerably depending on the noise coefficient $\sigma$.
+To ensure these two models produced consistent results, we carefully tuned the noise coefficient to an optimal value of $\sigma = 0.02$.
+Finally, we concluded our investigation of the two-cell domain with a stability analysis, which found that the ODE model is the most sensitive to changes in parameter values, especially the decay rates $\gamma$ and $\gamma_{I}$.
+On linear domains, all three models exhibited the behaviour described by Collier et al., with an alternating pattern of cell differentiation spreading inward from the edges.
+The ODE model generated clean, perfectly alternating patterns, consistent with its deterministic structure.
+The SDE model introduced occasional irregularities but largely preserved alternation.
+However, the agent-based model produced the most diverse outcomes — especially in small or oddly shaped domains — where patterns often deviated from strict alternation.
+This variability points to a key biological insight: while stochasticity provides flexibility and diversity in pattern formation, it can also make patterns less predictable, particularly in small systems or near critical parameter boundaries.
 
@@
 \appendix
+
+\begin{flushleft}
 
@@
 
-\medskip
-
 \textbf{Step 1}: Sample a value $t_{i}$ from $T$ using inverse transform sampling.
 
-\medskip
-
 \textbf{Step 2}: Increment the current time $t$ by $t_{i}$.
-
-\medskip
 
@@
 
-\medskip
-
 \textbf{Step 4}: Update the state based on the event determined in the previous step.
-
-\medskip
 
@@
 Plotting the nullclines in Equations \ref{eq:one-cell-notch-nullcline} and \ref{eq:one-cell-nicd-nullcline} on the $N-I$ plane for small and large values of $D_{ext}$ (assuming the other parameters are suitably chosen) reveals a fold bifurcation.
-This is shown in Figure \ref{fig:one-cell-phase-plane}.
+This is shown in the figure below.
 

### 2025-04-11 snacks-presentation/main.tex
Revise paper, add Dora's discussion section.

@@
 \begin{frame}
+
   \frametitle{Our Chemical Equations}
-  Consider a cell $A$ with neighbour $B$. The Notch, Delta, and NICD concentrations of cell $A$ are given by $N_{A}$, $D_{A}$, and $I_{A}$ respectively (and similarly for cell $B$).
+
+  \onslide<1->{Consider a cell $A$ with neighbour $B$. The Notch, Delta, and NICD concentrations of cell $A$ are given by $N_{A}$, $D_{A}$, and $I_{A}$ respectively (and similarly for cell $B$).
+
+  \bigskip
+
   \begin{enumerate} 
@@
     \item $\emptyset \rightarrow D_{A}$: Delta production.
-  \end{enumerate}
+\end{enumerate}}
+
+  \bigskip
+
+  \onslide<2->{In the \textbf{Agent-Based} model, we simulate these reactions directly using the \emph{Gillespie Algorithm}. Under our assumptions, this is the most accurate model.}
+
 \end{frame}
@@
 
-  \frametitle{Our Models}
+  \frametitle{Differential Equation Models}
 
-  \onslide<1->{In the \textbf{Agent-Based} model, we simulate the chemical reactions directly using the \emph{Gillespie Algorithm}. Under our assumptions, this is the most accurate model.}
+  \onslide<1->{In the \textbf{ODE} model, we approximate the reaction kinetics by simulating the agent-based model \emph{in expectation}.
+Since this model is deterministic, we ensure cell differentiation with an initial perturbation before all simulations.
+
+$$\begin{aligned}
+  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\
+  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\
+  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
+\end{aligned}$$
+
+  }
 
@@
 
-  \onslide<2->{In the \textbf{ODE} model, we approximate the reaction kinetics by simulating the agent-based model \emph{in expectation}.
-To ensure cell differentiation, we apply an \emph{initial perturbation} to all simulations.}
-
-  \bigskip
-
-  \onslide<3->{In the \textbf{SDE} model, we reintroduce randomness by adding an independent \emph{Wiener Process} to each equation in the ODE model.
-This recreates some the stochasticity of the agent-based model.}
+  \onslide<2->{In the \textbf{SDE} model, we recreate some of the stochasticity of the agent-based model by adding an independent \emph{Wiener Process} to each equation above.}
 
 \end{frame}
-
-% NOTE: I can extend this section by talking about the Gillespie algorithm.
 
@@
   \onslide<1->{
-    We tested model stability by perturbing $5$ parameters: $N_{m}$, $D_{m}$, $K_{T}$, $\gamma$, and $\gamma_{I}$.
-    \begin{itemize}
-      \item $N_{m}$, $D_{m}$ are the maximum rates of Notch/Delta production.
-      \item $K_{T}$ is the binding rate.
-      \item $\gamma$, $\gamma_{I}$ are the Notch/Delta and NICD decay rates.
-    \end{itemize}
+    We tested model stability by perturbing $N_{m}$, $D_{m}$, $K_{T}$, $\gamma$, and $\gamma_{I}$.
+$$\begin{aligned}
+  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\
+  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\
+  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
+\end{aligned}$$
   }
-
-  \bigskip
 
@@
 \begin{frame}
-  \frametitle{Stability Analysis}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Simulations on Linear Domains}
+  \frametitle{Stability Analysis Results}
 \end{frame}
@@
 
-\begin{frame}
-  \frametitle{Patterns on Hexagonal Domains}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Patterns on Hexagonal Domains}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Patterns on Hexagonal Domains}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Patterns on Hexagonal Domains}
-\end{frame}
+\section{End}
 

### 2025-04-11 paper/paper.tex
Capitalise delta in Table1; rewrite key assumptions in Appendix C

@@
  $k_{T} N_{A}D_{B}$ &
- A Notch receptor from cell $A$ binds to a delta ligand from cell $B$, triggering the release of NICD in cell $A$. \\
+ A Notch receptor from cell $A$ binds to a Delta ligand from cell $B$, triggering the release of NICD in cell $A$. \\
  \hline
@@
 \subsection*{Key Assumptions}
+Throughout this derivation, we make the following assumptions:
+\begin{enumerate}
+  \item Fluctuations in molecular counts are small relative to their mean values; that is, $\sqrt{\text{Var}(N)} \ll \langle n \rangle$, $\sqrt{\text{Var}(D)} \ll \langle d \rangle$, and $\sqrt{\text{Var}(I)} \ll \langle i \rangle$.
 
-Throughout this derivation, we make the following assumptions:
+  \item Second moments can be approximated by the squares of the means, i.e., $\langle n^2 \rangle \approx \langle n \rangle^2$, $\langle d^2 \rangle \approx \langle d \rangle^2$, and $\langle i^2 \rangle \approx \langle i \rangle^2$.
 
-\begin{enumerate}
-  \item The standard deviation of the molecular counts is small compared to their mean values. Specifically:
-  \begin{itemize}
-    \item $\sqrt{\text{Var}(N)} \ll \langle n \rangle$
-    \item $\sqrt{\text{Var}(D)} \ll \langle d \rangle$
-    \item $\sqrt{\text{Var}(I)} \ll \langle i \rangle$
-  \end{itemize}
+  \item Molecular species are approximately uncorrelated, so that joint moments factorize, e.g., $\langle n \cdot d \rangle \approx \langle n \rangle \langle d \rangle$.
 
-  \item As a consequence of assumption 1, we can approximate:
-  \begin{itemize}
-    \item $\langle n^2 \rangle \approx \langle n \rangle^2$
-    \item $\langle d^2 \rangle \approx \langle d \rangle^2$
-    \item $\langle i^2 \rangle \approx \langle i \rangle^2$
-  \end{itemize}
-
-  \item The molecular counts are approximately independent, allowing us to factorize expected values:
-  \begin{itemize}
-    \item $\langle n \cdot d \rangle \approx \langle n \rangle \cdot \langle d \rangle$
-    \item $\langle n \cdot i \rangle \approx \langle n \rangle \cdot \langle i \rangle$
-    \item $\langle d \cdot i \rangle \approx \langle d \rangle \cdot \langle i \rangle$
-  \end{itemize}
-
-  \item For functions $H^+$ and $H^-$, we approximate:
-  \begin{itemize}
-    \item $\langle H^+(i) \rangle \approx H^+(\langle i \rangle)$
-    \item $\langle H^-(i) \rangle \approx H^-(\langle i \rangle)$
-    \item $\langle n \cdot H^+(i) \rangle \approx \langle n \rangle \cdot H^+(\langle i \rangle)$
-    \item $\langle d \cdot H^-(i) \rangle \approx \langle d \rangle \cdot H^-(\langle i \rangle)$
-  \end{itemize}
+  \item For Hill functions $H^+$ and $H^-$, we approximate expectation values by evaluating the function at the mean, i.e., $\langle H(i) \rangle \approx H(\langle i \rangle)$, and similarly $\langle x \cdot H(i) \rangle \approx \langle x \rangle H(\langle i \rangle)$.
 \end{enumerate}

### 2025-04-11 paper/paper.tex
Capitalise delta in Table1; rewrite key assumptions in Appendix C

@@
 \begin{enumerate}
-  \item Fluctuations in molecular counts are small relative to their mean values; that is, $\sqrt{\text{Var}(N)} \ll \langle n \rangle$, $\sqrt{\text{Var}(D)} \ll \langle d \rangle$, and $\sqrt{\text{Var}(I)} \ll \langle i \rangle$.
+  \item The standard deviation of molecular counts is small compared to their mean values, $\sqrt{\text{Var}(N)} \ll \langle n \rangle$, $\sqrt{\text{Var}(D)} \ll \langle d \rangle$, and $\sqrt{\text{Var}(I)} \ll \langle i \rangle$.
 
-  \item Second moments can be approximated by the squares of the means, i.e., $\langle n^2 \rangle \approx \langle n \rangle^2$, $\langle d^2 \rangle \approx \langle d \rangle^2$, and $\langle i^2 \rangle \approx \langle i \rangle^2$.
+  \item The second moments can be approximated by the squares of the means, i.e., $\langle n^2 \rangle \approx \langle n \rangle^2$, $\langle d^2 \rangle \approx \langle d \rangle^2$, and $\langle i^2 \rangle \approx \langle i \rangle^2$.
 
-  \item Molecular species are approximately uncorrelated, so that joint moments factorize, e.g., $\langle n \cdot d \rangle \approx \langle n \rangle \langle d \rangle$.
+  \item Molecular counts are approximately independent, so that joint moments factorize, e.g., $\langle n \cdot d \rangle \approx \langle n \rangle \langle d \rangle$.
 

### 2025-04-12 paper/paper.tex
Update results and discussion section with information about linear and hexagonal domains.

--- 2025-04-11 paper/paper.tex
@@
 However, in order to evaluate the quality of our model, we wanted to see if we could replicate the key result of cell differentiation under different parameter regimes.
-We identified five main parameters which effect whether the two-cell system differentiates.
+We identified five main parameters which determine whether the two-cell system differentiates.
 These parameters are the maximum rates of Notch and Delta production $n_{m}$ and $d_{m}$, the binding rate $k_{T}$, the Notch and Delta decay rate $\gamma$, and the NICD decay rate $\gamma_{I}$.
@@
 Collier et al. \cite{collier_pattern_1996} tested their model on linear domains of up to $40$ cells with zero-flux boundary conditions.
-They observed that cells on the boundary of the domain adopted the primary fate first.
+They observed that cells on the boundary of the domain adopted a fate first.
 Then, the alternating pattern of primary and secondary fated cells gradually spread inwards towards the center of the domain.
-In our first experiment on a linear domain, we replicated this result as shown in Figure.
+In our first experiment on a linear domain, we replicated this result as shown in Figure \ref{fig:linear-domain}.
 
-% TODO: Add linear domain figure here.
+\begin{figure}[!htp]
+   \makebox[\textwidth][c]{\includegraphics[width=1.3\textwidth]{img/linear_heatmap_Gillespie_V1.png}}
+   \caption{
+Results of the agent-based model run on a linear domain with 7 cells using Dirichlet boundary conditions.
+The first subplot shows Notch molecules per cell over time.
+Since noise is introduced every time step, the cell trajectories exhibit Brownian motion-like behaviour and thus have been smoothed out for better legibility.
+The second plot also shows Notch levels over time but in this visualization it is easier to see how cells influence their neighbours and how the cells on the edge of the domain are first to differentiate.
+The third plot shows the final Notch pattern, where primary fate cells are orange (indicating low Notch) and secondary fate cells are green (indicating high Notch).} 
+   \label{fig:linear-domain}
+\end{figure}
 
 We also attempted to broadly classify the types of patterns we observed on linear domains.
-In order to do this, we first need to define the notion of a \emph{pattern}.
-Since the ``default'' or ``regular'' outcome of a simulation on a linear domain is an alternating array of cells with each fate, we will call this pattern $0: [],\, 1: []$.
+In order to do this, we needed to define the notion of a \emph{pattern}.
+Since the ``default'' outcome of a simulation on a linear domain is an alternating array of cells with each fate, we will call this pattern $0: [],\, 1: []$.
 Then, each contiguous group of cells with the same fate (where $0$ is the primary fate and $1$ is the secondary fate) will add entries to the pattern arrays.
 For example, if there are $2$ contiguous pairs of cells with the primary fate and $1$ contiguous triplet of cells with the secondary fate, the simulation will have the pattern $0: [2, 2], \, 1: [3]$. 
-There is a lot more nuance related to pattern formation that we didn't have time to explore, but one interesting result is shown in Table \ref{tb:patterns}.
+There is a lot more nuance related to pattern formation that we were unable to explore, but one interesting result is shown in Table \ref{tb:patterns}.
 
@@
 
+Another phenomenon that we discovered, although we did not explore in detail, is that the SDE model was far more likely to allow for adjacent pairs of primary fate cells than the agent-based model when given identical parameter values and initial conditions.
+The agent-based model showed greater robustness to parameter change than the SDE model, as the general differentiation pattern noted by Collier et al.
+(shown in Figure \ref{fig:linear-domain}) could easily be disrupted by small adjustments to the binding coefficient $k_T$ or the Notch/Delta decay rate $\gamma$ leading to primary fate cell grouping.
+
 \subsection*{Hexagonal Domains}
 
-% TODO: Add results on hexagonal domains.
+Continuing to use the domains in Collier et al. \cite{collier_pattern_1996}, we next considered hexagonal grids of cells in a slanted parallelogram pattern. 
+We imposed both periodic and Dirichlet boundary conditions on these grids.
+In our simulations, we observed alternating patterns of primary fate cells appear on the $7$x$7$ grid (see Figure \ref{fig:hexagonal-domain}), especially for the agent-based model.
+Although the SDE model could replicate this same pattern depending on initial conditions, it was more likely to end up with pairs of adjacent low-notch cells.
+This behaviour was not noted by Collier et al. and points to some key differences between our models depending on the domain.
+
+\begin{figure}[!htp]
+   \makebox[\textwidth][c]{\includegraphics[width=1.3\textwidth]{img/HexDirichletPattern.png}}
+   \caption{Results of the agent-based model run on a hexagonal domain with 49 cells using Dirichlet boundary conditions.
+In this figure, black represents primary cell fate and white secondary cell fate.
+Although not all the secondary fated cells have the same Notch levels, this general pattern is stable long term with high Notch concentration around $300$ molecules.}
+   \label{fig:hexagonal-domain}
+\end{figure}
+
+The patterns on the hexagonal domains also exhibited increased sensitivity to boundary conditions and parameter values.
+In particular, identical starting conditions and parameters would often lead to markedly different final patterns on domains with periodic and Dirichlet boundary conditions.
+Figure \ref{fig:hexagonal-compare} highlights some of these differences.
+Increasing $\gamma$ from $0$ to $0.03$ could leave final patterns showing homogeneous cell fate, isolated primary cells, or even striping behaviour.
+Similar to the linear domain, cells on the edge of the hexagonal domain tended to differentiate first, especially when using Dirichlet boundary conditions.
+
+\begin{figure}[!htp]
+   \makebox[\textwidth][c]{\includegraphics[width=1.3\textwidth]{img/Hex_Compare.png}}
+   \caption{Results of running the SDE model on 49 hexagonal cells, with identical initial conditions comparing periodic and Dirichlet boundary conditions.
+Cells are coloured black if the notch level is less than $1$ at the final time step and left white otherwise.
+The first row (A) uses $\gamma = 0.01$ while the second row (B) uses $\gamma = 0.02$.
+All other parameters are equal for these simulations.
+}
+   \label{fig:hexagonal-compare}
+\end{figure}
 
@@
 
-Our comparison of the ODE, SDE, and agent-based models highlights a clear trade-off between simplicity, accuracy to the biology, and the role of stochasticity in modelling Delta-Notch signalling.
-The ODE model is computationally efficient and easy to analyze, but it can only break symmetry with an initial perturbation.
+Our comparison of the ODE, SDE, and agent-based models highlights a clear trade-off between simplicity and accuracy in modelling Delta-Notch signalling.
+The ODE model is computationally efficient and easy to analyze, but can only break symmetry with an initial perturbation.
 In other words, stable cell differentiation only occurs when the initial conditions are already imbalanced.
-The SDE model, by introducing intrinsic noise, allows cells to break symmetry without perturbations.
-This behaviour closely mirrors the agent-based model, which captures randomness at the level of individual molecules and allows differentiation to emerge purely from fluctuations in the receptor-ligand dynamics.
+Under the SDE model, the additional noise allows cells to break symmetry without perturbations.
+This behaviour closely mirrors the agent-based model, which captures randomness at the level of individual molecules.
+Although cell differentiation was easy to achieve with the stochastic models, controlling cell fate proved to be more challenging than we initially expected.
+Even when we imposed large initial perturbations in the two-cell system ($\pm 25\%$), the intended outcome only occurred about $80\%$ of the time.
+This result suggests that there are systemic biases in the Delta-Notch signalling pathway that induce specific fates in certain cells.
 
-Simulations of the agent-based model revealed that it is difficult to control cell fate using initial state alone.
-Even when we imposed large initial perturbations in the two-cell system ($\pm 25\%$), the intended outcome only occurred about $80\%$ of the time.
 When comparing the SDE model to the agent-based model, we found that the differentiation time varied considerably depending on the noise coefficient $\sigma$.
@@
 Finally, we concluded our investigation of the two-cell domain with a stability analysis, which found that the ODE model is the most sensitive to changes in parameter values, especially the decay rates $\gamma$ and $\gamma_{I}$.
-On linear domains, all three models exhibited the behaviour described by Collier et al., with an alternating pattern of cell differentiation spreading inward from the edges.
-The ODE model generated clean, perfectly alternating patterns, consistent with its deterministic structure.
-The SDE model introduced occasional irregularities but largely preserved alternation.
-However, the agent-based model produced the most diverse outcomes — especially in small or oddly shaped domains — where patterns often deviated from strict alternation.
-This variability points to a key biological insight: while stochasticity provides flexibility and diversity in pattern formation, it can also make patterns less predictable, particularly in small systems or near critical parameter boundaries.
+On linear domains, all three models exhibited the behaviour described by Collier et al., with an approximately alternating pattern of cell differentiation spreading inward from the edges.
+However, the SDE model and the agent-based model deviated from perfect alternation in the majority of simulations.
+Furthermore, the deviations were different between the two models, as contiguous groups of primary fated cells almost never appeared in agent-based simulations but appeared frequently in simulations of the SDE model.
+This variability points to a key biological insight: while randomness provides flexibility and diversity in pattern formation, it can also make patterns less predictable, particularly in small systems or near critical parameter boundaries.
+
+Finally, we replicated the qualitative results of Collier et al. on two-dimensional hexagonal domains. 
+We also found that our models were more sensitive to changes in parameter values and boundary conditions on the hexagonal domain compared to the linear and two-cell domains.
+However, due to computational limitations and time constraints we were unable to do a more thorough analysis of the models' behaviour on the hexagonal domain.
+Generally, our results on all three domains show that the choice of modelling formalism can drastically influence the results of simulations.
+Therefore, we suggest that modellers should also investigate whether their results are stable across different formalisms (a ``model stability analysis'') whenever possible.
 

### 2025-04-12 paper/paper.tex

@@
 \subsection*{Key Assumptions}
+Throughout this derivation, we make the following assumptions:
+\begin{enumerate}
+  \item The standard deviation of molecular counts is small compared to their mean values, $\sqrt{\text{Var}(N)} \ll \langle n \rangle$, $\sqrt{\text{Var}(D)} \ll \langle d \rangle$, and $\sqrt{\text{Var}(I)} \ll \langle i \rangle$.
 
-Throughout this derivation, we make the following assumptions:
+  \item The second moments can be approximated by the squares of the means, i.e., $\langle n^2 \rangle \approx \langle n \rangle^2$, $\langle d^2 \rangle \approx \langle d \rangle^2$, and $\langle i^2 \rangle \approx \langle i \rangle^2$.
 
-\begin{enumerate}
-  \item The standard deviation of the molecular counts is small compared to their mean values. Specifically:
-  \begin{itemize}
-    \item $\sqrt{\text{Var}(N)} \ll \langle n \rangle$
-    \item $\sqrt{\text{Var}(D)} \ll \langle d \rangle$
-    \item $\sqrt{\text{Var}(I)} \ll \langle i \rangle$
-  \end{itemize}
+  \item Molecular counts are approximately independent, so that joint moments factorize, e.g., $\langle n \cdot d \rangle \approx \langle n \rangle \langle d \rangle$.
 
-  \item As a consequence of assumption 1, we can approximate:
-  \begin{itemize}
-    \item $\langle n^2 \rangle \approx \langle n \rangle^2$
-    \item $\langle d^2 \rangle \approx \langle d \rangle^2$
-    \item $\langle i^2 \rangle \approx \langle i \rangle^2$
-  \end{itemize}
-
-  \item The molecular counts are approximately independent, allowing us to factorize expected values:
-  \begin{itemize}
-    \item $\langle n \cdot d \rangle \approx \langle n \rangle \cdot \langle d \rangle$
-    \item $\langle n \cdot i \rangle \approx \langle n \rangle \cdot \langle i \rangle$
-    \item $\langle d \cdot i \rangle \approx \langle d \rangle \cdot \langle i \rangle$
-  \end{itemize}
-
-  \item For functions $H^+$ and $H^-$, we approximate:
-  \begin{itemize}
-    \item $\langle H^+(i) \rangle \approx H^+(\langle i \rangle)$
-    \item $\langle H^-(i) \rangle \approx H^-(\langle i \rangle)$
-    \item $\langle n \cdot H^+(i) \rangle \approx \langle n \rangle \cdot H^+(\langle i \rangle)$
-    \item $\langle d \cdot H^-(i) \rangle \approx \langle d \rangle \cdot H^-(\langle i \rangle)$
-  \end{itemize}
+  \item For Hill functions $H^+$ and $H^-$, we approximate expectation values by evaluating the function at the mean, i.e., $\langle H(i) \rangle \approx H(\langle i \rangle)$, and similarly $\langle x \cdot H(i) \rangle \approx \langle x \rangle H(\langle i \rangle)$.
 \end{enumerate}
